---
title: "Threshold Cox + Output Differential Privacy (Demonstration)"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Threshold Cox + Output Differential Privacy (Demonstration)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F, purl=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
## Inline numbers in scientific notation as HTML (knitr's default
## prints a bare "1.8^{-6}" in HTML output).
sci <- function(x, digits = 2) {
  e <- floor(log10(abs(x)))
  if (x == 0 || e >= -2) return(formatC(x, format = "fg", digits = digits))
  sprintf("%s&nbsp;&times;&nbsp;10<sup>%d</sup>",
          formatC(x / 10^e, format = "f", digits = digits - 1), e)
}
## The encrypted chunks are gated `eval = RECOMPUTE` and are shown but
## not run when the vignette builds; their results are read from
## data(cox_threshold_dp_results), produced by
## data-raw/cox_threshold_dp_results.R from these same chunks.
## HOMOMORPHER_RECOMPUTE=true runs them for real.
if (!exists("RECOMPUTE"))
    RECOMPUTE <- isTRUE(as.logical(Sys.getenv("HOMOMORPHER_RECOMPUTE", "FALSE")))
data(cox_threshold_dp_results, package = "homomorpheR")
```

## Introduction
This vignette is a demonstration. `vignette("cox-threshold")` fits
the Cox model under threshold FHE: the fit matches the cleartext
fit, no single party can decrypt, and the aggregator sees only the
joint log-likelihood $\ell(\beta)$ at each optimizer query. Those
values are what the protocol reveals.

Here we ask what happens if each site also adds noise, so that the
released values satisfy **output differential privacy**. Adding the
noise takes one extra `rnorm()` call per site per query. With the
optimizers used here, the fits are poor at any noise level that
gives a small $\varepsilon$. The vignette runs the fits and reports
the numbers.

Threshold FHE without noise remains the package's main path. This
vignette shows what adding output DP on top costs.

## A brief output-DP primer

Given a query $f: \mathcal{D} \to \mathbb{R}$ on a dataset $D$,
the **Gaussian mechanism** [@DworkRoth2014, §3.5.3] releases
$\tilde f(D) = f(D) + \mathcal{N}(0, \sigma^2)$. If $f$'s
*sensitivity* (the largest change $f$ can undergo when one record
is added or removed) is $\Delta$, then for any
$\varepsilon \in (0, 1)$ and $\delta > 0$, choosing
$\sigma \geq \Delta \sqrt{2\ln(1.25/\delta)}/\varepsilon$ makes a
single release of $\tilde f$ satisfy $(\varepsilon, \delta)$
differential privacy.

The optimizer issues many queries, so the per-query budget
composes. We use **zCDP composition** [@BunSteinke2016]:
the Gaussian mechanism is $\rho$-zCDP with $\rho = (\Delta/\sigma)^2/2$;
$T$ queries compose linearly to $T \cdot \rho$ in zCDP units; and
that converts to $(\varepsilon, \delta)$ via
$\varepsilon = \rho + 2\sqrt{\rho \log(1/\delta)}$.

## The protocol

The setup is identical to `vignette("cox-threshold")`, except each
site adds an independent Gaussian noise term to its local
contribution before encryption. Since variances of independent
normals add, the decrypted total is
$\ell(\beta) + \mathcal{N}(0, \sigma^2)$ when each site adds
$\mathcal{N}(0, \sigma^2/N)$.

## The Cox setup (same DLBCL data as the other Cox vignettes)

```{r}
suppressPackageStartupMessages(library(survival))
library(homomorpheR)
library(stats4)
data(DLBCL)

cox_data <- split(
  DLBCL[, c("time", "status", "GCB_sig", "LN_sig",
            "Prolif_sig", "BMP6", "MHC2_sig", "Subgroup")],
  DLBCL$Subgroup)

agg_model <- coxph(Surv(time, status) ~ GCB_sig + LN_sig +
                       Prolif_sig + BMP6 + MHC2_sig +
                       strata(Subgroup),
                   data = DLBCL)
agg_coef <- coef(agg_model)

cph_control <- replace(coxph.control(), "iter.max", 0)

local_cox_nll <- function(data, beta) {
    fit <- tryCatch(
        coxph(Surv(time, status) ~ GCB_sig + LN_sig + Prolif_sig +
                  BMP6 + MHC2_sig,
              data    = data,
              init    = beta,
              control = cph_control),
        error = function(e) NULL)
    if (is.null(fit)) NA_real_ else -fit$loglik[1]
}
```

## Threshold setup and DP-noised workers

The threshold setup is the same as in `cox-threshold`. The only
change is in each worker's `contribution_fn`: it adds an
independent $\mathcal{N}(0, \sigma^2/N)$ draw to its local nLL
before returning it. The noisy value is then encrypted as usual.

```{r eval=RECOMPUTE}
cc <- openfhe.R::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L,
                           features             = c(openfhe.R::Feature$MULTIPARTY))

n_sites <- length(cox_data)

build_dp_workers <- function(sigma) {
    lapply(names(cox_data), function(nm) {
        make_worker(
            nm,
            data     = cox_data[[nm]],
            contribution_fn = function(data, beta) {
                nll <- local_cox_nll(data, beta)
                if (is.na(nll)) return(NA_real_)
                nll + rnorm(1L, mean = 0, sd = sigma / sqrt(n_sites))
            })
    })
}

fit_at_sigma <- function(sigma, method = "BFGS", seed = 1L) {
    set.seed(seed)   # stabilize the DP-noise draws across runs
    workers <- build_dp_workers(sigma)
    master  <- make_threshold_master("Aggregator",
                                     crypto_context = cc,
                                     sites          = workers)
    dp_nLL <- function(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig)
        master_aggregate(master, c(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig))
    stats4::mle(dp_nLL,
                start   = list(GCB_sig = 0, LN_sig = 0, Prolif_sig = 0,
                               BMP6    = 0, MHC2_sig = 0),
                method  = method,
                control = list(reltol = 1e-7))
}
```

## Mechanical correctness: $\sigma = 0$ reproduces `cox-threshold`

When the noise is zero the protocol reduces to the lossless
threshold protocol. The fitted coefficients match `coxph()` to
threshold-CKKS precision.

```{r eval=RECOMPUTE}
fit_clean <- fit_at_sigma(0)
clean_check <- data.frame(
    coefficient = names(agg_coef),
    cleartext   = unname(agg_coef),
    protocol    = unname(coef(fit_clean)[names(agg_coef)]),
    abs_diff    = abs(unname(coef(fit_clean)[names(agg_coef)] - agg_coef))
)
knitr::kable(clean_check, digits = 9,
             caption = "Threshold-DP protocol at sigma = 0 vs cleartext coxph()")
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
knitr::kable(cox_threshold_dp_results$clean_check, digits = 9,
             caption = "Threshold-DP protocol at sigma = 0 vs cleartext coxph()")
```

The maximum absolute coefficient difference at $\sigma = 0$ is
`r sci(max(cox_threshold_dp_results$clean_check$abs_diff), 3)`.
So any deviation from `coxph()` below comes from the noise.

## BFGS at increasing noise: where does it break?

We sweep $\sigma$ across four orders of magnitude. Each fit is
one full BFGS run over the threshold-DP encrypted nLL.

```{r eval=RECOMPUTE}
sigma_grid_bfgs <- c(1e-4, 1e-3, 1e-2, 1e-1, 1)
fits_bfgs <- lapply(sigma_grid_bfgs, fit_at_sigma, method = "BFGS")
names(fits_bfgs) <- sprintf("sigma_%g", sigma_grid_bfgs)

bfgs_table <- data.frame(
    sigma        = sigma_grid_bfgs,
    n_evals      = sapply(fits_bfgs, function(f) f@details$counts[["function"]]),
    GCB_sig      = sapply(fits_bfgs, function(f) coef(f)[["GCB_sig"]]),
    LN_sig       = sapply(fits_bfgs, function(f) coef(f)[["LN_sig"]]),
    Prolif_sig   = sapply(fits_bfgs, function(f) coef(f)[["Prolif_sig"]]),
    BMP6         = sapply(fits_bfgs, function(f) coef(f)[["BMP6"]]),
    MHC2_sig     = sapply(fits_bfgs, function(f) coef(f)[["MHC2_sig"]]),
    max_abs_diff = sapply(fits_bfgs, function(f)
        max(abs(coef(f) - agg_coef[names(coef(f))]))))
knitr::kable(bfgs_table, digits = 6,
             caption = "BFGS over the threshold-DP nLL at five noise scales")
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
knitr::kable(cox_threshold_dp_results$bfgs_table, digits = 6,
             caption = "BFGS over the threshold-DP nLL at five noise scales")
```

`max_abs_diff` grows slowly up to $\sigma = 0.01$ and jumps at
$\sigma = 0.1$. At $\sigma = 1$ BFGS converges, but the fit is
far from the cleartext one.

### Why BFGS breaks

BFGS estimates gradients by finite differences. With default
`ndeps = 1e-3` the gradient noise per coordinate inflates from
the function-value noise of $\sigma$ to roughly $\sigma\sqrt{2}/(2h)
\approx 707\,\sigma$. Near the optimum the true gradient is small
and the noise dominates. Any optimizer that differences nearby
function values has the same problem. The encryption is not
involved: the protocol releases $\ell(\beta) +
\mathcal{N}(0, \sigma^2)$ at every query, as intended.

## Nelder–Mead at the same noise scale

Nelder–Mead is gradient-free; it never amplifies noise through a
$1/h$ divisor. Noise still hurts it, but more gently than BFGS.

```{r eval=RECOMPUTE}
fit_nm_1 <- fit_at_sigma(1, method = "Nelder-Mead")

bfgs1 <- coef(fits_bfgs[["sigma_1"]])
nm1   <- coef(fit_nm_1)
coefs <- names(agg_coef)

nm_compare <- data.frame(
    coefficient  = coefs,
    coxph_agg = unname(agg_coef[coefs]),
    BFGS_sigma1  = unname(bfgs1[coefs]),
    NM_sigma1    = unname(nm1[coefs])
)
nm_caption <- sprintf(
    "BFGS vs Nelder-Mead at sigma = 1 (BFGS %d evals, NM %d evals)",
    fits_bfgs[["sigma_1"]]@details$counts[["function"]],
    fit_nm_1@details$counts[["function"]])
knitr::kable(nm_compare, digits = 6, caption = nm_caption)
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
knitr::kable(cox_threshold_dp_results$nm_compare, digits = 6,
             caption = cox_threshold_dp_results$nm_caption)
```

At $\sigma = 1$ BFGS gets at least one coefficient's sign wrong
relative to the cleartext fit. Nelder–Mead at the same $\sigma$
gets every sign right, but uses more function evaluations, and
each one adds to the privacy budget below.

## Privacy budget

For each fit, with sensitivity $\Delta = 1$ (placeholder) and
target $\delta = 10^{-5}$, zCDP composition gives:

```{r eval=RECOMPUTE}
zcdp_to_eps <- function(rho, delta = 1e-5) rho + 2 * sqrt(rho * log(1 / delta))

bfgs_budget <- data.frame(
    optimizer = "BFGS",
    sigma     = sigma_grid_bfgs,
    n_queries = sapply(fits_bfgs, function(f) f@details$counts[["function"]]),
    stringsAsFactors = FALSE)
nm_budget <- data.frame(
    optimizer = "Nelder-Mead",
    sigma     = 1,
    n_queries = fit_nm_1@details$counts[["function"]])
budget <- rbind(bfgs_budget, nm_budget)
budget$rho_per_query              <- (1 / budget$sigma)^2 / 2
budget$rho_total                  <- budget$n_queries * budget$rho_per_query
budget$epsilon_at_delta_1e_minus_5 <- zcdp_to_eps(budget$rho_total)
knitr::kable(budget, digits = 4,
             caption = "zCDP composition; sensitivity Delta = 1, target delta = 1e-5")
```

```{r eval=RECOMPUTE, echo=FALSE}
cox_threshold_dp_results <- list(clean_check = clean_check,
                                 bfgs_table  = bfgs_table,
                                 nm_compare  = nm_compare,
                                 nm_caption  = nm_caption,
                                 budget      = budget)
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
knitr::kable(cox_threshold_dp_results$budget, digits = 4,
             caption = "zCDP composition; sensitivity Delta = 1, target delta = 1e-5")
```

The `epsilon_at_delta_1e_minus_5` column shows that at noise
levels where the fits are still close, $\varepsilon$ is large.
Whether that is acceptable depends on the application.

### What it would take to push $\varepsilon$ smaller

Three directions, none pursued here:

1. **Tighter sensitivity bound.** $\Delta = 1$ is conservative;
   careful Cox-specific analysis could reduce it.
2. **Far fewer queries.** An optimizer that found the MLE in
   $\le 10$ queries would improve the budget by $\sim 10\times$;
   DP-SGD-style methods achieve this via subsampling and
   momentum.
3. **Subsampling amplification.** Per-record subsampling with the
   appropriate amplification accountant — incompatible with the
   current "compute the whole nLL at every site at every query"
   shape, but adaptable.

These are open research questions and outside the scope of the
package.

## Summary

1. At $\sigma = 0$ the protocol matches the cleartext `coxph()`
   fit to CKKS precision.
2. Threshold FHE controls who can decrypt; output DP limits what
   the released values reveal. Adding DP takes one `rnorm()` call
   per site per query and nothing new in homomorpheR.
3. The vignette fits five values of $\sigma$ with BFGS and one
   with Nelder–Mead, and tabulates the fits and privacy budgets.
4. BFGS amplifies the noise through finite-difference gradients
   and fails at moderate $\sigma$. Nelder–Mead at $\sigma = 1$
   gets the signs right but needs more queries. The accuracy loss
   comes from the optimizer's response to noise, not from the
   encryption.

## Limitations
- **Tight Cox sensitivity.** $\Delta = 1$ is a placeholder.
- **DP-SGD or stochastic gradient methods** [@AbadiCKMMTZ2016].
  Standard alternative for tight DP budgets; out of scope here.
- **Adaptive noise schedules.** All queries here use the same
  $\sigma$; adaptive schedules can be tighter.
- **Malicious-site protection.** A hostile site can corrupt its
  $\text{nLL}_i$. Out of scope.

For fits that match the cleartext fit, use
`vignette("cox-threshold")`; the aggregator then sees the exact
$\ell(\beta)$ values.

## References
