---
title: "Distributed Cox Regression with Threshold Key Generation"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Distributed Cox Regression with Threshold Key Generation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F, purl=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
## The encrypted chunks are gated `eval = RECOMPUTE` and are shown but
## not run when the vignette builds; their results are read from
## data(cox_threshold_results), produced by
## data-raw/cox_threshold_results.R from these same chunks.
## HOMOMORPHER_RECOMPUTE=true runs them for real.
if (!exists("RECOMPUTE"))
    RECOMPUTE <- isTRUE(as.logical(Sys.getenv("HOMOMORPHER_RECOMPUTE", "FALSE")))
data(cox_threshold_results, package = "homomorpheR")
```

## Introduction
The `cox` vignette fits a stratified Cox model across three sites
under CKKS. There the master holds the secret key and decrypts the
encrypted sum at every iteration of the optimizer, so the master
must be trusted not to decrypt anything else, such as a single
site's contribution.

In this vignette the three sites generate the CKKS key pair
jointly, and each keeps its own share of the secret key. No single
party holds the whole key. The master becomes an aggregator: it
adds the encrypted contributions and collects the sites' partial
decryptions, but it cannot decrypt anything by itself.

The data and the model are the same as in `vignette("cox")`. Only
who can decrypt changes.

## Threat model

Three sites and one untrusted aggregator:

- **Sites $1, 2, 3$** each hold private patient data and a secret
  key share $\mathit{sk}_i$. They are honest-but-curious among
  themselves and toward the aggregator.
- **Aggregator** holds no secret-key material. It receives the
  encrypted contributions, adds them, and sends the sum back to the
  sites for partial decryption. The encrypted values it handles
  tell it nothing on their own.

What the aggregator sees, by stage:

1. Encrypted local contributions $\mathit{ct}_i = E_{\mathit{pk}_{1..n}}(\ell_i)$. None decryptable alone.
2. The encrypted sum $\mathit{ct}_{\text{sum}} = \boxplus_i \mathit{ct}_i$. Not decryptable alone.
3. Partial decryptions $\rho_i$ contributed by each site. Not decryptable individually.
4. After combining the partial decryptions, the sum $\ell(\beta) = \sum_i \ell_i$ in the clear.

Step 4 reveals $\ell(\beta)$ to the aggregator, which is what the
master saw in `vignette("cox")`. The difference is that no single
party can decrypt an individual contribution or any intermediate
value. That takes a partial decryption from every site.

## The Cox setup (same DLBCL data as `cox.Rmd`)

```{r}
suppressPackageStartupMessages(library(survival))
library(homomorpheR)
data(DLBCL)

cox_data <- split(
  DLBCL[, c("time", "status", "GCB_sig", "LN_sig",
            "Prolif_sig", "BMP6", "MHC2_sig", "Subgroup")],
  DLBCL$Subgroup)
```

## The protocol

**Setup** (once):

1. Site 1 calls `key_gen(cc)` to produce its keypair $(\mathit{pk}_1, \mathit{sk}_1)$.
2. Site 2 calls `multiparty_key_gen(cc, pk_1)` to produce $(\mathit{pk}_{12}, \mathit{sk}_2)$.
3. Site 3 calls `multiparty_key_gen(cc, pk_{12})` to produce $(\mathit{pk}_{123}, \mathit{sk}_3)$.
4. The final $\mathit{pk}_{123}$ is the **joint public key**. Each site keeps its own $\mathit{sk}_i$.

**Per query** (called inside the optimizer):

1. Each site $i$ computes its local Cox negative log-likelihood
   $\ell_i(\beta)$ and encrypts it under the joint public key.
2. The aggregator sums the encrypted contributions homomorphically.
3. Each site partial-decrypts the sum using its own $\mathit{sk}_i$.
4. The aggregator fuses the partials to recover $\ell(\beta)$.

`make_threshold_master()` runs this chain across the sites in one
call and returns a `ThresholdMaster` holding the joint public key.
Each site keeps the share it generated. To decrypt, the master asks
every site for a partial decryption and combines them. This happens
inside the `decrypt()` method, so the `master_aggregate()` runner
from `cox.Rmd` works unchanged.

## Implementation

```{r}
cph_control <- replace(coxph.control(), "iter.max", 0)

local_cox_nll <- function(data, beta) {
    fit <- tryCatch(
        coxph(Surv(time, status) ~ GCB_sig + LN_sig + Prolif_sig +
                  BMP6 + MHC2_sig,
              data    = data,
              init    = beta,
              control = cph_control),
        error = function(e) NULL)
    if (is.null(fit)) NA_real_ else -fit$loglik[1]
}
```

The CKKS context needs the `MULTIPARTY` feature enabled so the
chained `multiparty_key_gen()` calls work:

```{r eval=RECOMPUTE}
cc <- openfhe.R::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L,
                           features             = c(openfhe.R::Feature$MULTIPARTY))
```

The sites come first, because the joint public key is built from
them. `make_threshold_master()` then takes the sites and returns
the master.

```{r eval=RECOMPUTE}
worker_gcb <- make_worker(name = "GCB",      data = cox_data[["GCB"]],
                          contribution_fn = local_cox_nll)
worker_abc <- make_worker(name = "ABC",      data = cox_data[["ABC"]],
                          contribution_fn = local_cox_nll)
worker_t3  <- make_worker(name = "Type III", data = cox_data[["Type III"]],
                          contribution_fn = local_cox_nll)

master <- make_threshold_master("Aggregator",
                                crypto_context = cc,
                                sites = list(worker_gcb, worker_abc, worker_t3))
```

The check below confirms that the master has no property holding
key shares and that the GCB site holds its own share:

```{r eval=RECOMPUTE}
share_check <- c(master_holds_shares = "secret_keys" %in% names(S7::props(master)),
                 gcb_holds_own_share = !is.null(worker_gcb@state$sk))
share_check
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
cox_threshold_results$share_check
```

## Iterative MLE through the threshold protocol

The optimizer code is the same as in `cox.Rmd`. Only the master
class differs.

```{r eval=RECOMPUTE}
library(stats4)

encrypted_nLL <- function(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig) {
    master_aggregate(master, c(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig))
}

fit <- mle(encrypted_nLL,
           start   = list(GCB_sig = 0, LN_sig = 0, Prolif_sig = 0,
                          BMP6    = 0, MHC2_sig = 0),
           method  = "BFGS",
           control = list(reltol = 1e-7))
summary(fit)
logLik(fit)
```

```{r eval=RECOMPUTE, echo=FALSE}
cox_threshold_results <- list(coef        = summary(fit)@coef,
                              loglik      = as.numeric(logLik(fit)),
                              counts      = fit@details$counts,
                              share_check = share_check)
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
cox_threshold_results$coef
cat(sprintf("'log Lik.' %f (df=5)\n", cox_threshold_results$loglik))
```

## Comparison with the cleartext fit

As in `vignette("cox")`, the check is the identical `mle()` objective
with the encrypted aggregation replaced by an ordinary sum of the
three sites' cleartext values.

```{r}
library(stats4)

plain_nLL <- function(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig) {
    beta <- c(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig)
    sum(vapply(cox_data, local_cox_nll, numeric(1), beta = beta))
}

fit_plain <- mle(plain_nLL,
                 start   = list(GCB_sig = 0, LN_sig = 0, Prolif_sig = 0,
                                BMP6    = 0, MHC2_sig = 0),
                 method  = "BFGS",
                 control = list(reltol = 1e-7))
```

```{r echo = FALSE}
thr_coefs   <- cox_threshold_results$coef[, "Estimate"]
plain_coefs <- coef(fit_plain)[names(thr_coefs)]
thr_diff    <- abs(thr_coefs - plain_coefs)
comparison <- data.frame(
    check.names           = FALSE,
    Coefficient           = names(thr_coefs),
    `mle(), threshold`    = unname(thr_coefs),
    `mle(), cleartext`    = unname(plain_coefs),
    `Absolute Difference` = formatC(unname(thr_diff), format = "e", digits = 2)
)
knitr::kable(comparison, digits = 7,
             caption = "Threshold-CKKS DLBCL Cox against the same objective evaluated in the clear.")
```

The threshold fit agrees with the cleartext fit to within
`r formatC(max(thr_diff), format = "e", digits = 2)` in every
coefficient.

## Discussion
1. **No single party holds the decryption key.** Each site generated
   its own share and kept it; the master holds only the joint public
   key, and has no property in which a share could sit. Encrypted
   intermediate values are undecryptable by any single party in the
   system, the aggregator included.
2. **The fit does not change.** The coefficients agree with the
   cleartext fit of the same objective to CKKS precision.
3. **Same optimizer, same callback shape, same code structure.**
   Compared to `cox.Rmd` only the setup changed: the workers are
   built first and handed to `make_threshold_master()` instead of
   being wired to a `make_ckks_master()` afterwards, because the
   joint key cannot exist before the sites do. The optimizer sees
   nothing different.

## Limitations
- **The aggregator sees $\ell(\beta)$ at every iteration.** That is
  the function value `mle()` asks for. The individual site
  contributions stay hidden. Hiding $\ell(\beta)$ as well would
  require running the optimizer on encrypted values, which is
  possible but considerably more complex.
- **Honest-but-curious is the trust model.** Sites are assumed to
  follow the protocol. A malicious site could submit a corrupted
  partial decryption to break the fit; detecting this requires
  additional protocol machinery (commitments, zero-knowledge
  proofs) that this vignette does not implement.
- **Output privacy is unchanged.** The released coefficients
  $\hat\beta$ are the same as the cleartext fit. Output-level
  attacks (membership inference, model inversion) remain in scope
  and motivate the differential-privacy demonstrations elsewhere
  in the package.

