---
title: "Threshold Cox + Output Differential Privacy (Demonstration)"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Threshold Cox + Output DP (Demonstration)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## Why this vignette exists

This vignette is a **demonstration**, not a recommendation.
`vignette("cox-threshold")` shows the lossless threshold-FHE
protocol: bit-identical to `coxph()`, no single decrypter, and the
aggregator sees only $\ell(\beta)$ at each optimiser query. That
trajectory of joint log-likelihood values is the residual exposure
the lossless protocol accepts as the cost of preserving accuracy.

A natural question is whether one can compose the threshold-FHE
channel with **output differential privacy** to bound that
residual exposure formally. The answer in this simulation is "yes,
the cryptographic protocol composes cleanly and the implementation
is one extra `rnorm()` call per site per query, but the
accuracy/privacy trade-off does not produce a usable point at
small $\varepsilon$ given the optimisers we use here." This
vignette runs that trade-off end-to-end and reports the numbers.

`feedback_he_not_dp` in the package's notes records the framing
explicitly: lossless threshold-FHE is the package's primary path;
DP demonstrations show *what happens if* a user composes output
DP on top, not the recommended deployment for precision-critical
workloads.

## A brief output-DP primer

Given a query $f: \mathcal{D} \to \mathbb{R}$ on a dataset $D$,
the **Gaussian mechanism** releases
$\tilde f(D) = f(D) + \mathcal{N}(0, \sigma^2)$. If $f$'s
*sensitivity* (the largest change $f$ can undergo when one record
is added or removed) is $\Delta$, then for any
$\varepsilon \in (0, 1)$ and $\delta > 0$, choosing
$\sigma \geq \Delta \sqrt{2\ln(1.25/\delta)}/\varepsilon$ makes a
single release of $\tilde f$ satisfy $(\varepsilon, \delta)$
differential privacy.

The optimiser issues many queries, so the per-query budget
composes. We use **zCDP composition** [Bun & Steinke, 2016]:
the Gaussian mechanism is $\rho$-zCDP with $\rho = (\Delta/\sigma)^2/2$;
$T$ queries compose linearly to $T \cdot \rho$ in zCDP units; and
that converts to $(\varepsilon, \delta)$ via
$\varepsilon = \rho + 2\sqrt{\rho \log(1/\delta)}$.

## The protocol

The setup is identical to `vignette("cox-threshold")`, except each
site adds an independent Gaussian noise term to its local
contribution before encryption. Since variances of independent
normals add, the decrypted total is
$\ell(\beta) + \mathcal{N}(0, \sigma^2)$ when each site adds
$\mathcal{N}(0, \sigma^2/N)$.

## The Cox setup (same simulation as the other Cox vignettes)

```{r}
suppressPackageStartupMessages(library(survival))
library(homomorpheR)
library(stats4)

sample_size <- c(n1 = 1000, n2 = 500, n3 = 1500)
set.seed(12345)

beta_true <- c(age = -0.015, sex = 0.2, bm = 0.001)
lambda_T  <- c(5, 4, 3)
lambda_C  <- 2

cox_data <- lapply(seq_along(sample_size), function(i) {
  n   <- sample_size[i]
  sex <- sample(c(0, 1), size = n, replace = TRUE)
  age <- sample(40:70, size = n, replace = TRUE)
  bm  <- rnorm(n)
  true_time <- rweibull(n, shape = 1,
                        scale = lambda_T[i] *
                          exp(beta_true["age"] * age +
                              beta_true["sex"] * sex +
                              beta_true["bm"]  * bm))
  cens_time <- rweibull(n, shape = 1, scale = lambda_C)
  time  <- pmin(true_time, cens_time)
  event <- (time == true_time)
  data.frame(stratum = i, sex = sex, age = age, bm = bm,
             time = time, event = event)
})

agg_data <- do.call(rbind, cox_data)
agg_model <- coxph(Surv(time, event) ~ sex + age + bm + strata(stratum),
                   data = agg_data)
oracle_coef <- coef(agg_model)

cph_control <- replace(coxph.control(), "iter.max", 0)

local_cox_nll <- function(data, beta) {
    fit <- tryCatch(
        coxph(Surv(time, event) ~ sex + age + bm,
              data    = data,
              init    = beta,
              control = cph_control),
        error = function(e) NULL)
    if (is.null(fit)) NA_real_ else -fit$loglik[1]
}
```

## Threshold setup and DP-noised workers

The threshold infrastructure is the same as `cox-threshold`. The
DP twist lives entirely inside the per-worker `local_fn`: each
worker adds an independent $\mathcal{N}(0, \sigma^2/N)$ draw to its
local nLL before returning. The master/worker runner sees only the
noisy local value and encrypts it as usual.

```{r}
cc <- openfhe::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L,
                           features             = c(openfhe::Feature$MULTIPARTY))

n_sites <- length(cox_data)

build_dp_workers <- function(sigma) {
    lapply(seq_len(n_sites), function(i) {
        make_worker(
            paste0("Site ", i),
            data     = cox_data[[i]],
            local_fn = function(data, beta) {
                nll <- local_cox_nll(data, beta)
                if (is.na(nll)) return(NA_real_)
                nll + rnorm(1L, mean = 0, sd = sigma / sqrt(n_sites))
            })
    })
}

fit_at_sigma <- function(sigma, method = "BFGS", seed = 1L) {
    set.seed(seed)   # stabilise the DP-noise draws across runs
    master <- make_threshold_master("Aggregator",
                                    crypto_context = cc,
                                    n_sites        = n_sites)
    set_workers(master, build_dp_workers(sigma))
    dp_nLL <- function(sex, age, bm)
        run_master_worker(master, c(sex, age, bm))
    stats4::mle(dp_nLL,
                start   = list(sex = 0, age = 0, bm = 0),
                method  = method,
                control = list(reltol = 1e-7))
}
```

## Mechanical correctness: $\sigma = 0$ reproduces `cox-threshold`

When the noise is zero the protocol reduces to the lossless
threshold protocol. The fitted coefficients match `coxph()` to
threshold-CKKS precision.

```{r}
fit_clean <- fit_at_sigma(0)
clean_check <- data.frame(
    coefficient = names(oracle_coef),
    cleartext   = unname(oracle_coef),
    protocol    = unname(coef(fit_clean)[names(oracle_coef)]),
    abs_diff    = abs(unname(coef(fit_clean)[names(oracle_coef)] - oracle_coef))
)
knitr::kable(clean_check, digits = 9,
             caption = "Threshold-DP protocol at sigma = 0 vs cleartext coxph()")
```

The maximum absolute coefficient difference at $\sigma = 0$ is
`r signif(max(clean_check$abs_diff), 3)`. The DP layer adds no
measurable error when the noise is zero, so the cryptographic
channel is doing exactly what it does in the lossless vignette.
Every deviation from `coxph()` below is attributable to the noise.

## BFGS at increasing noise: where does it break?

We sweep $\sigma$ across four orders of magnitude. Each fit is
one full BFGS run over the threshold-DP encrypted nLL.

```{r}
sigma_grid_bfgs <- c(1e-4, 1e-3, 1e-2, 1e-1, 1)
fits_bfgs <- lapply(sigma_grid_bfgs, fit_at_sigma, method = "BFGS")
names(fits_bfgs) <- sprintf("sigma_%g", sigma_grid_bfgs)

bfgs_table <- data.frame(
    sigma        = sigma_grid_bfgs,
    n_evals      = sapply(fits_bfgs, function(f) f@details$counts[["function"]]),
    sex          = sapply(fits_bfgs, function(f) coef(f)[["sex"]]),
    age          = sapply(fits_bfgs, function(f) coef(f)[["age"]]),
    bm           = sapply(fits_bfgs, function(f) coef(f)[["bm"]]),
    max_abs_diff = sapply(fits_bfgs, function(f)
        max(abs(coef(f) - oracle_coef[names(coef(f))]))))
knitr::kable(bfgs_table, digits = 6,
             caption = "BFGS over the threshold-DP nLL at five noise scales")
```

Read `max_abs_diff`: it grows roughly linearly with $\sigma$ in
the small-$\sigma$ regime (1e-4 → 1e-2) and the qualitative answer
collapses between $\sigma = 0.1$ and $\sigma = 1$. At $\sigma = 1$
BFGS exits cleanly but the fit is unusable.

### Why BFGS breaks

BFGS estimates gradients by finite differences. With default
`ndeps = 1e-3` the gradient noise per coordinate inflates from
the function-value noise of $\sigma$ to roughly $\sigma\sqrt{2}/(2h)
\approx 707\,\sigma$. Near the optimum the true gradient is small
and the noise dominates. Any optimiser that takes derivatives by
differencing nearby function values will have this exact failure
mode against any DP mechanism that adds enough noise to be
meaningful.

This is not a defect of the cryptographic protocol. The protocol
is doing exactly what it claims: releasing $\ell(\beta) +
\mathcal{N}(0, \sigma^2)$ at every query.

## Nelder–Mead at the same noise scale

Nelder–Mead is gradient-free; it never amplifies noise through a
$1/h$ divisor. Noise still hurts it, but more gently than BFGS.

```{r}
fit_nm_1 <- fit_at_sigma(1, method = "Nelder-Mead")
nm_compare <- data.frame(
    optimizer = c("BFGS", "Nelder-Mead", "coxph oracle"),
    sigma     = c(1, 1, NA),
    n_evals   = c(fits_bfgs[["sigma_1"]]@details$counts[["function"]],
                  fit_nm_1@details$counts[["function"]],
                  NA),
    sex       = c(coef(fits_bfgs[["sigma_1"]])[["sex"]],
                  coef(fit_nm_1)[["sex"]],
                  oracle_coef[["sex"]]),
    age       = c(coef(fits_bfgs[["sigma_1"]])[["age"]],
                  coef(fit_nm_1)[["age"]],
                  oracle_coef[["age"]]),
    bm        = c(coef(fits_bfgs[["sigma_1"]])[["bm"]],
                  coef(fit_nm_1)[["bm"]],
                  oracle_coef[["bm"]])
)
knitr::kable(nm_compare, digits = 6,
             caption = "BFGS vs Nelder-Mead at sigma = 1")
```

Nelder–Mead at $\sigma = 1$ recovers the right sign on `sex` and
keeps the substantive direction of the effect. BFGS at the same
$\sigma$ flipped the sign. The price is more function evaluations,
which feeds back into the privacy budget below.

## Privacy budget

For each fit, with sensitivity $\Delta = 1$ (placeholder) and
target $\delta = 10^{-5}$, zCDP composition gives:

```{r}
zcdp_to_eps <- function(rho, delta = 1e-5) rho + 2 * sqrt(rho * log(1 / delta))

bfgs_budget <- data.frame(
    optimizer = "BFGS",
    sigma     = sigma_grid_bfgs,
    n_queries = sapply(fits_bfgs, function(f) f@details$counts[["function"]]),
    stringsAsFactors = FALSE)
nm_budget <- data.frame(
    optimizer = "Nelder-Mead",
    sigma     = 1,
    n_queries = fit_nm_1@details$counts[["function"]])
budget <- rbind(bfgs_budget, nm_budget)
budget$rho_per_query              <- (1 / budget$sigma)^2 / 2
budget$rho_total                  <- budget$n_queries * budget$rho_per_query
budget$epsilon_at_delta_1e_minus_5 <- zcdp_to_eps(budget$rho_total)
knitr::kable(budget, digits = 4,
             caption = "zCDP composition; sensitivity Delta = 1, target delta = 1e-5")
```

Read the `epsilon_at_delta_1e_minus_5` column. At noise scales the
optimisers can tolerate without producing a broken fit, the
resulting $\varepsilon$ is large. Whether large $\varepsilon$ is
acceptable is a use-case decision; we report the numbers and stop.

### What it would take to push $\varepsilon$ smaller

Three directions, none pursued here:

1. **Tighter sensitivity bound.** $\Delta = 1$ is conservative;
   careful Cox-specific analysis could reduce it.
2. **Far fewer queries.** An optimiser that found the MLE in
   $\le 10$ queries would improve the budget by $\sim 10\times$;
   DP-SGD-style methods achieve this via subsampling and
   momentum.
3. **Subsampling amplification.** Per-record subsampling with the
   appropriate amplification accountant — incompatible with the
   current "compute the whole nLL at every site at every query"
   shape, but adaptable.

These are research questions, not package-level concerns. The
package's main story is the lossless cox progression.

## What this demonstrates

1. **The cryptographic protocol is correct under the noise.** At
   $\sigma = 0$ it reproduces `cox-threshold` exactly.
2. **The two layers compose without interaction.** Threshold FHE
   handles "who holds the key"; output DP handles "what the
   released values reveal." Adding the DP layer is one `rnorm()`
   call per site per query and zero new homomorpheR machinery.
3. **The accuracy/noise trade-off is characterised end-to-end.**
   The vignette runs five $\sigma$ values with BFGS and one with
   Nelder–Mead, and reports the recovered fits and the privacy
   budgets in tables.
4. **The optimiser-vs-noise interaction shapes the trade-off.**
   BFGS amplifies function-value noise through finite-difference
   gradients and breaks at moderate $\sigma$; Nelder–Mead at
   $\sigma = 1$ recovers a usable estimate at the cost of more
   queries. The cryptographic layer is not the limit — the
   optimiser is.

## What this does not address

- **Tight Cox sensitivity.** $\Delta = 1$ is a placeholder.
- **DP-SGD or stochastic gradient methods.** Standard alternative
  for tight DP budgets; out of scope here.
- **Adaptive noise schedules.** All queries here use the same
  $\sigma$; adaptive schedules can be tighter.
- **Malicious-site protection.** A hostile site can corrupt
  either its $\text{nLL}_i$ or its $z_i$. Out of scope.

Readers who need lossless fits should use
`vignette("cox-threshold")` and accept the residual
released-function leakage that comes with it.

## References

- Dwork & Roth (2014). *The Algorithmic Foundations of
  Differential Privacy.* §3.5.3 (Gaussian mechanism).
- Bun & Steinke (2016). *Concentrated Differential Privacy.* TCC.
  Source for the zCDP composition bound used above.
- Abadi et al. (2016). *Deep Learning with Differential Privacy.*
  CCS. The DP-SGD paper.
