---
title: "Threshold Cox + Output Differential Privacy (Demonstration)"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Threshold Cox + Output Differential Privacy (Demonstration)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r ktab, echo=FALSE}
## Tables through kableExtra. Math written as $...$ in headers, cells
## and captions becomes \( ... \) and `code` becomes <code>, because
## pandoc does not process math inside a raw HTML table.
ktab <- function(x, ..., col.names = names(x), caption = NULL) {
    tex <- function(s)
        gsub("`([^`]*)`", "<code>\\1</code>",
             gsub("\\$([^$]+)\\$", "\\\\(\\1\\\\)", s))
    chr <- vapply(x, is.character, logical(1))
    x[chr] <- lapply(x[chr], tex)
    tab <- knitr::kable(x, format = "html", escape = FALSE,
                        col.names = tex(col.names),
                        caption = if (!is.null(caption)) tex(caption), ...)
    kableExtra::kable_styling(tab, bootstrap_options = c("striped", "condensed"),
                              full_width = TRUE)
}
```

```{r echo=F, purl=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
## Inline numbers in scientific notation as HTML (knitr's default
## prints a bare "1.8^{-6}" in HTML output).
sci <- function(x, digits = 2) {
  e <- floor(log10(abs(x)))
  if (x == 0 || e >= -2) return(formatC(x, format = "fg", digits = digits))
  sprintf("%s&nbsp;&times;&nbsp;10<sup>%d</sup>",
          formatC(x / 10^e, format = "f", digits = digits - 1), e)
}
## The encrypted chunks are gated `eval = RECOMPUTE` and are shown but
## not run when the vignette builds; their results are read from
## data(cox_threshold_dp_results), produced by
## data-raw/cox_threshold_dp_results.R from these same chunks.
## HOMOMORPHER_RECOMPUTE=true runs them for real.
if (!exists("RECOMPUTE"))
    RECOMPUTE <- isTRUE(as.logical(Sys.getenv("HOMOMORPHER_RECOMPUTE", "FALSE")))
data(cox_threshold_dp_results, package = "homomorpheR")
```

## Introduction
This vignette is a demonstration. The previous `vignette("cox-threshold")` fits
the Cox model under threshold FHE: the fit matches the cleartext
fit, no single party can decrypt, and the aggregator sees only the
joint log-likelihood $\ell(\beta)$ at each optimizer query. 

Here we ask what happens if each site also adds Gaussian noise of the
kind an *output differential privacy* mechanism uses. The sensitivity
$\Delta = 1$ below is a placeholder, so the $\varepsilon$ values are
not a privacy guarantee. Adding the
noise takes one extra `rnorm()` call per site per query. With the
optimizers used here, the fits are poor at any noise level that
gives a small $\varepsilon$. The vignette runs the fits and reports
the numbers.

## A brief output-DP primer

Given a query $f: \mathcal{D} \to \mathbb{R}$ on a dataset $D$,
the *Gaussian mechanism* [@DworkRoth2014, §3.5.3] releases
$\tilde f(D) = f(D) + \mathcal{N}(0, \sigma^2)$. If $f$'s
*sensitivity* (the largest change $f$ can undergo when one record
is added or removed) is $\Delta$, then for any
$\varepsilon \in (0, 1)$ and $\delta > 0$, choosing
$\sigma \geq \Delta \sqrt{2\ln(1.25/\delta)}/\varepsilon$ makes a
single release of $\tilde f$ satisfy $(\varepsilon, \delta)$
differential privacy.

The optimizer issues many queries, so the per-query budget
composes. We use *zCDP composition* [@BunSteinke2016]:
the Gaussian mechanism is $\rho$-zCDP with $\rho = (\Delta/\sigma)^2/2$;
$T$ queries compose linearly to $T \cdot \rho$ in zCDP units; and
that converts to $(\varepsilon, \delta)$ via
$\varepsilon = \rho + 2\sqrt{\rho \log(1/\delta)}$.

## The protocol

The setup is identical to `vignette("cox-threshold")`, except each
site adds an independent Gaussian noise term to its local
contribution before encryption. Since variances of independent
normals add, the decrypted total is
$\ell(\beta) + \mathcal{N}(0, \sigma^2)$ when each site adds
$\mathcal{N}(0, \sigma^2/N)$.

## The Cox setup (same DLBCL data as the other Cox vignettes)

```{r}
suppressPackageStartupMessages(library(survival))
library(homomorpheR)
library(stats4)
data(DLBCL)

cox_data <- split(
  DLBCL[, c("time", "status", "GCB_sig", "LN_sig",
            "Prolif_sig", "BMP6", "MHC2_sig", "Subgroup")],
  DLBCL$Subgroup)

agg_model <- coxph(Surv(time, status) ~ GCB_sig + LN_sig +
                       Prolif_sig + BMP6 + MHC2_sig +
                       strata(Subgroup),
                   data = DLBCL)
agg_coef <- coef(agg_model)

cph_control <- replace(coxph.control(), "iter.max", 0)

local_cox_nll <- function(data, beta) {
    fit <- tryCatch(
        coxph(Surv(time, status) ~ GCB_sig + LN_sig + Prolif_sig +
                  BMP6 + MHC2_sig,
              data    = data,
              init    = beta,
              control = cph_control),
        error = function(e) NULL)
    if (is.null(fit)) NA_real_ else -fit$loglik[1]
}
```

## Threshold setup and DP-noised workers

The threshold setup is also the same as in `cox-threshold`. The only
change is in each worker's `contribution_fn`: it adds an independent
$\mathcal{N}(0, \sigma^2/N)$ draw to its local nLL before returning
it. The noisy value is then encrypted as usual.

```{r eval=RECOMPUTE}
cc <- openfhe.R::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L,
                           features             = c(openfhe.R::Feature$MULTIPARTY))

n_sites <- length(cox_data)

build_dp_workers <- function(sigma) {
    lapply(names(cox_data), function(nm) {
        make_worker(
            nm,
            data     = cox_data[[nm]],
            contribution_fn = function(data, beta) {
                nll <- local_cox_nll(data, beta)
                if (is.na(nll)) return(NA_real_)
                nll + rnorm(1L, mean = 0, sd = sigma / sqrt(n_sites))
            })
    })
}

fit_at_sigma <- function(sigma, method = "BFGS", seed = 1L) {
    set.seed(seed)   # stabilize the DP-noise draws across runs
    workers <- build_dp_workers(sigma)
    master  <- make_threshold_master("Aggregator",
                                     crypto_context = cc,
                                     sites          = workers)
    ## Every call is one decrypted release, including the calls optim()
    ## makes for finite-difference gradients and mle() for the Hessian.
    n_queries <- 0L
    dp_nLL <- function(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig) {
        n_queries <<- n_queries + 1L
        master_aggregate(master, c(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig))
    }
    fit <- stats4::mle(dp_nLL,
                       start   = list(GCB_sig = 0, LN_sig = 0, Prolif_sig = 0,
                                      BMP6    = 0, MHC2_sig = 0),
                       method  = method,
                       control = list(reltol = 1e-7))
    list(fit = fit, n_queries = n_queries)
}
```

## Mechanical correctness: $\sigma = 0$ reproduces `cox-threshold`

When the noise is zero the protocol reduces to the lossless
threshold protocol. The fitted coefficients match `coxph()` to
threshold-CKKS precision.

```{r echo=FALSE}
## Table display in mathematical notation. The stored tables keep
## their plain column names; only what is printed changes.
tex_pow10 <- function(x)
    ifelse(x == 1, "$1$", sprintf("$10^{%d}$", as.integer(round(log10(x)))))
tex_sci <- function(x, digits = 4) {
    e <- floor(log10(abs(x)))
    ifelse(x == 0 | (e >= -2 & e < 4),
           sprintf("$%s$", trimws(formatC(x, format = "fg", digits = digits))),
           sprintf("$%s \\times 10^{%d}$",
                   trimws(formatC(x / 10^e, format = "fg", digits = digits)),
                   as.integer(e)))
}
show_clean <- function(tab) {
    tab$abs_diff <- tex_sci(tab$abs_diff)
    ktab(tab, digits = 9, col.names = clean_cols, caption = clean_caption,
                 align = "lrrr")
}
clean_cols  <- c("Coefficient", "$\\hat\\beta$, `coxph()`",
                 "$\\hat\\beta$, protocol at $\\sigma = 0$",
                 "$\\lvert \\text{difference} \\rvert$")
clean_caption <- "Threshold-DP protocol at $\\sigma = 0$ vs cleartext `coxph()`"
sweep_cols  <- c("$\\sigma$", "Queries","GCB_sig", "LN_sig", "Prolif_sig",
                 "BMP6", "MHC2_sig",
                 "$\\max_j \\lvert \\hat\\beta_j - \\hat\\beta_j^{\\text{coxph}} \\rvert$")
budget_cols <- c("Optimizer", "$\\sigma$", "Queries $k$", "$\\rho$ per query",
                 "$\\rho_{\\text{total}} = k\\rho$", "$\\varepsilon$ at $\\delta = 10^{-5}$")
budget_caption <- "zCDP composition; sensitivity $\\Delta = 1$, target $\\delta = 10^{-5}$"
show_sweep <- function(tab, optimizer) {
    tab$sigma <- tex_pow10(tab$sigma)
    ktab(tab, digits = 6, row.names = FALSE, col.names = sweep_cols,
                 caption = sprintf("%s over the threshold-DP nLL at %d values of $\\sigma$",
                                   optimizer, nrow(tab)))
}
show_budget <- function(tab) {
    tab$sigma <- tex_pow10(tab$sigma)
    for (v in c("rho_per_query", "rho_total", "epsilon_at_delta_1e_minus_5"))
        tab[[v]] <- tex_sci(tab[[v]])
    ktab(tab, row.names = FALSE, col.names = budget_cols,
                 caption = budget_caption, align = "lrrrrr")
}
```

```{r eval=RECOMPUTE}
fit_clean <- fit_at_sigma(0)$fit
clean_check <- data.frame(
    coefficient = names(agg_coef),
    cleartext   = unname(agg_coef),
    protocol    = unname(coef(fit_clean)[names(agg_coef)]),
    abs_diff    = abs(unname(coef(fit_clean)[names(agg_coef)] - agg_coef))
)
show_clean(clean_check)
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
show_clean(cox_threshold_dp_results$clean_check)
```

The maximum absolute coefficient difference at $\sigma = 0$ is
`r sci(max(cox_threshold_dp_results$clean_check$abs_diff), 3)`.
So any deviation from `coxph()` below comes from the noise.

## Increasing noise: where does it break?

We sweep $\sigma$ across five orders of magnitude. Each fit is one
full BFGS run over the threshold-DP encrypted nLL. BFGS estimates the
gradient by finite differences of these nLL values.

```{r eval=RECOMPUTE}
sigma_grid <- c(1e-5, 1e-4, 1e-3, 1e-2, 1e-1, 1)

sweep_table <- function(fits) data.frame(
    sigma        = sigma_grid,
    n_queries    = sapply(fits, `[[`, "n_queries"),
    GCB_sig      = sapply(fits, function(f) coef(f$fit)[["GCB_sig"]]),
    LN_sig       = sapply(fits, function(f) coef(f$fit)[["LN_sig"]]),
    Prolif_sig   = sapply(fits, function(f) coef(f$fit)[["Prolif_sig"]]),
    BMP6         = sapply(fits, function(f) coef(f$fit)[["BMP6"]]),
    MHC2_sig     = sapply(fits, function(f) coef(f$fit)[["MHC2_sig"]]),
    max_abs_diff = sapply(fits, function(f)
        max(abs(coef(f$fit) - agg_coef[names(coef(f$fit))]))))

fits_bfgs  <- lapply(sigma_grid, fit_at_sigma, method = "BFGS")
bfgs_table <- sweep_table(fits_bfgs)
show_sweep(bfgs_table, "BFGS")
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
show_sweep(cox_threshold_dp_results$bfgs_table, "BFGS")
```

## Nelder–Mead at the same noise scales

The same sweep with Nelder–Mead.

```{r eval=RECOMPUTE}
fits_nm  <- lapply(sigma_grid, fit_at_sigma, method = "Nelder-Mead")
nm_table <- sweep_table(fits_nm)
show_sweep(nm_table, "Nelder–Mead")
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
show_sweep(cox_threshold_dp_results$nm_table, "Nelder–Mead")
```

Fidelity decays as $\sigma$ grows.

## Privacy budget

For the fits at the first three values of $\sigma$, with
sensitivity $\Delta = 1$ (placeholder) and target
$\delta = 10^{-5}$, zCDP composition gives:

```{r eval=RECOMPUTE}
zcdp_to_eps <- function(rho, delta = 1e-5) rho + 2 * sqrt(rho * log(1 / delta))

budget_rows <- function(optimizer, tab) data.frame(
    optimizer = optimizer,
    sigma     = tab$sigma[1:3],
    n_queries = tab$n_queries[1:3])
budget <- rbind(budget_rows("BFGS", bfgs_table),
                budget_rows("Nelder–Mead", nm_table))
budget$rho_per_query              <- (1 / budget$sigma)^2 / 2
budget$rho_total                  <- budget$n_queries * budget$rho_per_query
budget$epsilon_at_delta_1e_minus_5 <- zcdp_to_eps(budget$rho_total)
show_budget(budget)
```

```{r eval=RECOMPUTE, echo=FALSE}
cox_threshold_dp_results <- list(clean_check = clean_check,
                                 bfgs_table  = bfgs_table,
                                 nm_table    = nm_table,
                                 budget      = budget)
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
show_budget(cox_threshold_dp_results$budget)
```

The $\varepsilon$ column shows that at noise
levels where the fits are still close, $\varepsilon$ is large.
Whether that is acceptable depends on the application.

One could also explore tighter sensitivity bound $\Delta$ or allow far
fewer queries, etc. We don't do that here.

## References
