---
title: "Distributed Cox Regression with Threshold Key Generation"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Distributed Cox Regression with Threshold Key Generation}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## Introduction
The companion `cox` vignette fits a stratified Cox model across three
sites under CKKS via master/worker. The master holds the secret key
and decrypts the encrypted partial likelihood at every iteration of
the optimizer. This is *operationally* convenient — one party owns
decryption — but it leaves a meaningful piece of the threat model
unaddressed: the master is trusted not to decrypt anything it
shouldn't.

The end of the `cox` vignette flagged this:

> **Threshold key generation**: in a real deployment, the secret key
> would be split across the sites (n-of-n threshold) so that no
> single party — not even the master — can decrypt intermediate
> values unilaterally.

This vignette implements exactly that. Three sites jointly generate
a CKKS key pair so that no single party holds the joint secret key.
The master's role is reduced to an aggregator: it sums encrypted
contributions and orchestrates partial-decryption fan-in, but it
cannot decrypt anything by itself.

The mathematical content is identical to `vignette("cox")` — same
data, same fit, same coefficients to floating-point precision. Only
the trust model changes.

## Threat model

Three sites and one untrusted aggregator:

- **Sites $1, 2, 3$** each hold private patient data and a secret
  key share $\mathit{sk}_i$. They are honest-but-curious among
  themselves and toward the aggregator.
- **Aggregator** holds no secret-key material. It receives encrypted
  contributions, sums them homomorphically, and broadcasts the sum
  back for partial decryption. It is *fully untrusted*: a curious
  or compromised aggregator gains no information from the
  encrypted values it processes.

What the aggregator sees, by stage:
1. Encrypted local contributions $\mathit{ct}_i = E_{\mathit{pk}_{1..n}}(\ell_i)$. None decryptable alone.
2. The encrypted sum $\mathit{ct}_{\text{sum}} = \boxplus_i \mathit{ct}_i$. Not decryptable alone.
3. Partial decryptions $\rho_i$ contributed by each site. Not decryptable individually.
4. After fusion, the *sum in the clear* $\ell(\beta) = \sum_i \ell_i$. This appears at the aggregator only after step 4.

Step 4 reveals $\ell(\beta)$ to the aggregator — the same thing the
master saw in `vignette("cox")`. The strict improvement is that
*nobody else* can decrypt anything along the way: a captured log of
encrypted values or partial decryptions is useless without the joint
fusion.

## The Cox setup (same DLBCL data as `cox.Rmd`)

```{r}
suppressPackageStartupMessages(library(survival))
library(homomorpheR)
data(DLBCL)

cox_data <- split(
  DLBCL[, c("time", "status", "GCB_sig", "LN_sig",
            "Prolif_sig", "BMP6", "MHC2_sig", "Subgroup")],
  DLBCL$Subgroup)

agg_model <- coxph(Surv(time, status) ~ GCB_sig + LN_sig +
                       Prolif_sig + BMP6 + MHC2_sig +
                       strata(Subgroup),
                   data = DLBCL)
```

## The protocol

**Setup** (once):

1. Site 1 calls `key_gen(cc)` to produce its keypair $(\mathit{pk}_1, \mathit{sk}_1)$.
2. Site 2 calls `multiparty_key_gen(cc, pk_1)` to produce $(\mathit{pk}_{12}, \mathit{sk}_2)$.
3. Site 3 calls `multiparty_key_gen(cc, pk_{12})` to produce $(\mathit{pk}_{123}, \mathit{sk}_3)$.
4. The final $\mathit{pk}_{123}$ is the **joint public key**. Each site keeps its own $\mathit{sk}_i$.

**Per query** (called inside the optimizer):

1. Each site $i$ computes its local Cox negative log-likelihood
   $\ell_i(\beta)$ and encrypts it under the joint public key.
2. The aggregator sums the encrypted contributions homomorphically.
3. Each site partial-decrypts the sum using its own $\mathit{sk}_i$.
4. The aggregator fuses the partials to recover $\ell(\beta)$.

`homomorpheR` exports a `make_threshold_master()` constructor that
runs the chained key generation in one call and returns a
[ThresholdMaster] holding the joint public key plus the per-site
secret shares. The same `set_workers()` and `master_aggregate()`
helpers used by the single-decrypter `cox.Rmd` work unchanged here:
the threshold-specific encrypt/partial-decrypt/fuse logic lives
inside `master_encrypt()` and `master_decrypt()` methods on
`ThresholdMaster`, so the per-iteration code looks the same as the
single-decrypter case.

## Implementation

```{r}
cph_control <- replace(coxph.control(), "iter.max", 0)

local_cox_nll <- function(data, beta) {
    fit <- tryCatch(
        coxph(Surv(time, status) ~ GCB_sig + LN_sig + Prolif_sig +
                  BMP6 + MHC2_sig,
              data    = data,
              init    = beta,
              control = cph_control),
        error = function(e) NULL)
    if (is.null(fit)) NA_real_ else -fit$loglik[1]
}
```

The CKKS context needs the `MULTIPARTY` feature enabled so the
chained `multiparty_key_gen()` calls work:

```{r}
cc <- openfhe.R::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L,
                           features             = c(openfhe.R::Feature$MULTIPARTY))
```

`make_threshold_master(name, cc, n_sites)` runs the chained key
generation and returns a master holding the joint public key and a
list of per-site secret shares. The master's `master_encrypt` /
`master_decrypt` methods automatically use the joint pk for
encryption and the partial-decrypt fan-in for decryption.

```{r}
master <- make_threshold_master("Aggregator",
                                crypto_context = cc,
                                n_sites        = 3)

worker_gcb <- make_worker("GCB",      data = cox_data[["GCB"]],      local_fn = local_cox_nll)
worker_abc <- make_worker("ABC",      data = cox_data[["ABC"]],      local_fn = local_cox_nll)
worker_t3  <- make_worker("Type III", data = cox_data[["Type III"]], local_fn = local_cox_nll)

set_workers(master, list(worker_gcb, worker_abc, worker_t3))
```

## Iterative MLE through the threshold protocol

The optimizer-facing code is *identical* to `cox.Rmd`. Same
`master_aggregate()` runner, same `mle()` driver — only the
underlying master class changed.

```{r}
library(stats4)

encrypted_nLL <- function(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig) {
    master_aggregate(master, c(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig))
}

fit <- mle(encrypted_nLL,
           start   = list(GCB_sig = 0, LN_sig = 0, Prolif_sig = 0,
                          BMP6    = 0, MHC2_sig = 0),
           method  = "BFGS",
           control = list(reltol = 1e-7))
summary(fit)
logLik(fit)
```

## Comparison with the cleartext fit

```{r}
summary(agg_model)
cat(sprintf("logLik(threshold-distributed encrypted): %f\n",
            as.numeric(logLik(fit))))
cat(sprintf("logLik(aggregated cleartext)            : %f\n",
            agg_model$loglik[2]))
```

```{r echo = FALSE}
mle_coefs   <- coef(fit)
coxph_coefs <- coef(agg_model)
comparison <- data.frame(
    coefficient            = names(mle_coefs),
    threshold_distributed  = unname(mle_coefs),
    aggregated_cleartext   = unname(coxph_coefs[names(mle_coefs)]),
    abs_diff               = unname(abs(mle_coefs - coxph_coefs[names(mle_coefs)]))
)
knitr::kable(comparison, digits = 9,
             caption = "Threshold-distributed BFGS vs. aggregated-cleartext coxph()")
```

## Discussion
1. **No single party holds the decryption key.** `make_threshold_master()`
   distributes the secret across three sites; the master holds
   only the joint public key. Encrypted intermediate values are
   undecryptable by any single party in the system.
2. **Same statistical fit as the trusted-master version.** The
   coefficients and standard errors agree with `coxph()` to the
   precision the optimizer cares about. Threshold key generation
   is a trust-model improvement; it does not change the fit.
3. **Same optimizer, same callback shape, same code structure.**
   The only line that changed compared to `cox.Rmd` is the master
   constructor (`make_threshold_master()` instead of
   `make_ckks_master()`). The optimizer sees nothing different.

## Limitations
- **The aggregator sees $\ell(\beta)$ at every iteration.** That is
  the optimizer's *function value* — what `mle()` is asking for.
  The information leak is "the value of the joint partial
  log-likelihood at every $\beta$ visited," not "individual
  per-site contributions" (which are properly hidden). Hiding
  $\ell(\beta)$ would require running the optimizer inside the
  encrypted domain — feasible but considerably more complex.
- **Honest-but-curious is the trust model.** Sites are assumed to
  follow the protocol. A malicious site could submit a corrupted
  partial decryption to break the fit; detecting this requires
  additional protocol machinery (commitments, zero-knowledge
  proofs) that this vignette does not implement.
- **Output privacy is unchanged.** The released coefficients
  $\hat\beta$ are the same as the cleartext fit. Output-level
  attacks (membership inference, model inversion) remain in scope
  and motivate the differential-privacy demonstrations elsewhere
  in the package.

## Where this fits

Read `vignette("cox")` first if you have not — that vignette
establishes the master/worker protocol and the optimizer-driven
shape this one inherits. The only difference here is the master:
threshold instead of single-decrypter. Same code, stronger trust
model.
