---
title: "Distributed Stratified Cox Regression under CKKS"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Distributed Cox under CKKS}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## The statistical problem

The Cox proportional hazards model is one of the workhorses of
medical statistics. Given covariates $x_i$ for subject $i$, an event
time $t_i$, and an event indicator $\delta_i$, the hazard is modeled
as

$$
h(t \mid x_i) \;=\; h_0(t) \, \exp(\beta^\top x_i)
$$

where $h_0(t)$ is an unspecified baseline hazard and $\beta$ is the
vector of regression coefficients we want to estimate. The
**partial log-likelihood** depends only on $\beta$:

$$
\ell(\beta) \;=\; \sum_{i: \delta_i = 1}
  \left[\, \beta^\top x_i \;-\; \log\!\!\sum_{j \in R_i} \exp(\beta^\top x_j) \,\right]
$$

where $R_i$ is the risk set at time $t_i$.

For **stratified** Cox regression — when baseline hazards differ
across strata (e.g. across study sites) but the coefficients $\beta$
are shared — the partial log-likelihood becomes a sum over strata:

$$
\ell(\beta) \;=\; \sum_{s=1}^{S} \ell_s(\beta).
$$

This additive decomposition is exactly what we need for a
privacy-preserving distributed protocol. The same decomposition
underlies the master/worker architecture used by `distcomp`,
`DataSHIELD`, and `WebDISCO`: a central master broadcasts the
current $\beta$, each site computes its local $\ell_s(\beta)$ on
its private data, and the master combines the local contributions.
We adopt the same topology, with the encrypted-aggregation step
performed under CKKS so that the master sees only the *sum* of
local contributions and not the individual addends.

## DLBCL lymphoma cohort

```{r}
#| echo: false
suppressPackageStartupMessages(library(homomorpheR))
data(DLBCL)
.sg  <- c("GCB", "ABC", "Type III")
.coh <- list(
  n      = nrow(DLBCL),
  deaths = sum(DLBCL$status),
  mfu    = format(round(median(DLBCL$time), 1), nsmall = 1),
  sgn    = vapply(.sg, function(s) sum(DLBCL$Subgroup == s), 1L),
  sgd    = vapply(.sg, function(s) sum(DLBCL$status[DLBCL$Subgroup == s]), 1L))
```

We use the diffuse large-B-cell lymphoma (DLBCL) cohort of
@Rosenwald2002, the same dataset that @BayleFanLou2025 use to
motivate distributed Cox estimation. The `data(DLBCL)` table
shipped with homomorpheR excludes the five patients with zero
follow-up time (following @BayleFanLou2025), leaving
`r .coh$n` patients with `r .coh$deaths` deaths over a median
follow-up of `r .coh$mfu` years. The published outcome
predictor combines five expression signatures: germinal-center B
cell, lymph node, proliferation, BMP6, and MHC class II. We model
the hazard as a function of those five signatures, stratified by
molecular subgroup (GCB, ABC, Type III) and *use the subgroup
itself as the site boundary*. This is biologically and
operationally realistic: GCB, ABC, and Type III tumors arise from
different cells of origin and tend to be diagnosed at different
referral centers. The three sites are imbalanced in size (GCB
$n=`r .coh$sgn["GCB"]`$, ABC $n=`r .coh$sgn["ABC"]`$, Type III
$n=`r .coh$sgn["Type III"]`$, with `r .coh$sgd["GCB"]`,
`r .coh$sgd["ABC"]`, and `r .coh$sgd["Type III"]` deaths
respectively), which the master/worker protocol handles
transparently.

```{r}
suppressPackageStartupMessages(library(survival))
library(homomorpheR)
data(DLBCL)

cox_data <- split(
  DLBCL[, c("time", "status", "GCB_sig", "LN_sig",
            "Prolif_sig", "BMP6", "MHC2_sig", "Subgroup")],
  DLBCL$Subgroup)
sapply(cox_data, function(df) c(n = nrow(df), events = sum(df$status)))
```

## The aggregated fit

If all data were in one place, fitting the stratified Cox model is
trivial:

```{r}
agg_model <- coxph(Surv(time, status) ~ GCB_sig + LN_sig +
                       Prolif_sig + BMP6 + MHC2_sig +
                       strata(Subgroup),
                   data = DLBCL)
agg_model
agg_model$loglik
```

The first log-likelihood is at $\beta = 0$ (the null model); the
second is at the MLE. The privacy goal: reproduce these estimates
*without* the three sites ever pooling their raw data.

## The protocol

We use the same master/worker topology as the MLE vignette: master
broadcasts $\beta$, each worker computes its local Cox partial
log-likelihood at $\beta$, encrypts it under the master's public
key, and returns the ciphertext. The master sums the encrypted
contributions homomorphically and decrypts the total. Mathematically
nothing changes from the MLE case; only the local computation
differs.

The local computation exploits a well-known feature of `coxph()`:
`iter.max = 0` returns the partial log-likelihood evaluated at the
supplied `init` *without* taking any Newton-Raphson steps.

```{r}
cph_control <- replace(coxph.control(), "iter.max", 0)

local_cox_nll <- function(data, beta) {
    fit <- tryCatch(
        coxph(Surv(time, status) ~ GCB_sig + LN_sig + Prolif_sig +
                  BMP6 + MHC2_sig,
              data    = data,
              init    = beta,
              control = cph_control),
        error = function(e) NULL)
    if (is.null(fit)) NA_real_ else -fit$loglik[1]
}
```

`tryCatch` returns `NA_real_` if the local fit blows up at extreme
$\beta$ — `homomorpheR`'s `master_aggregate()` propagates that
`NA` back to the optimizer, which simply backs off.

## Wiring up the protocol

The summed stratified Cox nLL on this cohort has magnitude
$\approx 5 \times 10^2$ — comfortably within CKKS precision at
the default scaling parameters. We lift `scaling_mod_size` from
50 to 59 and set `first_mod_size = 60` for a safety margin.

```{r}
cc <- openfhe.R::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L)
keys <- openfhe.R::key_gen(cc)

worker_gcb <- make_worker("GCB",      data = cox_data[["GCB"]],      local_fn = local_cox_nll)
worker_abc <- make_worker("ABC",      data = cox_data[["ABC"]],      local_fn = local_cox_nll)
worker_t3  <- make_worker("Type III", data = cox_data[["Type III"]], local_fn = local_cox_nll)
master     <- make_ckks_master("Master", crypto_context = cc, keypair = keys)
set_workers(master, list(worker_gcb, worker_abc, worker_t3))
```

## Iterative MLE through the encrypted protocol

We hand `stats4::mle()` a function that looks like a standard
multivariate negative log-likelihood. Each call drives one
master/worker round and returns a single decrypted scalar.

```{r}
library(stats4)

encrypted_nLL <- function(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig) {
    master_aggregate(master, c(GCB_sig, LN_sig, Prolif_sig, BMP6, MHC2_sig))
}

## Parameter order matches the coxph formula above, so the
## side-by-side comparison below is honest. BFGS uses numerical
## finite-difference gradients, the fair comparison given a
## generic optimizer driven through an encrypted channel.
fit <- mle(encrypted_nLL,
           start   = list(GCB_sig = 0, LN_sig = 0, Prolif_sig = 0,
                          BMP6    = 0, MHC2_sig = 0),
           method  = "BFGS",
           control = list(reltol = 1e-7))
summary(fit)
logLik(fit)
```

## Comparison with the aggregated fit

```{r}
summary(agg_model)
cat(sprintf("logLik(distributed encrypted): %f\n", as.numeric(logLik(fit))))
cat(sprintf("logLik(aggregated cleartext) : %f\n", agg_model$loglik[2]))
```

The estimates and standard errors agree to high precision: the
encrypted distributed fit reproduces the aggregated cleartext fit
without any site revealing its raw patient data.

```{r echo = FALSE}
mle_coefs   <- coef(fit)
coxph_coefs <- coef(agg_model)
comparison <- data.frame(
    coefficient           = names(mle_coefs),
    encrypted_distributed = unname(mle_coefs),
    aggregated_cleartext  = unname(coxph_coefs[names(mle_coefs)]),
    abs_diff              = unname(abs(mle_coefs - coxph_coefs[names(mle_coefs)]))
)
knitr::kable(comparison, digits = 9,
             caption = "Encrypted-distributed BFGS vs. aggregated-cleartext coxph()")
```

## What just happened

We took `stats4::mle()` — a generic R optimizer with no awareness
of cryptography — and handed it a function that computes its
argument through a three-party homomorphic-encryption protocol.
The optimizer ran its normal BFGS loop, asked for the value of the
negative log-likelihood at a handful of points in $\mathbb{R}^5$,
and at each point our function performed a CKKS master/worker round
across three hospital sites and returned a single decrypted scalar.

`mle()` does not know — and does not need to know — that any of
this happened. From its perspective it called a function and got
a number. The result matches the aggregated `coxph()` fit at the
precision the optimizer cares about.

## A note on the comparison

`coxph()` uses Newton-Raphson with the **analytic gradient and
Hessian** of the Cox partial log-likelihood — a scheme-specific
advantage. We use BFGS with **numerical finite-difference
gradients**, which is the fair comparison given that we're driving
a generic optimizer through an encrypted channel. CKKS itself
introduces no measurable error at the parameter sizes we use.

## What this demonstrates

1. **Existing R optimizers work over encrypted channels.**
   Privacy-preserving distributed statistics does not require
   rewriting `mle`, `optim`, `nlm`, `glm`'s fitter, MCMC samplers,
   or any other established R machinery. As long as the routine
   accepts a likelihood (or loss) as a callback, you can drop in
   an FHE-aware callback and the routine will just work.
2. **Stratified Cox regression decomposes additively** across
   strata, so the same master/worker scheme used for Poisson MLE
   (`vignette("mle")`) works verbatim for survival analysis. Only
   the local computation changes (`coxph` instead of `dpois`).
3. **CKKS handles real-valued partial log-likelihoods directly.**
   The Paillier-based vignettes in this package have to split each
   value into integer and fractional parts and rationally
   approximate the fractional part with a denominator of $2^{256}$.
   With CKKS, the protocol becomes pure ciphertext arithmetic.
4. **The master/worker classes are reusable.** The same exported
   `Site` / `Master` classes and `master_aggregate()` runner
   drive both this Cox vignette and the Poisson MLE vignette —
   only the per-worker `local_fn` differs.

## Caveats and extensions

- **Performance**: each function evaluation requires three CKKS
  encryptions, three ciphertext additions, and one decryption.
  CKKS encrypt/decrypt dominates the wall-clock cost.
- **Information leakage**: the master sees the *value* of the
  joint log-likelihood at each $\beta$, which is more information
  than individual contributions but is what `mle()` needs. To
  hide even this would require running Newton-Raphson entirely
  inside the encrypted domain — feasible with CKKS but
  considerably more complex.
- **Threshold key generation**: in a real deployment, the secret
  key would be split across the sites (n-of-n threshold) so that
  no single party — not even the master — can decrypt
  intermediate ciphertexts unilaterally. The next vignette
  (`vignette("cox-threshold")`) builds on this one and adds
  threshold key generation.
- **Beyond Cox**: any likelihood whose log decomposes additively
  across data partitions admits this exact protocol — generalized
  linear models, mixed-effects models with site-specific random
  effects, frailty survival models. The structural pattern is
  identical; only the local likelihood evaluation changes.
