---
title: "Consensus ADMM + Output Differential Privacy (Demonstration)"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Consensus ADMM + Output Differential Privacy (Demonstration)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## Introduction
This vignette is a **demonstration**, not a recommendation.
`vignette("cvxr-consensus-admm")` shows the lossless consensus-ADMM
protocol over the threshold-FHE channel: bit-identical to the
centralized CVXR fit, no single decrypter, residuals drive
convergence. The aggregator sees the full trajectory $\{z^k\}$ in
the clear after each iteration's threshold fusion.

A natural question is whether one can compose that lossless
channel with **output differential privacy** to bound the
trajectory leakage formally. The answer in this simulation is: yes,
the cryptographic protocol composes cleanly and the implementation
is one extra `rnorm()` call per site per iteration, but the
accuracy/privacy trade-off does not produce a usable point at
small $\varepsilon$ given the optimizer structure here. This
vignette runs that trade-off end-to-end.

`feedback_he_not_dp` in the package's notes records the framing
explicitly: lossless threshold-FHE is the package's primary path;
DP demonstrations show *what happens if* a user composes output DP
on top, not the recommended deployment for precision-critical
workloads.

## A brief output-DP primer

Given a query $f: \mathcal{D} \to \mathbb{R}^p$, the **Gaussian
mechanism** releases $\tilde f(D) = f(D) + \mathcal{N}(0, \sigma^2 I)$.
Sensitivity-driven $\sigma$ gives single-query
$(\varepsilon, \delta)$ DP. Multi-query composition uses **zCDP**
[Bun & Steinke 2016]: each release is $(\Delta/\sigma)^2/2$-zCDP;
$T$ releases compose linearly to $T \cdot \rho$; convert back to
$(\varepsilon, \delta)$ via $\varepsilon = \rho + 2\sqrt{\rho \log(1/\delta)}$.

## The protocol modification

The setup is identical to `vignette("cvxr-consensus-admm")`. The
only change is **noise injection at each site**: at every ADMM
iteration site $i$ adds an independent draw
$\eta_i \sim \mathcal{N}(0, \sigma^2 N \cdot I)$ to $x_i + u_i$
*before* encrypting. The encrypted noise terms sum under the joint
key, the $1/N$ scaling contracts the variance back to $\sigma^2$
per coordinate, and the recovered $z$ has noise
$\mathcal{N}(0, \sigma^2 I)$.

Distributing the noise across sites (rather than centralizing it
at the aggregator) means:

1. No single point ever holds the noiseless value. Even if the
   aggregator is compromised it sees only encrypted *noised*
   contributions until the final fusion.
2. The trust model matches `cox-threshold-dp`: each site is its
   own randomness boundary.

Both are properties of the code below rather than of this
paragraph: `site_contribution_dp()` draws the noise *and* encrypts
inside the site, so `encrypted_consensus_dp()` receives ciphertexts
and nothing else. Had the aggregator done the noising — reading
each site's $x_i + u_i$ and adding a draw itself — every sentence
above would be false while the numbers came out identical.

## The stopping rule changes

Residual-based convergence checks are not meaningful under noise:
the residuals never shrink below the per-iteration noise floor.
So $T$ is **fixed in advance** rather than discovered.

That is forced twice over. Even without noise, a residual-based
stop is an adaptive, data-dependent decision — the iteration at
which you halt is a function of the records — and $T$ is exactly
the quantity that multiplies the privacy budget below. A protocol
that reads $T$ off the cohort and then bills $T$ releases against
$\varepsilon$ has left the accounting open at the step that
defines it. The next section picks $\rho$ and $T$ without touching
the cohort at all.

The $\sigma = 0$ row of the table below verifies that the protocol
with the noise mechanism turned off lands on the lossless ADMM
fit at that pre-committed $T$.

## Setup

```{r}
suppressPackageStartupMessages({
    library(homomorpheR)
    library(CVXR)
    library(S7)
})

N   <- 3L
p   <- 4L
lam <- 1
```

```{r}
build_local_problem <- function(X_i, y_i, rho_val) {
    x  <- Variable(p)
    zp <- Parameter(p)
    up <- Parameter(p)
    y_signs <- 2 * y_i - 1
    margins <- -y_signs * (X_i %*% x)
    local_loss <- sum(logistic(margins)) +
                  (lam / (2 * N)) * sum_squares(x)
    augmented  <- (rho_val / 2) * sum_squares(x - zp + up)
    prob <- Problem(Minimize(local_loss + augmented))
    value(zp) <- rep(0, p); value(up) <- rep(0, p)
    list(prob = prob, x = x, zp = zp, up = up)
}

## Inherits homomorpheR's abstract `Site` (which supplies `name` and the
## `state` environment), so it can take part in threshold key generation
## and keep its own share.
ConsensusSite <- new_class("ConsensusSite",
    parent     = homomorpheR::Site,
    properties = list(n = class_integer))

make_consensus_site <- function(name, X_i, y_i, rho_val) {
    st        <- new.env(parent = emptyenv())
    st$X      <- X_i
    st$y      <- y_i
    built     <- build_local_problem(X_i, y_i, rho_val)
    st$prob   <- built$prob
    st$x_var  <- built$x
    st$zp     <- built$zp
    st$up     <- built$up
    st$x_curr <- rep(0, ncol(X_i))
    st$u_curr <- rep(0, ncol(X_i))
    ConsensusSite(name = name, n = nrow(X_i), state = st)
}

local_update <- function(site, z_curr) {
    st <- site@state
    value(st$zp) <- z_curr
    value(st$up) <- st$u_curr
    suppressMessages(suppressWarnings(psolve(st$prob, solver = "CLARABEL")))
    if (!status(st$prob) %in% c("optimal", "optimal_inaccurate"))
        stop("Local CVXR solve at ", site@name, " did not reach optimal status.")
    st$x_curr <- as.numeric(value(st$x_var))
    invisible(st$x_curr)
}
```

## Simulated cohort

```{r}
set.seed(20260412)
n_per_site <- c(500L, 1000L, 1500L)
beta_true  <- c(intercept = -0.5, age = 0.4, bmi = -0.3, sex = 0.6)

make_site_data <- function(n) {
    X  <- cbind(1, rnorm(n), rnorm(n), rbinom(n, 1, 0.5))
    pr <- plogis(as.numeric(X %*% beta_true))
    y  <- as.integer(runif(n) < pr)
    list(X = X, y = y)
}
site_data <- lapply(n_per_site, make_site_data)
```

## Choosing $\rho$ and $T$ without touching the cohort

The lossless ADMM vignette tunes $\rho$ by sweeping candidates
through the encrypted channel, *on the real data*. Under output DP
that sweep is not available to us. Every read of the cohort is a
release that has to be paid for, and a sweep is a large one: it
solves at every site, at every candidate, at every iteration, and
what comes out — the chosen $\rho$, and above all $T$ — is a
function of the records. Billing $T \cdot N$ Gaussian releases
while $T$ itself was read off the data is not an accounting, it is
an accounting with its own parameter left outside.

There are two honest ways to close that. One is to **pay for the
selection**: report-noisy-max or the exponential mechanism over
the grid, with negative iteration count as utility. That is the
principled route, but it needs a sensitivity bound for "iterations
to convergence" under a one-record change, and we do not have one
— it would trade this gap for a weaker one.

The other is to make the selection **data-independent**, which is
what DP deployments in practice do, and what we do here. The sweep
runs on a **surrogate cohort** built only from facts the protocol
already treats as public — how many sites there are, roughly how
large they are, and the covariate schema — together with nominal
effect sizes written into the analysis plan and claimed by nobody
to be correct. No record from any site enters it.

Two consequences worth being explicit about. First, this sweep
needs **no encrypted channel**: there is no secret in it to
protect. That is the exact opposite of the situation in
`vignette("cvxr-consensus-admm")`, where the sweep goes through
the threshold channel *because* it touches real data. Cleartext
here is a conclusion, not a convenience. Second, the cost is real
but it is paid in **utility, not privacy**: a surrogate that
misjudges the curvature returns a $\rho$ the cohort would not have
picked and a $T$ shorter than it needed. There is no way to check
that against the cohort without spending budget, so the $T$ below
is a pre-commitment — and since the budget grows linearly in $T$,
pre-committing generously is not free either.

```{r}
tol      <- 1e-3
max_iter <- 60L

## Public design facts: three sites of these approximate sizes, four
## covariates of these types. Nominal effect sizes, not the cohort's.
beta_nominal   <- c(0, 0.5, 0.5, 0.5)
surrogate_seed <- 20260413L

set.seed(surrogate_seed)
surrogate_data <- lapply(n_per_site, function(n) {
    X  <- cbind(1, rnorm(n), rnorm(n), rbinom(n, 1, 0.5))
    pr <- plogis(as.numeric(X %*% beta_nominal))
    list(X = X, y = as.integer(runif(n) < pr))
})

sweep_one_rho <- function(cohort, rho_val) {
    built <- lapply(cohort,
                    function(s) build_local_problem(s$X, s$y, rho_val))
    x_curr <- u_curr <- replicate(N, rep(0, p), simplify = FALSE)
    z      <- rep(0, p)
    k_conv <- NA_integer_
    for (k in seq_len(max_iter)) {
        for (i in seq_len(N)) {
            value(built[[i]]$zp) <- z
            value(built[[i]]$up) <- u_curr[[i]]
            suppressMessages(suppressWarnings(
                psolve(built[[i]]$prob, solver = "CLARABEL")))
            x_curr[[i]] <- as.numeric(value(built[[i]]$x))
        }
        z_prev <- z
        z <- Reduce(`+`, Map(`+`, x_curr, u_curr)) / N
        for (i in seq_len(N)) u_curr[[i]] <- u_curr[[i]] + x_curr[[i]] - z
        pri <- sqrt(sum(vapply(seq_len(N),
            function(i) sum((x_curr[[i]] - z)^2), 0)) / N)
        dua <- rho_val * sqrt(sum((z - z_prev)^2))
        if (pri < tol && dua < tol) { k_conv <- k; break }
    }
    data.frame(rho = rho_val,
               iters = if (is.na(k_conv)) max_iter else k_conv,
               converged = !is.na(k_conv))
}

rho_grid  <- c(10, 20, 50, 100, 500)
rho_sweep <- do.call(rbind,
                     lapply(rho_grid,
                            function(r) sweep_one_rho(surrogate_data, r)))
knitr::kable(rho_sweep,
             caption = "Consensus-ADMM convergence on the surrogate cohort")

converged_rows <- rho_sweep[rho_sweep$converged, ]
if (nrow(converged_rows) == 0L)
    stop("No rho in the grid converged within max_iter on the surrogate.")

rho_chosen <- converged_rows$rho[which.min(converged_rows$iters)]
T_fixed    <- converged_rows$iters[converged_rows$rho == rho_chosen]
cat(sprintf("Pre-committed rho = %g, T = %d (surrogate cohort).\n",
            rho_chosen, T_fixed))
```

These two numbers are now constants of the protocol. The DP-ADMM
loop below runs for exactly $T = `r T_fixed`$ iterations
regardless of residuals, and nothing downstream is allowed to
revisit them in light of what the cohort does.

## Threshold-FHE setup

```{r}
cc <- openfhe.R::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L,
                           features             = c(openfhe.R::Feature$MULTIPARTY))
```

The DP version of the consensus step. The only change from the
lossless ADMM vignette's `encrypted_consensus()` is the
`rnorm(p, ..., sd = sigma * sqrt(Nv))` term inside the per-site
loop:

```{r}
## Site-side, and this is the whole point of the DP variant: the site
## draws its own noise, adds it, and encrypts with the public bundle it
## was handed at setup -- all before anything leaves. The noiseless
## x_i + u_i exists nowhere but here, which is what "each site is its
## own randomness boundary" has to mean.
site_contribution_dp <- function(site, sigma, Nv) {
    st <- site@state
    if (is.null(st$params))
        stop("Site ", site@name, " has no public parameters.")
    noised <- st$x_curr + st$u_curr + rnorm(p, mean = 0, sd = sigma * sqrt(Nv))
    encrypt_under(st$params, noised)
}

## Aggregator-side: sum the ciphertexts, scale, threshold-decrypt. The
## 1/N scaling contracts the summed noise variance back to sigma^2.
encrypted_consensus_dp <- function(threshold_master, sites, sigma) {
    Nv  <- length(sites)
    cts <- lapply(sites, site_contribution_dp, sigma = sigma, Nv = Nv)
    ct_avg <- Reduce(`+`, cts) * (1 / Nv)
    master_decrypt(threshold_master, ct_avg, len = p)
}
```

## The DP-ADMM loop

```{r}
run_dp_admm <- function(sigma, T_iter = T_fixed, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    ## The sites exist first: the joint public key is built from them,
    ## each keeping the share it generates.
    sites <- list(
        make_consensus_site("Site 1", site_data[[1]]$X, site_data[[1]]$y, rho_chosen),
        make_consensus_site("Site 2", site_data[[2]]$X, site_data[[2]]$y, rho_chosen),
        make_consensus_site("Site 3", site_data[[3]]$X, site_data[[3]]$y, rho_chosen))
    master <- make_threshold_master("Aggregator",
                                    crypto_context = cc,
                                    sites          = sites)

    z_curr <- rep(0, p)
    z_hist <- matrix(NA_real_, nrow = T_iter, ncol = p,
                     dimnames = list(NULL, names(beta_true)))
    for (k in seq_len(T_iter)) {
        for (s in sites) local_update(s, z_curr)
        z_curr <- encrypted_consensus_dp(master, sites, sigma)
        for (s in sites) {
            s@state$u_curr <- s@state$u_curr + (s@state$x_curr - z_curr)
        }
        z_hist[k, ] <- z_curr
    }
    list(z = z_curr, z_hist = z_hist)
}
```

## Centralized CVXR fit

This fit, and every `max_dev` column derived from it below, is a
**simulation diagnostic and not part of the protocol**. It pools
the raw data, which the deployed protocol never does, and it is
not charged to the budget because it is not released — it exists
so that this document can show you how far the noise moved the
answer. A real deployment has no access to it, which is precisely
why $\rho$ and $T$ had to be pre-committed above.

```{r}
X_pooled  <- do.call(rbind, lapply(site_data, `[[`, "X"))
y_pooled  <- unlist(lapply(site_data, `[[`, "y"))
beta_var  <- Variable(p)
y_signs_p <- 2 * y_pooled - 1
margins_p <- -y_signs_p * (X_pooled %*% beta_var)
suppressMessages(suppressWarnings(
    psolve(Problem(Minimize(sum(logistic(margins_p)) +
                            (lam / 2) * sum_squares(beta_var))),
           solver = "CLARABEL")))
beta_central <- as.numeric(value(beta_var))
names(beta_central) <- names(beta_true)
```

## The $\sigma$ sweep

Six $\sigma$ values from zero to one. The $\sigma = 0$ row is the
sanity check that the DP mechanism is a no-op when off.

```{r}
sigma_grid    <- c(0, 1e-4, 1e-3, 1e-2, 1e-1, 1)
sweep_results <- vector("list", length(sigma_grid))
for (j in seq_along(sigma_grid)) {
    sweep_results[[j]] <- run_dp_admm(sigma = sigma_grid[j], seed = 100L + j)
}
names(sweep_results) <- sprintf("sigma=%.0e", sigma_grid)
```

## $\sigma = 0$ sanity check

```{r}
clean_dev <- max(abs(sweep_results[[1]]$z - beta_central))
agree_tol <- 10 * tol
if (clean_dev > agree_tol)
    stop("DP-ADMM at sigma = 0 disagrees with the centralized fit.")
cat(sprintf("Sigma = 0 max coefficient deviation: %.2e (tol %.0e)\n",
            clean_dev, agree_tol))
```

## Summary table

```{r}
summary_df <- do.call(rbind, lapply(seq_along(sigma_grid), function(j) {
    z <- sweep_results[[j]]$z
    data.frame(sigma     = sigma_grid[j],
               intercept = z[1],
               age       = z[2],
               bmi       = z[3],
               sex       = z[4],
               max_dev   = max(abs(z - beta_central)))
}))
central_row <- data.frame(sigma = NA, intercept = beta_central[1],
                         age = beta_central[2], bmi = beta_central[3],
                         sex = beta_central[4], max_dev = 0)
summary_table <- rbind(summary_df, central_row)
rownames(summary_table) <- c(sprintf("sigma=%g", sigma_grid), "centralized")
knitr::kable(summary_table, digits = 6,
             caption = "DP-ADMM coefficients vs centralized CVXR")
```

The deviation grows roughly linearly with $\sigma$ in the
small-$\sigma$ regime and then leaves the linear band — at
$\sigma = 1$ the recovered coefficients drift far from the
centralized fit. The protocol is faithfully running the same
algorithm at every $\sigma$; the deterioration is what the noise
mechanism does to the optimization, not what the cryptographic
channel does.

## Privacy budget

Per-iteration zCDP: $\rho_{\text{iter}} = (\Delta/\sigma)^2/2$ per
*site* per coordinate. Across $T = `r T_fixed`$ iterations and
$N = `r N`$ sites, total $\rho = T \cdot N \cdot (\Delta/\sigma)^2/2$.
Convert to $(\varepsilon, \delta)$ via the standard formula.

```{r}
zcdp_to_eps <- function(rho, delta = 1e-5) rho + 2 * sqrt(rho * log(1 / delta))

budget <- data.frame(sigma = sigma_grid[sigma_grid > 0])
budget$rho_total                  <- T_fixed * N * (1 / budget$sigma)^2 / 2
budget$epsilon_at_delta_1e_minus_5 <- zcdp_to_eps(budget$rho_total)
knitr::kable(budget, digits = 4,
             caption = "zCDP composition; sensitivity Delta = 1, delta = 1e-5")
```

This total is the whole of the procedure's data-dependent
exposure, which is the point of having chosen $\rho$ and $T$ off
the surrogate: the selection step contributes nothing to it,
because it read nothing. The one quantity still taken on faith is
the sensitivity $\Delta = 1$, and that is flagged as a placeholder
in the limitations below — an honest budget with one declared
placeholder, rather than a budget whose own $T$ came from an
unbilled read.

Read `epsilon_at_delta_1e_minus_5`. The smallest $\varepsilon$
attained — at $\sigma = 1$ where the fit is already broken — is
still in the dozens. Whether large $\varepsilon$ is acceptable is a
use-case decision; we report the numbers and stop.

## What this demonstrates

1. **The cryptographic protocol composes cleanly with output
   DP.** At $\sigma = 0$ the DP-ADMM protocol reproduces the
   lossless ADMM fit. The DP layer is one `rnorm()` call per site
   per iteration and zero new homomorpheR machinery.
2. **The accuracy/privacy trade-off is reported end-to-end.** Six
   $\sigma$ values, one fixed-$T$ DP-ADMM run each, summary table
   showing the recovered coefficients and the deviation from the
   centralized fit.
3. **At noise scales the optimizer tolerates, $\varepsilon$ is
   large.** zCDP composition over $T \cdot N$ Gaussian releases
   inflates the budget; tightening it requires either fewer
   iterations, tighter sensitivity, or subsampling amplification —
   all out of scope for this demonstration.
4. **Hyperparameter selection is part of the budget, or it is
   outside the guarantee.** The step that picks $\rho$ and $T$
   looks like setup rather than analysis, which is exactly why it
   escapes scrutiny; here it also *sets the multiplier on the
   budget*. Making it data-independent keeps the accounting closed
   and costs utility instead. Composing a cryptographic channel
   with a DP guarantee does not make that step go away — encryption
   hides intermediates, not choices.

## Limitations
- **Tight sensitivity bounds** for the consensus update. $\Delta = 1$
  is a placeholder.
- **Paying for the selection instead of avoiding it.** Report-noisy-max
  or the exponential mechanism over the $\rho$ grid would let the
  cohort inform the choice for a declared cost. That needs a
  sensitivity bound on iterations-to-convergence under a one-record
  change, which we do not have.
- **Surrogate quality.** A surrogate that misjudges the cohort's
  curvature costs accuracy — a worse $\rho$, or a $T$ too short to
  converge — and there is no way to detect that from inside the
  protocol without spending budget.
- **Adaptive noise** schedules across iterations.
- **Subsampling amplification** of the DP accountant.
- **Malicious-site protection.** A hostile site can corrupt either
  its $(x_i + u_i)$ or its noise draw; out of scope.

Readers who need the lossless fit should use
`vignette("cvxr-consensus-admm")` and accept the trajectory
exposure that comes with it.

## References

- Bun & Steinke (2016). *Concentrated Differential Privacy.* TCC.
- Cyffers, Bellet & Upadhyay (2023). *Muffliato:
  Peer-to-Peer Privacy Amplification for Decentralized
  Optimisation and Averaging.* The DP-ADMM analytic companion to
  this empirical demonstration.
