---
title: "Consensus ADMM + Output Differential Privacy (Demonstration)"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Consensus ADMM + Output Differential Privacy (Demonstration)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F, purl=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
## The expensive chunks are gated `eval = RECOMPUTE` and are shown but
## not run when the vignette builds; their results are read from
## data(cvxr_admm_dp_results), produced by
## data-raw/cvxr_admm_dp_results.R from these same chunks.
## HOMOMORPHER_RECOMPUTE=true runs them for real.
if (!exists("RECOMPUTE"))
    RECOMPUTE <- isTRUE(as.logical(Sys.getenv("HOMOMORPHER_RECOMPUTE", "FALSE")))
data(cvxr_admm_dp_results, package = "homomorpheR")
```

## Introduction
This vignette is a demonstration. It fits an L2-regularized
logistic regression across $N$ sites by consensus ADMM under
threshold FHE, the protocol `vignette("cvxr-cox-lasso-dlbcl")` uses
for the Cox lasso. Each site solves its local problem with CVXR,
and the aggregator decrypts only the average of the sites'
encrypted vectors.

## The problem and its consensus split

With $N$ sites, local data $(X_i, y_i)$ at site $i$, and a shared
coefficient $x \in \mathbb{R}^p$, the global problem is

$$
\min_{x \in \mathbb{R}^p}
\sum_{i=1}^{N} \ell_i(x; X_i, y_i)
+ \frac{\lambda}{2}\, \lVert x \rVert_2^2
$$

where $\ell_i$ is the logistic loss on site $i$. The standard
consensus split (Boyd, Parikh, Chu, Peleato, Eckstein, 2011)
introduces local copies $x_i \in \mathbb{R}^p$ and a single global
consensus $z \in \mathbb{R}^p$:

$$
\min_{\{x_i\}, z}
\sum_i \ell_i(x_i; X_i, y_i) + \frac{\lambda}{2}\, \lVert z \rVert_2^2
\quad \text{s.t.}\quad x_i = z, \ \forall i.
$$

The augmented-Lagrangian iteration is:

$$
\begin{aligned}
x_i^{k+1} &= \arg\min_{x_i}\, \ell_i(x_i) + \frac{\lambda}{2N}\lVert x_i\rVert_2^2 + \frac{\rho}{2}\lVert x_i - z^k + u_i^k\rVert_2^2, \\
z^{k+1} &= \frac{1}{N}\sum_i (x_i^{k+1} + u_i^k), \\
u_i^{k+1} &= u_i^k + (x_i^{k+1} - z^{k+1}).
\end{aligned}
$$

The $x$-update is local at each site; the $z$-update is the
consensus average that has to traverse the encrypted channel; the
$u$-update is local again. Only the $z$-update needs cryptography.

## Adding noise

Here we ask what happens if each site also adds noise, so that the
released iterates satisfy **output differential privacy**. Adding
the noise takes one extra `rnorm()` call per site per iteration.
With this optimizer, the fits are poor at any noise level that
gives a small $\varepsilon$. The vignette runs the fits and reports
the numbers.

Threshold FHE without noise remains the package's main path. This
vignette shows what adding output DP on top costs.

## A brief output-DP primer

Given a query $f: \mathcal{D} \to \mathbb{R}^p$, the **Gaussian
mechanism** releases $\tilde f(D) = f(D) + \mathcal{N}(0, \sigma^2 I)$.
Sensitivity-driven $\sigma$ gives single-query
$(\varepsilon, \delta)$ DP. Multi-query composition uses **zCDP**
[@BunSteinke2016]: each release is $(\Delta/\sigma)^2/2$-zCDP;
$T$ releases compose linearly to $T \cdot \rho$; convert back to
$(\varepsilon, \delta)$ via $\varepsilon = \rho + 2\sqrt{\rho \log(1/\delta)}$.

## The protocol modification

Without noise, site $i$ encrypts $x_i + u_i$ and the aggregator
decrypts only their average $z$. The change here is **noise
injection at each site**: at every ADMM
iteration site $i$ adds an independent draw
$\eta_i \sim \mathcal{N}(0, \sigma^2 N \cdot I)$ to $x_i + u_i$
*before* encrypting. The encrypted noise terms sum under the joint
key, the $1/N$ scaling contracts the variance back to $\sigma^2$
per coordinate, and the recovered $z$ has noise
$\mathcal{N}(0, \sigma^2 I)$.

Each site adds its own noise, rather than the aggregator adding it.
So:

1. Only site $i$ sees its noiseless $x_i + u_i$. The aggregator
   sees encrypted noisy contributions and, after decryption, their
   noisy average.
2. As in `cox-threshold-dp`, each site draws its own noise.

In the code below, `site_contribution_dp()` adds the noise and
encrypts inside the site, so `encrypted_consensus_dp()` receives
only encrypted values. If the aggregator added the noise instead,
the numbers would be the same, but the aggregator would see each
site's noiseless $x_i + u_i$.

## The stopping rule changes

Residual-based convergence checks are not meaningful under noise:
the residuals never shrink below the per-iteration noise floor.
So $T$ is **fixed in advance** rather than discovered.

There is a second reason. Even without noise, stopping on the
residuals makes $T$ depend on the data, and $T$ multiplies the
privacy budget below. Choosing $T$ from the cohort would be a
release the budget does not count. The next section picks $\rho$
and $T$ without using the cohort.

The $\sigma = 0$ row of the table below checks that, with no
noise, the protocol reaches the centralized fit in that fixed $T$.

## Setup

```{r}
suppressPackageStartupMessages({
    library(homomorpheR)
    library(CVXR)
    library(S7)
})

N   <- 3L
p   <- 4L
lam <- 1
```

```{r}
build_local_problem <- function(X_i, y_i, rho_val) {
    x  <- Variable(p)
    zp <- Parameter(p)
    up <- Parameter(p)
    y_signs <- 2 * y_i - 1
    margins <- -y_signs * (X_i %*% x)
    local_loss <- sum(logistic(margins)) +
                  (lam / (2 * N)) * sum_squares(x)
    augmented  <- (rho_val / 2) * sum_squares(x - zp + up)
    prob <- Problem(Minimize(local_loss + augmented))
    value(zp) <- rep(0, p); value(up) <- rep(0, p)
    list(prob = prob, x = x, zp = zp, up = up)
}

## Inherits homomorpheR's abstract `Site` (which supplies `name` and the
## `state` environment), so it can take part in threshold key generation
## and keep its own share.
ConsensusSite <- new_class("ConsensusSite",
    parent     = homomorpheR::Site,
    properties = list(n = class_integer))

make_consensus_site <- function(name, X_i, y_i, rho_val) {
    st        <- new.env(parent = emptyenv())
    st$X      <- X_i
    st$y      <- y_i
    built     <- build_local_problem(X_i, y_i, rho_val)
    st$prob   <- built$prob
    st$x_var  <- built$x
    st$zp     <- built$zp
    st$up     <- built$up
    st$x_curr <- rep(0, ncol(X_i))
    st$u_curr <- rep(0, ncol(X_i))
    ConsensusSite(name = name, n = nrow(X_i), state = st)
}

local_update <- function(site, z_curr) {
    st <- site@state
    value(st$zp) <- z_curr
    value(st$up) <- st$u_curr
    suppressMessages(suppressWarnings(psolve(st$prob, solver = "CLARABEL")))
    if (!status(st$prob) %in% c("optimal", "optimal_inaccurate"))
        stop("Local CVXR solve at ", site@name, " did not reach optimal status.")
    st$x_curr <- as.numeric(value(st$x_var))
    invisible(st$x_curr)
}
```

## Simulated cohort

```{r}
set.seed(20260412)
n_per_site <- c(500L, 1000L, 1500L)
beta_true  <- c(intercept = -0.5, age = 0.4, bmi = -0.3, sex = 0.6)

make_site_data <- function(n) {
    X  <- cbind(1, rnorm(n), rnorm(n), rbinom(n, 1, 0.5))
    pr <- plogis(as.numeric(X %*% beta_true))
    y  <- as.integer(runif(n) < pr)
    list(X = X, y = y)
}
site_data <- lapply(n_per_site, make_site_data)
```

## Choosing $\rho$ and $T$ without touching the cohort

Without noise, one could tune $\rho$ by running candidate values
through the encrypted channel on the real data. Under output
DP that sweep is itself a release. It solves at every site, for
every candidate, at every iteration, and its outputs (the chosen
$\rho$ and especially $T$) depend on the records. The budget below
counts $T$ Gaussian releases. If $T$ were chosen from the
data, that choice would not be counted.

There are two ways to handle this. One is to pay for the selection
with a DP selection mechanism, such as report-noisy-max or the
exponential mechanism over the grid, with negative iteration count
as the utility. That needs a sensitivity bound for iterations to
convergence under a one-record change, and we do not have one.

The other is to choose without the data, which is what we do. The
sweep runs on a **surrogate cohort** built only from facts the
protocol already treats as public: the number of sites, their
approximate sizes, and the covariate schema. The effect sizes are
nominal values fixed in the analysis plan. No record from any site
enters it.

This has two consequences. First, the sweep needs no encryption,
because it uses no private data. Second, the cost is in accuracy rather than privacy. A
surrogate that misjudges the curvature gives a $\rho$ the cohort
would not have picked, and possibly a $T$ that is too short. We
cannot check this against the cohort without spending budget. Since
the budget grows linearly in $T$, a generous $T$ is not free either.

```{r eval=RECOMPUTE}
tol      <- 1e-3
max_iter <- 60L

## Public design facts: three sites of these approximate sizes, four
## covariates of these types. Nominal effect sizes, not the cohort's.
beta_nominal   <- c(0, 0.5, 0.5, 0.5)
surrogate_seed <- 20260413L

set.seed(surrogate_seed)
surrogate_data <- lapply(n_per_site, function(n) {
    X  <- cbind(1, rnorm(n), rnorm(n), rbinom(n, 1, 0.5))
    pr <- plogis(as.numeric(X %*% beta_nominal))
    list(X = X, y = as.integer(runif(n) < pr))
})

sweep_one_rho <- function(cohort, rho_val) {
    built <- lapply(cohort,
                    function(s) build_local_problem(s$X, s$y, rho_val))
    x_curr <- u_curr <- replicate(N, rep(0, p), simplify = FALSE)
    z      <- rep(0, p)
    k_conv <- NA_integer_
    for (k in seq_len(max_iter)) {
        for (i in seq_len(N)) {
            value(built[[i]]$zp) <- z
            value(built[[i]]$up) <- u_curr[[i]]
            suppressMessages(suppressWarnings(
                psolve(built[[i]]$prob, solver = "CLARABEL")))
            x_curr[[i]] <- as.numeric(value(built[[i]]$x))
        }
        z_prev <- z
        z <- Reduce(`+`, Map(`+`, x_curr, u_curr)) / N
        for (i in seq_len(N)) u_curr[[i]] <- u_curr[[i]] + x_curr[[i]] - z
        pri <- sqrt(sum(vapply(seq_len(N),
            function(i) sum((x_curr[[i]] - z)^2), 0)) / N)
        dua <- rho_val * sqrt(sum((z - z_prev)^2))
        if (pri < tol && dua < tol) { k_conv <- k; break }
    }
    data.frame(rho = rho_val,
               iters = if (is.na(k_conv)) max_iter else k_conv,
               converged = !is.na(k_conv))
}

rho_grid  <- c(10, 20, 50, 100, 500)
rho_sweep <- do.call(rbind,
                     lapply(rho_grid,
                            function(r) sweep_one_rho(surrogate_data, r)))
knitr::kable(rho_sweep,
             caption = "Consensus-ADMM convergence on the surrogate cohort")

converged_rows <- rho_sweep[rho_sweep$converged, ]
if (nrow(converged_rows) == 0L)
    stop("No rho in the grid converged within max_iter on the surrogate.")

rho_chosen <- converged_rows$rho[which.min(converged_rows$iters)]
T_fixed    <- converged_rows$iters[converged_rows$rho == rho_chosen]
cat(sprintf("Pre-committed rho = %g, T = %d (surrogate cohort).\n",
            rho_chosen, T_fixed))
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
tol        <- cvxr_admm_dp_results$tol
rho_chosen <- cvxr_admm_dp_results$rho_chosen
T_fixed    <- cvxr_admm_dp_results$T_fixed
knitr::kable(cvxr_admm_dp_results$rho_sweep,
             caption = "Consensus-ADMM convergence on the surrogate cohort")
cat(sprintf("Pre-committed rho = %g, T = %d (surrogate cohort).\n",
            rho_chosen, T_fixed))
```

These two values are now fixed. The DP-ADMM loop below runs
exactly $T = `r T_fixed`$ iterations, whatever the residuals do.

## Threshold-FHE setup

```{r eval=RECOMPUTE}
cc <- openfhe.R::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L,
                           features             = c(openfhe.R::Feature$MULTIPARTY))
```

The DP version of the consensus step. The only change from the
lossless ADMM vignette's `encrypted_consensus()` is the
`rnorm(p, ..., sd = sigma * sqrt(Nv))` term inside the per-site
loop:

```{r}
## Site-side: the site draws its own noise, adds it, and encrypts with
## the public parameters it received at setup, all before anything
## leaves the site. The noiseless x_i + u_i never leaves.
site_contribution_dp <- function(site, sigma, Nv) {
    st <- site@state
    noised <- st$x_curr + st$u_curr + rnorm(p, mean = 0, sd = sigma * sqrt(Nv))
    encrypt(site, noised)
}

## Aggregator-side: sum the encrypted values, scale, threshold-decrypt. The
## 1/N scaling contracts the summed noise variance back to sigma^2.
encrypted_consensus_dp <- function(threshold_master, sites, sigma) {
    Nv  <- length(sites)
    cts <- lapply(sites, site_contribution_dp, sigma = sigma, Nv = Nv)
    ct_avg <- Reduce(`+`, cts) * (1 / Nv)
    decrypt(threshold_master, ct_avg, len = p)
}
```

## The DP-ADMM loop

```{r}
run_dp_admm <- function(sigma, T_iter = T_fixed, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    ## The sites exist first: the joint public key is built from them,
    ## each keeping the share it generates.
    sites <- list(
        make_consensus_site("Site 1", site_data[[1]]$X, site_data[[1]]$y, rho_chosen),
        make_consensus_site("Site 2", site_data[[2]]$X, site_data[[2]]$y, rho_chosen),
        make_consensus_site("Site 3", site_data[[3]]$X, site_data[[3]]$y, rho_chosen))
    master <- make_threshold_master("Aggregator",
                                    crypto_context = cc,
                                    sites          = sites)

    z_curr <- rep(0, p)
    z_hist <- matrix(NA_real_, nrow = T_iter, ncol = p,
                     dimnames = list(NULL, names(beta_true)))
    for (k in seq_len(T_iter)) {
        for (s in sites) local_update(s, z_curr)
        z_curr <- encrypted_consensus_dp(master, sites, sigma)
        for (s in sites) {
            s@state$u_curr <- s@state$u_curr + (s@state$x_curr - z_curr)
        }
        z_hist[k, ] <- z_curr
    }
    list(z = z_curr, z_hist = z_hist)
}
```

## Centralized CVXR fit

This fit pools the raw data, so it is not part of the protocol. We
use it only to measure how far the noise moves the answer, in the
`max_dev` column below. It is not released, so it is not charged
to the budget. A real deployment has no such reference, which is
why $\rho$ and $T$ were fixed in advance.

```{r}
X_pooled  <- do.call(rbind, lapply(site_data, `[[`, "X"))
y_pooled  <- unlist(lapply(site_data, `[[`, "y"))
beta_var  <- Variable(p)
y_signs_p <- 2 * y_pooled - 1
margins_p <- -y_signs_p * (X_pooled %*% beta_var)
suppressMessages(suppressWarnings(
    psolve(Problem(Minimize(sum(logistic(margins_p)) +
                            (lam / 2) * sum_squares(beta_var))),
           solver = "CLARABEL")))
beta_central <- as.numeric(value(beta_var))
names(beta_central) <- names(beta_true)
```

## The $\sigma$ sweep

Six $\sigma$ values from zero to one. The $\sigma = 0$ row checks
that the protocol without noise matches the centralized fit.

```{r eval=RECOMPUTE}
sigma_grid    <- c(0, 1e-4, 1e-3, 1e-2, 1e-1, 1)
sweep_results <- vector("list", length(sigma_grid))
for (j in seq_along(sigma_grid)) {
    sweep_results[[j]] <- run_dp_admm(sigma = sigma_grid[j], seed = 100L + j)
}
names(sweep_results) <- sprintf("sigma=%.0e", sigma_grid)
```

## $\sigma = 0$ sanity check

```{r eval=RECOMPUTE}
clean_dev <- max(abs(sweep_results[[1]]$z - beta_central))
agree_tol <- 10 * tol
if (clean_dev > agree_tol)
    stop("DP-ADMM at sigma = 0 disagrees with the centralized fit.")
cat(sprintf("Sigma = 0 max coefficient deviation: %.2e (tol %.0e)\n",
            clean_dev, agree_tol))
```

## Summary table

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
cat(sprintf("Sigma = 0 max coefficient deviation: %.2e (tol %.0e)\n",
            cvxr_admm_dp_results$clean_dev, 10 * tol))
```

```{r eval=RECOMPUTE}
summary_df <- do.call(rbind, lapply(seq_along(sigma_grid), function(j) {
    z <- sweep_results[[j]]$z
    data.frame(sigma     = sigma_grid[j],
               intercept = z[1],
               age       = z[2],
               bmi       = z[3],
               sex       = z[4],
               max_dev   = max(abs(z - beta_central)))
}))
central_row <- data.frame(sigma = NA, intercept = beta_central[1],
                         age = beta_central[2], bmi = beta_central[3],
                         sex = beta_central[4], max_dev = 0)
summary_table <- rbind(summary_df, central_row)
rownames(summary_table) <- c(sprintf("sigma=%g", sigma_grid), "centralized")
knitr::kable(summary_table, digits = 6,
             caption = "DP-ADMM coefficients vs centralized CVXR")
```

```{r eval=RECOMPUTE, echo=FALSE}
cvxr_admm_dp_results <- list(tol           = tol,
                             rho_sweep     = rho_sweep,
                             rho_chosen    = rho_chosen,
                             T_fixed       = T_fixed,
                             sigma_grid    = sigma_grid,
                             clean_dev     = clean_dev,
                             summary_table = summary_table)
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
sigma_grid <- cvxr_admm_dp_results$sigma_grid
knitr::kable(cvxr_admm_dp_results$summary_table, digits = 6,
             caption = "DP-ADMM coefficients vs centralized CVXR")
```

```{r echo=FALSE, purl=FALSE}
dev_at <- function(s) with(cvxr_admm_dp_results$summary_table,
                           formatC(max_dev[which(sigma == s)],
                                   format = "fg", digits = 2))
```

The deviation grows with $\sigma$: it is `r dev_at(1e-4)` at
$\sigma = 10^{-4}$, `r dev_at(0.01)` at $\sigma = 0.01$, and
`r dev_at(0.1)` at $\sigma = 0.1$. At $\sigma = 1$ it is
`r dev_at(1)`, and the coefficients are far from the centralized
fit. The algorithm is the same at every $\sigma$; the
loss comes from the noise, not from the encryption.

## Privacy budget

Each iteration releases one noisy average $z$, with noise
$\mathcal{N}(0, \sigma^2 I)$. Let $\Delta$ be the largest change in
$z$ (L2 norm over all $p$ coordinates) from adding or removing one
record. Each record sits at one site, and the aggregator decrypts
only the average, so one iteration is a single Gaussian release with
$\rho_{\text{iter}} = (\Delta/\sigma)^2/2$. Over $T = `r T_fixed`$
iterations the total is $\rho = T \cdot (\Delta/\sigma)^2/2$,
whatever the number of sites. This is the same accounting as
`cox-threshold-dp`. Convert to $(\varepsilon, \delta)$ via the
standard formula.

```{r}
zcdp_to_eps <- function(rho, delta = 1e-5) rho + 2 * sqrt(rho * log(1 / delta))

budget <- data.frame(sigma = sigma_grid[sigma_grid > 0])
budget$rho_total                  <- T_fixed * (1 / budget$sigma)^2 / 2
budget$epsilon_at_delta_1e_minus_5 <- zcdp_to_eps(budget$rho_total)
knitr::kable(budget, digits = 4,
             caption = "zCDP composition; sensitivity Delta = 1, delta = 1e-5")
```

Because $\rho$ and $T$ were chosen on the surrogate, this total
covers everything the procedure releases about the data. The one
assumption is the sensitivity $\Delta = 1$, a placeholder noted in
the limitations below.

The smallest $\varepsilon$, at $\sigma = 1$ where the fit is
already poor, is
`r signif(min(budget$epsilon_at_delta_1e_minus_5), 2)`. Whether
that is acceptable depends on the application.

## Summary

1. At $\sigma = 0$ the protocol matches the centralized fit.
   Adding DP takes one `rnorm()` call per site per iteration and
   nothing new in homomorpheR.
2. The vignette runs six values of $\sigma$, each for the fixed
   $T$, and tabulates the coefficients and their deviation from
   the centralized fit.
3. At noise levels where the fit is still close, $\varepsilon$ is
   large, because zCDP composition runs over $T$ Gaussian
   releases. Reducing it would need fewer iterations, a tighter
   sensitivity bound, or subsampling amplification, all outside
   this demonstration.
4. Choosing $\rho$ and $T$ counts against privacy too. It looks
   like setup, but it sets $T$, the multiplier on the budget.
   Choosing them without the data keeps the budget complete, at a
   cost in accuracy. Encryption hides the intermediate values, not
   these choices.

## Limitations
- **Tight sensitivity bounds** for the consensus update. $\Delta = 1$
  is a placeholder.
- **Paying for the selection instead of avoiding it.** Report-noisy-max
  or the exponential mechanism over the $\rho$ grid would let the
  cohort inform the choice for a declared cost. That needs a
  sensitivity bound on iterations-to-convergence under a one-record
  change, which we do not have.
- **Surrogate quality.** A surrogate that misjudges the cohort's
  curvature costs accuracy — a worse $\rho$, or a $T$ too short to
  converge — and there is no way to detect that from inside the
  protocol without spending budget.
- **Adaptive noise** schedules across iterations.
- **Subsampling amplification** of the DP accountant.
- **Decentralized averaging.** Replacing the aggregator with gossip
  averaging among the sites can amplify privacy
  [@CyffersEvenBelletMassoulie2022]. This protocol keeps a central
  aggregator.
- **Malicious-site protection.** A hostile site can corrupt either
  its $(x_i + u_i)$ or its noise draw; out of scope.

For consensus ADMM without noise, on real data, see
`vignette("cvxr-cox-lasso-dlbcl")`.

## References
