---
title: "Federated Consensus ADMM with CVXR"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Federated Consensus ADMM with CVXR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F, purl=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
## The encrypted chunks are gated `eval = RECOMPUTE` and are shown but
## not run when the vignette builds; their results are read from
## data(cvxr_admm_results), produced by data-raw/cvxr_admm_results.R
## from these same chunks. HOMOMORPHER_RECOMPUTE=true runs them for real.
if (!exists("RECOMPUTE"))
    RECOMPUTE <- isTRUE(as.logical(Sys.getenv("HOMOMORPHER_RECOMPUTE", "FALSE")))
data(cvxr_admm_results, package = "homomorpheR")
```

## Introduction
The `cox` and `cox-threshold` vignettes use optimizers such as
`mle()` and `optim()` that take an objective function. We give them
a function that runs one encrypted round across the sites, and the
optimizer uses it like any other objective.

That works for any objective whose log-likelihood is a sum over
sites. It does not work for general convex programs written in
[CVXR]{}. CVXR does not call a function we supply. It takes a
symbolic problem, converts it once to a standard form, and passes
that to a numerical solver.

For CVXR we use **federated consensus ADMM**. The global problem
is split into one local subproblem per site, each solved by CVXR in
the clear at that site. Only the cross-site **consensus update**
goes through the threshold-FHE channel: the per-site
$(x_i + u_i)$ vectors are summed and averaged encrypted, and the
result is recovered by threshold decryption. The local CVXR solves
stay fast because they use DPP (disciplined parametrized
programming): between iterations only the parameter values change,
not the problem structure.

This is the architectural complement to `cox-threshold.Rmd`:

| | `cox-threshold` | `cvxr-consensus-admm` |
|---|---|---|
| Optimizer | `stats4::mle()` (BFGS) | Consensus ADMM (one outer loop) |
| Local solver | `coxph(iter.max = 0)` | `CVXR::psolve()` |
| FHE call site | every BFGS function evaluation | every ADMM iteration |
| Topology | master/worker fan-out + fan-in | peer-to-peer with threshold aggregator |
| Threshold keys | yes (n-of-n) | yes (n-of-n) |

Both compose existing R machinery with a threshold-FHE channel
without rewriting the optimizer.

## The global problem and its consensus split

We pick L2-regularized logistic regression. With $N$ sites, local
data $(X_i, y_i)$ at site $i$, a shared coefficient
$x \in \mathbb{R}^p$, the global problem is

$$
\min_{x \in \mathbb{R}^p}
\sum_{i=1}^{N} \ell_i(x; X_i, y_i)
+ \frac{\lambda}{2}\, \lVert x \rVert_2^2
$$

where $\ell_i$ is the logistic loss on site $i$. The standard
consensus split (Boyd, Parikh, Chu, Peleato, Eckstein, 2011)
introduces
local copies $x_i \in \mathbb{R}^p$ and a single global consensus
$z \in \mathbb{R}^p$:

$$
\min_{\{x_i\}, z}
\sum_i \ell_i(x_i; X_i, y_i) + \frac{\lambda}{2}\, \lVert z \rVert_2^2
\quad \text{s.t.}\quad x_i = z, \ \forall i.
$$

The augmented-Lagrangian iteration is:

$$
\begin{aligned}
x_i^{k+1} &= \arg\min_{x_i}\, \ell_i(x_i) + \frac{\lambda}{2N}\lVert x_i\rVert_2^2 + \frac{\rho}{2}\lVert x_i - z^k + u_i^k\rVert_2^2, \\
z^{k+1} &= \frac{1}{N}\sum_i (x_i^{k+1} + u_i^k), \\
u_i^{k+1} &= u_i^k + (x_i^{k+1} - z^{k+1}).
\end{aligned}
$$

The $x$-update is local at each site; the $z$-update is the
consensus average that has to traverse the encrypted channel; the
$u$-update is local again. Only the $z$-update needs cryptography.

## The local CVXR subproblem

```{r}
library(homomorpheR)
suppressPackageStartupMessages(library(CVXR))

N   <- 3L
p   <- 4L
lam <- 1
```

We build the local problem so that DPP can engage: $z$ and $u$ are
`Parameter`s (so canonicalization is shared across iterations);
$X_i$, $y_i$, and $\rho$ are *constants* baked in at construction
time (so the augmented Lagrangian term enters affinely in the
parameters and the DPP fast path is not broken).

```{r}
build_local_problem <- function(X_i, y_i, rho_val) {
    n_i <- nrow(X_i)
    p_i <- ncol(X_i)

    x  <- Variable(p_i)
    zp <- Parameter(p_i)
    up <- Parameter(p_i)

    y_signs <- 2 * y_i - 1
    margins <- -y_signs * (X_i %*% x)

    local_loss <- sum(logistic(margins)) +
                  (lam / (2 * N)) * sum_squares(x)
    augmented  <- (rho_val / 2) * sum_squares(x - zp + up)

    prob <- Problem(Minimize(local_loss + augmented))

    value(zp) <- rep(0, p_i)
    value(up) <- rep(0, p_i)

    list(prob = prob, x = x, zp = zp, up = up)
}
```

A vignette-local site class wraps the local CVXR problem plus
the ADMM state. The exported `Site` from this package is shaped
for the master/worker pattern; ADMM is peer-to-peer and needs its
own per-site state, so we define `ConsensusSite` inline.

```{r}
library(S7)

## It inherits from homomorpheR's abstract `Site`, which supplies
## `name` and the `state` environment. That inheritance is what lets it
## take part in threshold key generation: `keygen_round()` and
## `partial_decrypt()` dispatch on `Site`, so each ConsensusSite
## generates and keeps its own share like any other party.
ConsensusSite <- new_class("ConsensusSite",
    parent     = homomorpheR::Site,
    properties = list(
        n = class_integer
    )
)

## A site is a *party*: it persists across the whole analysis, because
## it holds a secret share that threshold key generation put there. Its
## local CVXR problem is a property of the current run, not of the
## party, so it is set separately and can be rebuilt for a new rho
## without destroying the share.
make_consensus_site <- function(name, X_i, y_i) {
    st        <- new.env(parent = emptyenv())
    st$X      <- X_i
    st$y      <- y_i
    st$x_curr <- rep(0, ncol(X_i))
    st$u_curr <- rep(0, ncol(X_i))
    ConsensusSite(name = name, n = nrow(X_i), state = st)
}

set_rho <- function(site, rho_val) {
    st        <- site@state
    built     <- build_local_problem(st$X, st$y, rho_val)
    st$prob   <- built$prob
    st$x_var  <- built$x
    st$zp     <- built$zp
    st$up     <- built$up
    st$x_curr <- rep(0, ncol(st$X))
    st$u_curr <- rep(0, ncol(st$X))
    invisible(site)
}

local_update <- function(site, z_curr) {
    st <- site@state
    value(st$zp) <- z_curr
    value(st$up) <- st$u_curr
    psolve(st$prob, solver = "CLARABEL")
    if (!status(st$prob) %in% c("optimal", "optimal_inaccurate"))
        stop("Local solve at ", site@name, " did not reach optimal status.")
    st$x_curr <- as.numeric(value(st$x_var))
    invisible(st$x_curr)
}
```

## The threshold-FHE consensus aggregation

The consensus update needs to compute
$z^{k+1} = \frac{1}{N}\sum_i (x_i^{k+1} + u_i^k)$. This is a
length-$p$ vector average: pack each site's $x_i + u_i$ into the
slots of one value, encrypt under the joint public key, sum the
encrypted vectors, multiply by the unencrypted constant $1/N$
(one multiplication, which uses one level of multiplicative
depth), and threshold-decrypt the result.

We use the same threshold infrastructure as `cox-threshold.Rmd`:
a CKKS context with `Feature$MULTIPARTY` enabled and
`make_threshold_master()`, which walks the chained
`multiparty_key_gen()` setup across the sites.

Each site encrypts its own $(x_i + u_i)$ under the joint public
key and hands back an encrypted value. The aggregator adds them,
scales, and calls `decrypt()` to threshold-decrypt the total. If
the aggregator did the encrypting instead, it would see each site's
vector in the clear, which is what the protocol is meant to
prevent. Decryption works the same way: the aggregator sends the
encrypted total to each site, and each site applies its own share.
The shares never leave the sites.

```{r eval=RECOMPUTE}
cc <- openfhe.R::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L,
                           features             = c(openfhe.R::Feature$MULTIPARTY))
```

Threshold key generation happens once, up front, and it runs *through*
the sites: each generates its own share, keeps it, and passes on only
a public key. So the sites have to exist before the aggregator does —
we set them up below, once the data is in hand. The keys do not depend
on $\rho$, so the same sites and the same master serve both the tuning
sweep and the final run; only each site's local problem is rebuilt when
$\rho$ changes.

```{r}
## Site-side. The site forms x_i + u_i and encrypts it with the public
## bundle it was handed at setup -- context and joint public key, no
## secrets. The cleartext vector never leaves this function.
site_contribution <- function(site) {
    st <- site@state
    ## Encrypts with the setup bundle; errors if the site was never
    ## configured.
    encrypt(site, st$x_curr + st$u_curr)
}

## Aggregator-side. It receives encrypted values, adds them, scales by 1/N,
## and threshold-decrypts. At no point does it hold an individual
## (x_i + u_i).
encrypted_consensus <- function(threshold_master, sites) {
    ct_avg <- Reduce(`+`, lapply(sites, site_contribution)) * (1 / length(sites))
    decrypt(threshold_master, ct_avg, len = p)
}
```

What the protocol reveals: at every ADMM iteration the aggregator
learns the new consensus $z^k$. The per-site $(x_i + u_i)$ vectors
are never in the clear outside their sites. The additions and the
scalar multiply are done encrypted, and decryption needs all $N$
shares. So the aggregator's records contain the sequence
$\{z^k\}$ but no site's contribution. One site's records contain
its own $X_i, y_i$ and its own share, but no other site's
contribution, and one share alone cannot decrypt anything.

## Simulated cohort

Three sites with deliberately uneven sample sizes; same true
coefficients across sites.

```{r}
set.seed(98765)

beta_true <- c(0.5, -1.0, 0.3, 0.8)
n_per     <- c(400, 250, 350)

site_data <- lapply(seq_len(N), function(i) {
    X <- matrix(rnorm(n_per[i] * p), nrow = n_per[i], ncol = p)
    eta <- X %*% beta_true
    y <- rbinom(n_per[i], size = 1, prob = 1 / (1 + exp(-eta)))
    list(X = X, y = y)
})

X_all <- do.call(rbind, lapply(site_data, `[[`, "X"))
y_all <- unlist(lapply(site_data, `[[`, "y"))
```

### The parties, and the key-generation chain

Now the sites exist, so the chain can run through them. Each generates
its own share and keeps it; what comes back is an aggregator holding
the joint public key and nothing secret.

```{r eval=RECOMPUTE}
sites <- lapply(seq_len(N), function(i)
    make_consensus_site(paste0("Site ", i),
                        site_data[[i]]$X, site_data[[i]]$y))

master <- make_threshold_master("Aggregator",
                                crypto_context = cc,
                                sites          = sites)

share_check <- c(aggregator_has_share = !is.null(master@state$sk),
                 every_site_has_one   = all(vapply(sites, function(s) !is.null(s@state$sk), TRUE)))
share_check
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
cvxr_admm_results$share_check
```

## Centralized CVXR fit

For the comparison at the end of the run, we fit the same problem
centrally. This is the target the federated fit should match.

```{r}
x_central <- Variable(p)
y_signs  <- 2 * y_all - 1
margins  <- -y_signs * (X_all %*% x_central)
central_prob <- Problem(Minimize(
    sum(logistic(margins)) + (lam / 2) * sum_squares(x_central)
))
psolve(central_prob, solver = "CLARABEL")
beta_central <- as.numeric(value(x_central))
```

## Tuning $\rho$ before the loop

ADMM convergence depends on $\rho$. We sweep three candidate
values and keep the one that converges fastest, fixing it for the
run that follows.

The sweep uses the same encrypted channel as the main loop.

Sweeping in the clear first would not be safe. To average
$(x_i + u_i)$ in the clear, each site would have to send that
vector to the aggregator, once per iteration and once per
candidate. $x_i^{k+1}$ minimizes site $i$'s local loss, so it is a
direct function of that site's $(X_i, y_i)$.

Running the sweep encrypted needs nothing beyond what one ADMM run
needs. The $x$-update is a CVXR solve on the site's own data, and
the $u$-update is arithmetic at the site. Only the consensus
average is encrypted, with one addition and one multiplication by
a constant per iteration, so the sweep fits in
`multiplicative_depth = 1L`. Each iteration encrypts fresh values,
so neither depth nor noise builds up across iterations or across
candidates. The stopping test and the `which.min` over the grid are
comparisons, which CKKS does not support, but they use only $z$,
which the protocol decrypts every iteration anyway, and each
site's own $x_i$ and $u_i$. $\rho$ is a public constant that all
sites must agree on.

The extra cost is small: the encrypted channel carries a
length-$p$ vector, while each iteration also runs $N$
interior-point solves.

```{r eval=RECOMPUTE}
rho_grid    <- c(10, 50, 200)
sweep_iters <- integer(length(rho_grid))

for (g in seq_along(rho_grid)) {
    rho_val <- rho_grid[g]
    ## The parties persist -- they hold the key shares. Only the local
    ## problem is rebuilt for this rho.
    for (s in sites) set_rho(s, rho_val)
    z_curr <- rep(0, p)
    converged <- FALSE
    for (k in seq_len(40)) {
        for (s in sites) local_update(s, z_curr)
        z_new <- encrypted_consensus(master, sites)
        for (s in sites) {
            s@state$u_curr <- s@state$u_curr + (s@state$x_curr - z_new)
        }
        primal_res <- sqrt(mean(unlist(lapply(sites, function(s)
            sum((s@state$x_curr - z_new)^2)))))
        dual_res <- sqrt(N) * rho_val * sqrt(sum((z_new - z_curr)^2))
        z_curr <- z_new
        if (primal_res < 1e-3 && dual_res < 1e-3) {
            converged <- TRUE
            sweep_iters[g] <- k
            break
        }
    }
    if (!converged) sweep_iters[g] <- NA_integer_
}

sweep_table <- data.frame(rho = rho_grid, iters = sweep_iters)
knitr::kable(sweep_table,
             caption = "Threshold-FHE ADMM iterations to convergence")

rho_chosen <- rho_grid[which.min(sweep_iters)]
cat(sprintf("Selected rho = %g (converged in %d iterations).\n",
            rho_chosen, min(sweep_iters, na.rm = TRUE)))
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
knitr::kable(cvxr_admm_results$sweep_table,
             caption = "Threshold-FHE ADMM iterations to convergence")
cat(sprintf("Selected rho = %g (converged in %d iterations).\n",
            cvxr_admm_results$rho_chosen,
            min(cvxr_admm_results$sweep_table$iters, na.rm = TRUE)))
```

## The ADMM run at the chosen $\rho$

Local problems rebuilt at the chosen $\rho$ — the same parties, holding
the same shares from the key-generation chain above — then the main
loop, identical to the sweep body except that it records the
trajectory:

```{r eval=RECOMPUTE}
for (s in sites) set_rho(s, rho_chosen)

z_curr    <- rep(0, p)
max_iter  <- 40
reltol    <- 1e-3
converged <- FALSE
trajectory <- vector("list", max_iter)

for (k in seq_len(max_iter)) {
    for (s in sites) local_update(s, z_curr)
    z_new <- encrypted_consensus(master, sites)
    for (s in sites) {
        s@state$u_curr <- s@state$u_curr + (s@state$x_curr - z_new)
    }
    primal_res <- sqrt(mean(unlist(lapply(sites, function(s)
        sum((s@state$x_curr - z_new)^2)))))
    dual_res <- sqrt(N) * rho_chosen * sqrt(sum((z_new - z_curr)^2))
    trajectory[[k]] <- z_new
    z_curr <- z_new
    if (primal_res < reltol && dual_res < reltol) {
        converged <- TRUE
        trajectory <- trajectory[seq_len(k)]
        break
    }
}

if (!converged)
    stop("Encrypted ADMM did not converge within max_iter; rerun the rho sweep.")

cat(sprintf("Encrypted ADMM converged in %d iterations.\n", length(trajectory)))
```

```{r eval=RECOMPUTE, echo=FALSE}
cvxr_admm_results <- list(share_check = share_check,
                          sweep_table = sweep_table,
                          rho_chosen  = rho_chosen,
                          trajectory  = trajectory,
                          z_final     = z_curr,
                          reltol      = reltol)
```

```{r echo=FALSE, purl=FALSE, eval=!RECOMPUTE}
cat(sprintf("Encrypted ADMM converged in %d iterations.\n",
            length(cvxr_admm_results$trajectory)))
z_curr <- cvxr_admm_results$z_final
reltol <- cvxr_admm_results$reltol
```

## Comparison with the centralized fit

```{r}
comparison <- data.frame(
    coefficient            = paste0("beta_", seq_len(p)),
    threshold_distributed  = z_curr,
    centralized_cvxr       = beta_central,
    abs_diff               = abs(z_curr - beta_central)
)
knitr::kable(comparison, digits = 6,
             caption = "Threshold-FHE consensus ADMM vs. centralized CVXR")

max_diff <- max(abs(z_curr - beta_central))
cat(sprintf("Max absolute coefficient difference vs. centralized fit: %.2e\n",
            max_diff))

if (max_diff > 10 * reltol)
    stop("Encrypted ADMM agreement with the aggregated cleartext fit is too loose; ",
         "investigate before publishing this run.")
```

## Discussion
1. **CVXR problems work unchanged.** Each site solves its local
   CVXR problem in the clear. Only the consensus update is
   encrypted. The `Problem(Minimize(...))` is the same one you
   would write for pooled cleartext data.
2. **No single party holds the secret key.** Each site generated
   its own share during key generation and kept it. The aggregator
   holds only the joint public key. No single party can decrypt an
   intermediate value, and the consensus is decrypted only when
   every site contributes a partial decryption.
3. **DPP keeps the inner loop fast.** Each site's CVXR problem is
   built once at setup, and ADMM iterations only update the
   parameter values. Without DPP, CVXR would convert the problem
   to standard form again at every iteration.

## Limitations
- **Honest-but-curious trust.** A site that misreports its local
  $x_i + u_i$ can corrupt the consensus. Detecting this requires
  commitments / zero-knowledge proofs that this vignette does
  not implement.
- **The aggregator sees the trajectory $\{z^k\}$.** The consensus
  is decrypted at every iteration so the loop can test for
  convergence. The aggregator therefore learns one trajectory per
  candidate $\rho$, since the sweep runs the same protocol. Hiding
  the trajectory would mean keeping the residuals encrypted. The
  stopping test and the `which.min` would then be comparisons on
  encrypted values: a fixed number of iterations in place of
  `break`, and an approximate encrypted comparison in place of `<`.
  That is possible, but it is a different and much more expensive
  protocol.
- **No output privacy.** The released $\hat\beta = z^\star$ is the
  same coefficient vector as the centralized fit. Output-level
  attacks are out of scope here; the companion
  `vignette("cvxr-consensus-admm-dp")` demonstrates output DP
  composition.
