---
title: "Federated Consensus ADMM with CVXR over Threshold CKKS"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Federated Consensus ADMM under Threshold CKKS}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## Introduction
The `cox` and `cox-threshold` vignettes show that *function-call*
optimizers — `mle()`, `optim()`, `coxph`'s Newton-Raphson via
callback — compose with FHE: hand them a callback that performs an
encrypted master/worker round, and the optimizer cannot tell.

That pattern fits any objective with a sum-decomposable
log-likelihood, but it does not generalize to arbitrary convex
programs. The richer R toolkit for convex optimization —
[CVXR]{} — operates on a *symbolic* problem, not a callback. We
cannot hand CVXR a function that secretly does FHE on every call,
because CVXR does not call our function: it canonicalizes the
problem once and ships the canonical form to a numerical solver.

The replacement is **federated consensus ADMM**. We split the
global problem into per-site local subproblems each handled by
CVXR in cleartext, and use the threshold-FHE channel only to
compute the cross-site **consensus update**. The local CVXR solves
exploit DPP (disciplined parametric programming) and stay fast
because their Parameter values change but the problem structure
does not. The encrypted channel earns its keep on the consensus
aggregation, where the per-site $(x_i + u_i)$ vectors are summed
and averaged under encryption and the result is recovered via
threshold decryption.

This is the architectural complement to `cox-threshold.Rmd`:

| | `cox-threshold` | `cvxr-consensus-admm` |
|---|---|---|
| Optimizer | `stats4::mle()` (BFGS) | Consensus ADMM (one outer loop) |
| Local solver | `coxph(iter.max = 0)` | `CVXR::psolve()` |
| FHE call site | every BFGS function evaluation | every ADMM iteration |
| Topology | master/worker fan-out + fan-in | peer-to-peer with threshold aggregator |
| Threshold keys | yes (n-of-n) | yes (n-of-n) |

Both compose existing R machinery with a threshold-FHE channel
without rewriting the optimizer.

## The global problem and its consensus split

We pick L2-regularised logistic regression. With $N$ sites, local
data $(X_i, y_i)$ at site $i$, a shared coefficient
$x \in \mathbb{R}^p$, the global problem is

$$
\min_{x \in \mathbb{R}^p}
\sum_{i=1}^{N} \ell_i(x; X_i, y_i)
+ \frac{\lambda}{2}\, \lVert x \rVert_2^2
$$

where $\ell_i$ is the logistic loss on site $i$. The standard
consensus split (Boyd, Parikh, Chu, Peleato, Eckstein, 2011)
introduces
local copies $x_i \in \mathbb{R}^p$ and a single global consensus
$z \in \mathbb{R}^p$:

$$
\min_{\{x_i\}, z}
\sum_i \ell_i(x_i; X_i, y_i) + \frac{\lambda}{2}\, \lVert z \rVert_2^2
\quad \text{s.t.}\quad x_i = z, \ \forall i.
$$

The augmented-Lagrangian iteration is:

$$
\begin{aligned}
x_i^{k+1} &= \arg\min_{x_i}\, \ell_i(x_i) + \frac{\lambda}{2N}\lVert x_i\rVert_2^2 + \frac{\rho}{2}\lVert x_i - z^k + u_i^k\rVert_2^2, \\
z^{k+1} &= \frac{1}{N}\sum_i (x_i^{k+1} + u_i^k), \\
u_i^{k+1} &= u_i^k + (x_i^{k+1} - z^{k+1}).
\end{aligned}
$$

The $x$-update is local at each site; the $z$-update is the
consensus average that has to traverse the encrypted channel; the
$u$-update is local again. Only the $z$-update needs cryptography.

## The local CVXR subproblem

```{r}
library(homomorpheR)
suppressPackageStartupMessages(library(CVXR))

N   <- 3L
p   <- 4L
lam <- 1
```

We build the local problem so that DPP can engage: $z$ and $u$ are
`Parameter`s (so canonicalization is shared across iterations);
$X_i$, $y_i$, and $\rho$ are *constants* baked in at construction
time (so the augmented Lagrangian term enters affinely in the
parameters and the DPP fast path is not broken).

```{r}
build_local_problem <- function(X_i, y_i, rho_val) {
    n_i <- nrow(X_i)
    p_i <- ncol(X_i)

    x  <- Variable(p_i)
    zp <- Parameter(p_i)
    up <- Parameter(p_i)

    y_signs <- 2 * y_i - 1
    margins <- -y_signs * (X_i %*% x)

    local_loss <- sum(logistic(margins)) +
                  (lam / (2 * N)) * sum_squares(x)
    augmented  <- (rho_val / 2) * sum_squares(x - zp + up)

    prob <- Problem(Minimize(local_loss + augmented))

    value(zp) <- rep(0, p_i)
    value(up) <- rep(0, p_i)

    list(prob = prob, x = x, zp = zp, up = up)
}
```

A vignette-local site class wraps the local CVXR problem plus
the ADMM state. The exported `Site` from this package is shaped
for the master/worker pattern; ADMM is peer-to-peer and needs its
own per-site state, so we define `ConsensusSite` inline.

```{r}
library(S7)

ConsensusSite <- new_class("ConsensusSite",
    properties = list(
        name  = class_character,
        n     = class_integer,
        state = class_any
    )
)

make_consensus_site <- function(name, X_i, y_i, rho_val) {
    st        <- new.env(parent = emptyenv())
    st$X      <- X_i
    st$y      <- y_i
    built     <- build_local_problem(X_i, y_i, rho_val)
    st$prob   <- built$prob
    st$x_var  <- built$x
    st$zp     <- built$zp
    st$up     <- built$up
    st$x_curr <- rep(0, ncol(X_i))
    st$u_curr <- rep(0, ncol(X_i))
    ConsensusSite(name = name, n = nrow(X_i), state = st)
}

local_update <- function(site, z_curr) {
    st <- site@state
    value(st$zp) <- z_curr
    value(st$up) <- st$u_curr
    psolve(st$prob, solver = "CLARABEL")
    if (!status(st$prob) %in% c("optimal", "optimal_inaccurate"))
        stop("Local solve at ", site@name, " did not reach optimal status.")
    st$x_curr <- as.numeric(value(st$x_var))
    invisible(st$x_curr)
}
```

## The threshold-FHE consensus aggregation

The consensus update needs to compute
$z^{k+1} = \frac{1}{N}\sum_i (x_i^{k+1} + u_i^k)$. This is a
length-$p$ vector average: pack each site's $x_i + u_i$ into a
length-$p$ packed plaintext, encrypt under the joint public key,
sum the ciphertexts, multiply by the plaintext constant $1/N$
(one ciphertext-plaintext multiplication, depth 1), and
threshold-decrypt the result.

We use the same threshold infrastructure as `cox-threshold.Rmd`:
a CKKS context with `Feature$MULTIPARTY` enabled and
`make_threshold_master()`, which runs the chained
`multiparty_key_gen()` setup automatically. The consensus
function calls `master_encrypt()` to encrypt each site's
contribution under the joint public key and `master_decrypt()` to
threshold-decrypt the result; the partial-decrypt fan-in across
sites is handled inside the master class.

```{r}
cc <- openfhe.R::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L,
                           features             = c(openfhe.R::Feature$MULTIPARTY))
```

```{r}
encrypted_consensus <- function(threshold_master, sites) {
    cts <- vector("list", length(sites))
    for (i in seq_along(sites)) {
        st  <- sites[[i]]@state
        val <- st$x_curr + st$u_curr
        cts[[i]] <- master_encrypt(threshold_master, val)
    }
    ct_sum <- Reduce(`+`, cts)
    ct_avg <- ct_sum * (1 / length(sites))
    master_decrypt(threshold_master, ct_avg, len = p)
}
```

What this protocol hides and reveals: every ADMM iteration the
aggregator recovers the *new consensus* $z^k$ exactly. The
per-site $(x_i + u_i)$ vectors never appear in plaintext anywhere
— additions and the scalar multiply happen under encryption, and
the decryption is n-of-n. A subpoena to the aggregator yields the
trajectory $\{z^k\}$ but no individual site's contribution. A
subpoena to any one site yields that site's local $X_i, y_i$ and
its own secret share, but no other site's contribution and no
decryption power.

## Simulated cohort

Three sites with deliberately uneven sample sizes; same true
coefficients across sites.

```{r}
set.seed(98765)

beta_true <- c(0.5, -1.0, 0.3, 0.8)
n_per     <- c(400, 250, 350)

site_data <- lapply(seq_len(N), function(i) {
    X <- matrix(rnorm(n_per[i] * p), nrow = n_per[i], ncol = p)
    eta <- X %*% beta_true
    y <- rbinom(n_per[i], size = 1, prob = 1 / (1 + exp(-eta)))
    list(X = X, y = y)
})

X_all <- do.call(rbind, lapply(site_data, `[[`, "X"))
y_all <- unlist(lapply(site_data, `[[`, "y"))
```

## Centralized CVXR fit

For the comparison at the end of the run, we fit the same problem
centrally. This is the target the federated fit should match.

```{r}
x_central <- Variable(p)
y_signs  <- 2 * y_all - 1
margins  <- -y_signs * (X_all %*% x_central)
central_prob <- Problem(Minimize(
    sum(logistic(margins)) + (lam / 2) * sum_squares(x_central)
))
psolve(central_prob, solver = "CLARABEL")
beta_central <- as.numeric(value(x_central))
```

## Tuning $\rho$ before the loop

ADMM convergence depends on $\rho$. We sweep three candidate
values against a *cleartext* copy of the protocol (no FHE) and
pick the one that converges fastest. The chosen value is fixed
for the encrypted run.

```{r}
rho_grid    <- c(10, 50, 200)
sweep_iters <- integer(length(rho_grid))

for (g in seq_along(rho_grid)) {
    rho_val <- rho_grid[g]
    sites <- lapply(seq_len(N), function(i)
        make_consensus_site(paste0("Site ", i),
                            site_data[[i]]$X, site_data[[i]]$y,
                            rho_val = rho_val))
    z_curr <- rep(0, p)
    converged <- FALSE
    for (k in seq_len(40)) {
        for (s in sites) local_update(s, z_curr)
        means_xu <- Reduce(`+`,
            lapply(sites, function(s) s@state$x_curr + s@state$u_curr)) / N
        z_new <- means_xu
        for (s in sites) {
            s@state$u_curr <- s@state$u_curr + (s@state$x_curr - z_new)
        }
        primal_res <- sqrt(mean(unlist(lapply(sites, function(s)
            sum((s@state$x_curr - z_new)^2)))))
        dual_res <- sqrt(N) * rho_val * sqrt(sum((z_new - z_curr)^2))
        z_curr <- z_new
        if (primal_res < 1e-3 && dual_res < 1e-3) {
            converged <- TRUE
            sweep_iters[g] <- k
            break
        }
    }
    if (!converged) sweep_iters[g] <- NA_integer_
}

sweep_table <- data.frame(rho = rho_grid, iters = sweep_iters)
knitr::kable(sweep_table, caption = "Cleartext ADMM iterations to convergence")

rho_chosen <- rho_grid[which.min(sweep_iters)]
cat(sprintf("Selected rho = %g (converged in %d iterations).\n",
            rho_chosen, min(sweep_iters, na.rm = TRUE)))
```

## The encrypted ADMM loop

Threshold key generation, sites built with the chosen $\rho$, then
the main loop:

```{r}
master <- make_threshold_master("Aggregator",
                                crypto_context = cc,
                                n_sites        = N)

sites <- lapply(seq_len(N), function(i)
    make_consensus_site(paste0("Site ", i),
                        site_data[[i]]$X, site_data[[i]]$y,
                        rho_val = rho_chosen))

z_curr    <- rep(0, p)
max_iter  <- 40
reltol    <- 1e-3
converged <- FALSE
trajectory <- vector("list", max_iter)

for (k in seq_len(max_iter)) {
    for (s in sites) local_update(s, z_curr)
    z_new <- encrypted_consensus(master, sites)
    for (s in sites) {
        s@state$u_curr <- s@state$u_curr + (s@state$x_curr - z_new)
    }
    primal_res <- sqrt(mean(unlist(lapply(sites, function(s)
        sum((s@state$x_curr - z_new)^2)))))
    dual_res <- sqrt(N) * rho_chosen * sqrt(sum((z_new - z_curr)^2))
    trajectory[[k]] <- z_new
    z_curr <- z_new
    if (primal_res < reltol && dual_res < reltol) {
        converged <- TRUE
        trajectory <- trajectory[seq_len(k)]
        break
    }
}

if (!converged)
    stop("Encrypted ADMM did not converge within max_iter; rerun the rho sweep.")

cat(sprintf("Encrypted ADMM converged in %d iterations.\n", length(trajectory)))
```

## Comparison with the centralized fit

```{r}
comparison <- data.frame(
    coefficient            = paste0("beta_", seq_len(p)),
    threshold_distributed  = z_curr,
    centralized_cvxr       = beta_central,
    abs_diff               = abs(z_curr - beta_central)
)
knitr::kable(comparison, digits = 6,
             caption = "Threshold-FHE consensus ADMM vs. centralized CVXR")

max_diff <- max(abs(z_curr - beta_central))
cat(sprintf("Max absolute coefficient difference vs. centralized fit: %.2e\n",
            max_diff))

if (max_diff > 10 * reltol)
    stop("Encrypted ADMM agreement with the aggregated cleartext fit is too loose; ",
         "investigate before publishing this run.")
```

## Discussion
1. **CVXR symbolic problems compose with threshold FHE.** The
   local CVXR solve runs cleartext at each site; only the
   cross-site consensus update goes through the encrypted
   channel. The reader does not have to rewrite their CVXR model
   for an encrypted setting — the same `Problem(Minimize(...))`
   they would write for cleartext data is used here verbatim.
2. **No single party holds the secret key.** `make_threshold_master()`
   distributes the secret across all sites; the aggregator holds
   only the joint public key. Encrypted intermediates are
   undecryptable by any single party, and the consensus result
   appears only after the n-of-n partial-decryption fusion.
3. **DPP keeps the inner loop fast.** Each site's CVXR problem is
   built once at setup; ADMM iterations only update the
   parameter values. Without DPP, the canonicalization would
   re-run every iteration and the vignette would be infeasible.

## Limitations
- **Honest-but-curious trust.** A site that misreports its local
  $x_i + u_i$ can corrupt the consensus. Detecting this requires
  commitments / zero-knowledge proofs that this vignette does
  not implement.
- **The aggregator sees the trajectory $\{z^k\}$.** Per-iteration
  consensus values are revealed in plaintext (after fusion) so
  the optimizer can decide convergence. A subpoena to the
  aggregator yields this trajectory.
- **No output privacy.** The released $\hat\beta = z^\star$ is the
  same coefficient vector as the centralized fit. Output-level
  attacks are out of scope here; the companion DP vignettes (not
  yet ported) demonstrate output DP composition.
