---
title: "Federated Cox-Lasso via Consensus ADMM on DLBCL"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Federated Cox-Lasso via Consensus ADMM on DLBCL}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r ktab, echo=FALSE}
## Tables through kableExtra. Math written as $...$ in headers, cells
## and captions becomes \( ... \) and `code` becomes <code>, because
## pandoc does not process math inside a raw HTML table.
ktab <- function(x, ..., col.names = names(x), caption = NULL) {
    tex <- function(s)
        gsub("`([^`]*)`", "<code>\\1</code>",
             gsub("\\$([^$]+)\\$", "\\\\(\\1\\\\)", s))
    chr <- vapply(x, is.character, logical(1))
    x[chr] <- lapply(x[chr], tex)
    tab <- knitr::kable(x, format = "html", escape = FALSE,
                        col.names = tex(col.names),
                        caption = if (!is.null(caption)) tex(caption), ...)
    kableExtra::kable_styling(tab, bootstrap_options = c("striped", "condensed"),
                              full_width = TRUE)
}
```

```{r knitr-opts, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error   = FALSE,
  tidy    = FALSE,
  cache   = FALSE
)
## Inline numbers in scientific notation as HTML (knitr's default
## prints a bare "8.3^{-4}" in HTML output).
sci <- function(x, digits = 2) {
  e <- floor(log10(abs(x)))
  if (x == 0 || e >= -2) return(formatC(x, format = "fg", digits = digits))
  sprintf("%s&nbsp;&times;&nbsp;10<sup>%d</sup>",
          formatC(x / 10^e, format = "f", digits = digits - 1), e)
}
```

```{r setup, purl=FALSE, include=FALSE}
## This vignette is the single source of truth for the encrypted
## Cox-lasso consensus-ADMM example. The code chunks below contain the
## real pipeline, gated `eval = RECOMPUTE`: RECOMPUTE defaults to FALSE
## here, so the chunks are shown but NOT run when the vignette builds
## (the iterated encrypted ADMM is far too slow for that). The
## displayed numbers are read from data(cvxr_consensus), which was
## produced by knitr::purl()-ing exactly these chunks and running them
## (see data-raw/cvxr_consensus.R and the verification note below).
## That script writes the purled chunks to inst/scripts/cvxr-consensus.R,
## which the openfhe-jss manuscript reads with knitr::read_chunk(), so
## vignette, script, manuscript, and data object cannot drift.
## Default stays FALSE, but an environment variable can turn the real
## pipeline on for a whole build -- `HOMOMORPHER_RECOMPUTE=true R CMD
## build ...` or the same in front of a pkgdown run. Without a hook at
## this level there is no way to make a full build actually execute the
## encrypted pipeline, and a build that replays the shipped artifact
## cannot tell you the code still works.
if (!exists("RECOMPUTE"))
    RECOMPUTE <- isTRUE(as.logical(Sys.getenv("HOMOMORPHER_RECOMPUTE", "FALSE")))

suppressPackageStartupMessages(library(homomorpheR))
data(cvxr_consensus, package = "homomorpheR")
data(DLBCL,          package = "homomorpheR")
data(DLBCL_gex,      package = "homomorpheR")

K          <- cvxr_consensus$params$K
LAMBDA     <- cvxr_consensus$params$LAMBDA
RHO        <- cvxr_consensus$params$RHO
MAX_ITER   <- cvxr_consensus$params$MAX_ITER
TOL        <- cvxr_consensus$params$TOL
P_raw      <- ncol(DLBCL_gex)
N_total    <- nrow(DLBCL)
top_idx    <- cvxr_consensus$top_idx
sigma_K    <- cvxr_consensus$sigma_K
agg_beta   <- cvxr_consensus$agg_beta
z_ref      <- cvxr_consensus$z_ref
z_enc      <- cvxr_consensus$z_enc
trajectory <- cvxr_consensus$trajectory
n_iter_ref <- cvxr_consensus$n_iter_ref
n_iter_enc <- cvxr_consensus$n_iter_enc
pool_agree <- cvxr_consensus$pool_agree
```

## Introduction

This vignette runs the threshold-FHE consensus-ADMM protocol on a
real, high-dimensional survival problem:
a stratified **Cox lasso** on the diffuse large-B-cell lymphoma (DLBCL)
gene-expression cohort of @Rosenwald2002, the dataset @BayleFanLou2025
use to motivate distributed Cox estimation.

The `cox` and `cox-threshold` vignettes give an optimizer such as
`mle()` an objective function that runs one encrypted round across
the sites. That does not work for [CVXR][], which does not call a
function we supply: it converts a symbolic problem once to a
standard form and passes that to a solver. Instead we use
**federated consensus ADMM** (Boyd, Parikh, Chu, Peleato and
Eckstein, 2011). The global problem is split into per-site
subproblems, each solved by [CVXR][] in the clear, and only the
cross-site **consensus average** is encrypted. Disciplined
parametrized programming (DPP) keeps the per-site solves fast:
between iterations only the `Parameter` values change, not the
problem structure.

We develop the fit in two passes. First we run the whole thing **in the
clear** to fix the target: standardize, screen, and solve the consensus
ADMM with an ordinary unencrypted average, checking against the
centralized [CVXR][] solve. Then we put it **under threshold FHE**,
replacing each cross-site sum — the standardization moments, the
screening statistics, and the per-iteration consensus average — with
one round of the encrypted-summation primitive, and confirm the
encrypted fit reproduces the in-the-clear reference.

> **Reproducibility and verification.** The code chunks in this
> vignette *are* the pipeline; they are gated `eval = RECOMPUTE` and do
> not run when the vignette builds (every encrypted ADMM iteration
> repeats the per-site [CVXR][] solves). The displayed numbers come
> from `data(cvxr_consensus)`, which was produced by extracting these
> chunks with `knitr::purl()` and running them. To verify the
> results, do the same:
>
> ```r
> vig <- system.file("doc", "cvxr-cox-lasso-dlbcl.Rmd",
>                    package = "homomorpheR")
> RECOMPUTE <- TRUE                                  # un-gate the chunks
> src <- knitr::purl(vig, output = tempfile(fileext = ".R"), quiet = TRUE)
> source(src)                                        # runs the pipeline (minutes)
> fresh <- cvxr_consensus                            # just recomputed
> data(cvxr_consensus, package = "homomorpheR")      # the shipped copy
> max(abs(fresh$z_enc - cvxr_consensus$z_enc))       # small, not zero
> ```
>
> Threshold key generation is randomized, so a re-run reproduces the
> shipped coefficients up to CKKS approximation noise, not
> bit-for-bit. This `purl()`-then-`source()` is exactly what
> `data-raw/cvxr_consensus.R` does to build the shipped object.
> Building the vignette with the environment variable
> `HOMOMORPHER_RECOMPUTE=true` runs the chunks in place, and every
> number displayed then comes from that run.

## The global problem and its consensus split

The three molecular subgroups (GCB, ABC, Type III) are our sites: they
arise from different cells of origin and have systematically
different prognosis. Using them as sites is a choice made for this
demonstration.
With $N$ sites, per-site standardized design matrices
$X_k \in \mathbb{R}^{n_k \times p}$, event times $t_k$, status
$\delta_k$, a shared coefficient $\beta \in \mathbb{R}^p$, and the
per-stratum Cox partial log-likelihood $\ell_k$ in Breslow form, the
global problem is

$$
\min_{\beta \in \mathbb{R}^p}\;
-\sum_{k=1}^{N}\ell_k(\beta)\;+\;\lambda\lVert\beta\rVert_1 .
$$

Because the partial likelihood factorizes additively across strata, the
consensus split introduces per-site copies $x_k$ and a global consensus
$z$:

$$
\min_{\{x_k\},\,z}\;
\sum_k\bigl(-\ell_k(x_k)\bigr) + \lambda\lVert z\rVert_1
\quad\text{s.t.}\quad x_k = z,\ \forall k,
$$

with augmented-Lagrangian iteration

$$
\begin{aligned}
x_k^{t+1} &= \arg\min_x -\ell_k(x) + \tfrac{\rho}{2}\lVert x - z^t + u_k^t\rVert_2^2,\\
z^{t+1}   &= S_{\lambda/(N\rho)}\!\Bigl(\tfrac{1}{N}\sum_k (x_k^{t+1}+u_k^t)\Bigr),\\
u_k^{t+1} &= u_k^t + (x_k^{t+1}-z^{t+1}),
\end{aligned}
$$

where $S_\tau(v)=\operatorname{sign}(v)\max(|v|-\tau,0)$ is
soft-thresholding (the proximal map of $\tau\lVert\cdot\rVert_1$). The
$x$-update is the per-site [CVXR][] solve; the $z$-update needs the
cross-site average $\bar w=\tfrac1N\sum_k(x_k+u_k)$, which traverses
the encrypted channel at every iteration; the soft-threshold is
closed-form at the aggregator and adds no cryptographic depth.

The pipeline needs `survival` and [CVXR][] alongside the encryption
packages.

```{r cvxr-libs, eval=RECOMPUTE}
library(survival)
library(CVXR)
library(openfhe.R)
library(homomorpheR)
```

## The federated fit in the clear

The three subgroups are our sites. We load the cohort, split it by
subgroup, and keep each site's raw expression matrix, event times, and
status.

```{r cvxr-data, eval=RECOMPUTE}
data(DLBCL,     package = "homomorpheR")  # 235 patients: survival + signatures
data(DLBCL_gex, package = "homomorpheR")  # 235 patients x 6416 Lymphochip probes
dlbcl <- DLBCL
dlbcl$Subgroup <- factor(dlbcl$Subgroup, levels = c("GCB","ABC","Type III"))
stopifnot(identical(as.character(dlbcl$ID), rownames(DLBCL_gex)))  # rows aligned

sites_raw <- lapply(levels(dlbcl$Subgroup), function(s) {
    idx <- which(dlbcl$Subgroup == s)
    list(name = s, X = DLBCL_gex[idx, ], time = dlbcl$time[idx],
         status = dlbcl$status[idx])
})
names(sites_raw) <- levels(dlbcl$Subgroup)
N_sites <- length(sites_raw)
N_total <- sum(vapply(sites_raw, function(s) nrow(s$X), 1L))
P_raw   <- ncol(DLBCL_gex)
```

We standardize the features so the L1 penalty applies uniformly across
predictors on different scales. The pooled mean and variance are
additive over sites: site $k$ contributes $S_k=\sum_{i\in k}x_i$ and
$Q_k=\sum_{i\in k}x_i^2$, from which
$\mu=\tfrac1{N_{\mathrm{tot}}}\sum_k S_k$ and
$\sigma^2=\tfrac1{N_{\mathrm{tot}}}\sum_k Q_k-\mu^2$. In the clear this
is a pair of `colSums`.

```{r cvxr-standardize-plain, eval=RECOMPUTE}
pool_plain <- function(sites, n_total) {
    s <- Reduce(`+`, lapply(sites, function(s) colSums(s$X)))
    q <- Reduce(`+`, lapply(sites, function(s) colSums(s$X^2)))
    mu     <- s / n_total
    sigma2 <- pmax(q / n_total - mu^2, .Machine$double.eps)
    list(mu = mu, sigma = sqrt(sigma2))
}
pool      <- pool_plain(sites_raw, N_total)
sites_std <- lapply(sites_raw, function(s) list(
    name = s$name,
    X    = sweep(sweep(s$X, 2, pool$mu, "-"), 2, pool$sigma, "/"),
    time = s$time, status = s$status))
```

Rather than fit all $p = `r P_raw`$ probes, we pre-screen to the
$K = `r K`$ probes with the strongest univariate association with survival — the
screen-then-fit pattern @BayleFanLou2025 use on this cohort. For probe
$g$ on stratum $k$ in event-time order with risk set $R_i^{(k)}$ at
event $i$,
$U_g^{(k)}=\sum_{i\in k,\delta_i=1}(X_{i,g}-\overline X_{R_i^{(k)},g})$
is the score and
$I_g^{(k)}=\sum_{i\in k,\delta_i=1}\widehat{\operatorname{Var}}_{R_i^{(k)}}(X_{:,g})$
the information; both sum across sites, and the screen ranks probes by
$|Z_g|=|U_g|/\sqrt{I_g}$.

```{r cvxr-screen-plain, eval=RECOMPUTE}
K <- 100L

score_info_at_zero <- function(X, time, status) {
    p <- ncol(X)
    ord <- order(time)
    X_o <- X[ord, , drop = FALSE]; stat_o <- status[ord]; time_o <- time[ord]
    n <- nrow(X_o); U <- numeric(p); I <- numeric(p)
    ## Risk set at event i: every row with time >= t_i, tied rows included.
    first <- match(time_o, time_o)
    for (i in seq_len(n)) {
        if (stat_o[i] == 1L) {
            risk <- X_o[first[i]:n, , drop = FALSE]
            mu_R <- colMeans(risk)
            U <- U + (X_o[i, ] - mu_R)
            I <- I + colSums(sweep(risk, 2, mu_R, "-")^2) / nrow(risk)
        }
    }
    list(U = U, I = I)
}

screen_plain <- function(sites, K) {
    UI <- lapply(sites, function(s) score_info_at_zero(s$X, s$time, s$status))
    U  <- Reduce(`+`, lapply(UI, `[[`, "U"))
    I  <- Reduce(`+`, lapply(UI, `[[`, "I"))
    Z  <- U / sqrt(pmax(I, .Machine$double.eps))
    order(abs(Z), decreasing = TRUE)[seq_len(K)]
}
top_idx  <- screen_plain(sites_std, K)
sites_KS <- lapply(sites_std, function(s) list(
    name = s$name, X = s$X[, top_idx],
    time = s$time, status = s$status))
sigma_K  <- pool$sigma[top_idx]
```

`sites_KS` now holds each stratum on the $K = `r K`$ screened probes,
and `sigma_K` keeps their pooled SDs for the back-transform to the
original gene-expression scale.

The centralized stratified Cox-lasso fit — our ground truth — is a
single [CVXR][] solve summing the per-stratum Breslow partial
likelihoods plus the L1 penalty.

```{r cvxr-centralized, eval=RECOMPUTE}
LAMBDA <- 5
build_cox_breslow_nll <- function(beta_var, X_s, time_s, status_s) {
    ord <- order(time_s)
    X_o <- X_s[ord, , drop = FALSE]; stat_o <- status_s[ord]
    time_o <- time_s[ord]
    n <- nrow(X_o); eta_o <- X_o %*% beta_var
    ## Risk set at event i: every row with time >= t_i, tied rows included.
    first <- match(time_o, time_o)
    terms <- list()
    for (i in seq_len(n)) {
        if (stat_o[i] == 1L) {
            terms[[length(terms) + 1L]] <-
                log_sum_exp(eta_o[first[i]:n, 1]) - eta_o[i, 1]
        }
    }
    Reduce(`+`, terms)
}

beta_var <- Variable(K, name = "beta")
nll_per  <- lapply(sites_KS, function(s)
    build_cox_breslow_nll(beta_var, s$X, s$time, s$status))
agg_prob <- Problem(Minimize(Reduce(`+`, nll_per) +
                                 LAMBDA * p_norm(beta_var, 1)))
suppressMessages(suppressWarnings(
    psolve(agg_prob, solver = "CLARABEL", verbose = FALSE)))
agg_beta <- as.numeric(value(beta_var))
```

The distributed fit splits this objective into per-site subproblems
tied by a consensus variable. Each local problem is built so DPP
applies: $z$ and $u$ enter as `Parameter`s whose values change every
ADMM iteration while the symbolic structure does not; $X_k$, time,
status, and $\rho$ are constants.

```{r cvxr-local, eval=RECOMPUTE}
RHO <- 50

build_local <- function(X_k, time_k, status_k, rho) {
    p <- ncol(X_k)
    x  <- Variable(p); zp <- Parameter(p); up <- Parameter(p)
    nll <- build_cox_breslow_nll(x, X_k, time_k, status_k)
    aug <- (rho / 2) * sum_squares(x - zp + up)
    list(prob = Problem(Minimize(nll + aug)), x = x, zp = zp, up = up)
}
sites_problem <- lapply(sites_KS, function(s)
    build_local(s$X, s$time, s$status, RHO))
```

The ADMM driver below is the whole federated algorithm. How the
cross-site average is formed is left to its `consensus` argument, a
function of the per-site $(x_k,u_k)$ vectors. Everything else (the
local [CVXR][] solves, the soft-threshold $z$-update, the dual
update, the stopping rule) is ordinary R. We call it now with an
unencrypted average, and later, unchanged, with an encrypted one.

A note on the constants. We fix $\rho = `r RHO`$ and a cap of
`r MAX_ITER` iterations, and stop when the absolute rule
`primal < TOL && dual < TOL` holds, with `TOL = `r TOL``; the dual
residual is $\rho\lVert z^{t+1}-z^t\rVert$.

```{r cvxr-admm, eval=RECOMPUTE}
MAX_ITER <- 200L; TOL <- 5e-3
soft_threshold <- function(v, tau) sign(v) * pmax(abs(v) - tau, 0)

run_admm <- function(sites_problem, consensus) {
    site_x <- replicate(N_sites, rep(0, K), simplify = FALSE)
    site_u <- replicate(N_sites, rep(0, K), simplify = FALSE)
    z_curr <- rep(0, K); trajectory <- list()
    for (iter in seq_len(MAX_ITER)) {
        for (i in seq_len(N_sites)) {
            value(sites_problem[[i]]$zp) <- z_curr
            value(sites_problem[[i]]$up) <- site_u[[i]]
            suppressMessages(suppressWarnings(
                psolve(sites_problem[[i]]$prob, solver = "CLARABEL",
                       verbose = FALSE)))
            site_x[[i]] <- as.numeric(value(sites_problem[[i]]$x))
        }
        w_avg  <- consensus(site_x, site_u)
        z_new  <- soft_threshold(w_avg, LAMBDA / (N_sites * RHO))
        site_u <- Map(function(u, x) u + (x - z_new), site_u, site_x)
        primal <- sqrt(mean(vapply(site_x, function(x) sum((x - z_new)^2), 0)))
        dual   <- RHO * sqrt(sum((z_new - z_curr)^2))
        z_curr <- z_new; trajectory[[iter]] <- z_new
        if (primal < TOL && dual < TOL) break
    }
    list(z = z_curr, trajectory = trajectory)
}

plain_consensus <- function(site_x, site_u)
    Reduce(`+`, Map(`+`, site_x, site_u)) / length(site_x)

ref        <- run_admm(sites_problem, plain_consensus)
z_ref      <- ref$z
n_iter_ref <- length(ref$trajectory)
```

In the clear the consensus is a single line — the average of the
$(x_k+u_k)$ vectors. The unencrypted ADMM converges in `r n_iter_ref`
iterations and matches the centralized [CVXR][] fit to
`r sci(max(abs(z_ref - agg_beta)))` in maximum absolute
coefficient difference, so `agg_beta` — equivalently `z_ref` — is the
target the encrypted protocol must reproduce.

## The same fit under threshold FHE

Four quantities cross a site boundary: the standardization moments
$(S_k,Q_k)$, the screening statistics $(U^{(k)},I^{(k)})$, and, at each
ADMM iteration, the consensus sum $\sum_k(x_k+u_k)$ and the squared
primal residuals $\lVert x_k-z\rVert^2$ of the stopping rule.
`run_admm` forms the primal residual in the clear here; in a deployment
it is one more sum of the same kind. Each of the first three is a
sum over sites, so each becomes one round of the same threshold-FHE
summation primitive the master/worker fits use: every site encrypts its
contribution under the joint public key, the aggregator adds the
encrypted contributions, and the total is recovered by $n$-of-$n$ partial
decryption. For the consensus, the encrypted total is first scaled by
$1/N$, so what is decrypted is the average $\bar w$. The local [CVXR][] work and `run_admm` are untouched.

We use the same threshold infrastructure as `cox-threshold.Rmd`: a CKKS
context with `Feature$MULTIPARTY` and `make_threshold_master()`, which
walks the `multiparty_key_gen()` chain across the sites. Each site
generates its own share and keeps it; only public keys move along the
chain, so no party — the aggregator included — ever holds the joint
secret.

So the sites have to exist before the aggregator does. Here we
give each subgroup a site object to hold its share.

```{r cvxr-context, eval=RECOMPUTE}
cc <- fhe_context("CKKS",
                  multiplicative_depth = 1L,
                  scaling_mod_size     = 59L,
                  first_mod_size       = 60L,
                  batch_size           = 8192L,
                  features             = c(Feature$MULTIPARTY))

key_sites <- lapply(levels(dlbcl$Subgroup), function(s)
    make_worker(s, data = NULL, contribution_fn = function(data, theta) NULL))

master <- make_threshold_master("Aggregator",
                                crypto_context = cc, sites = key_sites)
```

Apart from the encrypted values in each round, this setup is the
only exchange between the aggregator and the sites. Each site kept
its own secret share and a copy of the public parameters, and
everything below encrypts with `encrypt(site, value)`, which uses
those parameters. `site_params(site)` returns them as an
`OpenFHEParams` object. Its class has two properties, the crypto
context and the joint public key, and none for a key share:

```{r cvxr-params-shown, purl=FALSE}
homomorpheR::OpenFHEParams
```

The standardization round is `pool_plain` with the two `colSums`
encrypted: each site encrypts $S_k$ and $Q_k$, the aggregator sums under
encryption and threshold-decrypts the pooled moments. (For brevity we
treat $N_{\mathrm{tot}}$ as known to the aggregator; hiding the per-site
head-counts is one more sum of the same kind.)

```{r cvxr-pool-encrypt, eval=RECOMPUTE}
## Site side. Each site forms its own moments and encrypts them under
## the joint key; what leaves is encrypted. Encrypting at the
## aggregator instead would mean handing it the per-site column sums in
## the clear first, which is the disclosure this round exists to avoid.
site_moments <- function(site, s)
    list(sum   = encrypt(site, colSums(s$X)),
         sumsq = encrypt(site, colSums(s$X^2)))

## Aggregator side. It receives only the encrypted moments, adds them,
## and decrypts the totals.
pool_encrypted <- function(master, parts, n_total, p_raw) {
    pooled_sum   <- decrypt(
        master, Reduce(`+`, lapply(parts, `[[`, "sum")),   len = p_raw)
    pooled_sumsq <- decrypt(
        master, Reduce(`+`, lapply(parts, `[[`, "sumsq")), len = p_raw)
    mu     <- pooled_sum / n_total
    sigma2 <- pmax(pooled_sumsq / n_total - mu^2, .Machine$double.eps)
    list(mu = mu, sigma = sqrt(sigma2))
}
moment_parts <- Map(site_moments, key_sites, sites_raw)   # at the sites
fhe_pool     <- pool_encrypted(master, moment_parts, N_total, P_raw)
pool_agree   <- list(mu    = max(abs(fhe_pool$mu    - pool$mu)),
                     sigma = max(abs(fhe_pool$sigma - pool$sigma)))
```

The encrypted moments agree with the unencrypted `pool` to
`r sci(pool_agree$mu)` (mean) and
`r sci(pool_agree$sigma)` (SD), the CKKS
approximation error. The screening round works the same way: the
`score_info_at_zero` summands $(U^{(k)},I^{(k)})$ are encrypted and
summed, and we confirm it selects the same probes.

```{r cvxr-screen-encrypt, eval=RECOMPUTE}
## Site side: compute the score and information at beta = 0 on the
## site's own rows, and encrypt both before returning them.
site_score_info <- function(site, s) {
    z <- score_info_at_zero(s$X, s$time, s$status)
    list(U = encrypt(site, z$U), I = encrypt(site, z$I))
}

## Aggregator side: it receives only the encrypted (U, I), adds them,
## and decrypts the totals.
screen_encrypted <- function(master, UI, p_raw, K) {
    U   <- decrypt(master, Reduce(`+`, lapply(UI, `[[`, "U")),
                   len = p_raw)
    I   <- decrypt(master, Reduce(`+`, lapply(UI, `[[`, "I")),
                   len = p_raw)
    Z   <- U / sqrt(pmax(I, .Machine$double.eps))
    order(abs(Z), decreasing = TRUE)[seq_len(K)]
}
UI_parts <- Map(site_score_info, key_sites, sites_std)   # at the sites
fhe_top  <- screen_encrypted(master, UI_parts, P_raw, K)
stopifnot(setequal(fhe_top, top_idx))   # same probes as the cleartext screen
```

Standardization and screening give the same design as in the
clear, so the consensus round is the only piece left to encrypt.
It mirrors `plain_consensus`. Each site encrypts its own
$(x_k+u_k)$. The aggregator sums the encrypted values, scales by
$1/N$ (one multiplication by an unencrypted constant, which uses
one level of multiplicative depth), and threshold-decrypts the
length-$K$ average. Soft-thresholding is applied to that average
in the clear at the aggregator.

```{r cvxr-consensus, eval=RECOMPUTE}
## Site side: site k forms x_k + u_k and encrypts it. What leaves the
## site is encrypted.
site_consensus_term <- function(site, x_k, u_k)
    encrypt(site, x_k + u_k)

## Aggregator side: it receives only the encrypted terms, adds them,
## scales by 1/N, and decrypts the average.
aggregate_consensus <- function(master, cts, K)
    decrypt(master, Reduce(`+`, cts) * (1 / length(cts)), len = K)

## One consensus round: the site step at every site, then the
## aggregator step.
encrypted_consensus <- function(site_x, site_u)
    aggregate_consensus(master,
                        Map(site_consensus_term, key_sites, site_x, site_u),
                        K)
fhe        <- run_admm(sites_problem, encrypted_consensus)
z_enc      <- fhe$z
trajectory <- fhe$trajectory
n_iter_enc <- length(trajectory)
```

```{r cvxr-assemble, eval=RECOMPUTE, echo=FALSE}
## (not shown in the rendered vignette; purled into the regeneration
## script) collect the primitives the vignette, manuscript, and data
## object share. Light derived quantities are recomputed downstream.
cvxr_consensus <- list(
    params     = list(K = K, LAMBDA = LAMBDA, RHO = RHO,
                      MAX_ITER = MAX_ITER, TOL = TOL),
    top_idx    = top_idx,
    sigma_K    = sigma_K,
    agg_beta   = agg_beta,
    z_ref      = z_ref,
    z_enc      = z_enc,
    trajectory = trajectory,
    n_iter_ref = n_iter_ref,
    n_iter_enc = n_iter_enc,
    pool_agree = pool_agree,
    screen_match = setequal(fhe_top, top_idx))
```

The only change is passing `encrypted_consensus` in place of
`plain_consensus`. The encrypted ADMM ran for `r n_iter_enc`
iterations, and its coefficients differ from the unencrypted run
by at most `r sci(max(abs(z_enc - z_ref)))`, the CKKS
approximation error.

## Comparison with the centralized fit

We compare the threshold-FHE consensus to the centralized [CVXR][] fit
on both the standardized scale (where ADMM lives) and the
back-transformed original gene-expression scale.

```{r cox-compare, purl=FALSE}
beta_orig_agg <- agg_beta / sigma_K
beta_orig_enc <- z_enc    / sigma_K
cmp <- data.frame(
    check.names    = FALSE,
    Scale          = c("standardized", "original"),
    `Max abs diff` = c(max(abs(z_enc - agg_beta)),
                       max(abs(beta_orig_enc - beta_orig_agg))),
    `L1 diff`      = c(sum(abs(z_enc - agg_beta)),
                       sum(abs(beta_orig_enc - beta_orig_agg))))
ktab(cmp, digits = 4,
             col.names = c("Scale",
                           "$\\max_j \\lvert \\hat\\beta_j^{\\text{ADMM}} - \\hat\\beta_j^{\\text{centralized}} \\rvert$",
                           "$\\lVert \\hat\\beta^{\\text{ADMM}} - \\hat\\beta^{\\text{centralized}} \\rVert_1$"),
             caption = "Threshold-FHE consensus ADMM vs. centralized Cox-lasso")

n_agg   <- sum(abs(agg_beta) > 1e-7)
n_enc   <- sum(abs(z_enc)    > 1e-7)
n_inter <- sum((abs(agg_beta) > 1e-7) & (abs(z_enc) > 1e-7))
```

The active set has `r n_agg` nonzero coefficients in the centralized fit
and `r n_enc` in the encrypted ADMM fit; the intersection is
`r n_inter` — every probe selected by the centralized fit is recovered
by the encrypted distributed protocol.

The figure below shows the consensus trajectory $z^t$ for the eight
probes with largest $|z|$ at convergence, with the centralized fit drawn
as a horizontal reference. The encrypted iterates follow ADMM's own
trajectory toward the centralized fit; they are not iterations of the
centralized solver.

```{r cox-fig, purl=FALSE, fig.width=8, fig.height=4.5, fig.cap="Consensus trajectories for the eight largest-magnitude coefficients. Solid lines are the encrypted ADMM iterates; dashed lines are the centralized CVXR fit. Standardized scale."}
top8 <- order(-abs(z_enc))[1:8]
labs <- paste("probe", colnames(DLBCL_gex)[top_idx][top8])
op <- par(mfrow = c(2, 4), mar = c(4, 4, 2, 1), cex = 0.8)
for (j in seq_along(top8)) {
    vals <- vapply(trajectory, function(z) z[top8[j]], numeric(1))
    plot(seq_along(vals), vals, type = "l", lwd = 1.6, col = "steelblue4",
         xlab = "ADMM iteration", ylab = expression(z^t), main = labs[j])
    abline(h = agg_beta[top8[j]], lty = 2)
}
par(op)
```

## What the protocol hides and reveals

The aggregator learns the pooled per-probe mean and SD (round 1), the
pooled per-probe univariate Cox score $U_g$ and information $I_g$ and the indices of
the top-`r K` screened probes (round 2), and the cross-site average
$\bar w$ before soft-thresholding at every ADMM iteration. Each is an aggregated population-level statistic
over `r N_total` patients, not patient-level data. None of the per-site
design matrices $X_k$, the per-site $(\mu,\sigma^2)$ contributions, the
per-site $(U,I)$ contributions, or the per-site $(x_k+u_k)$ vectors
reaches the aggregator in cleartext. Decryption at every step
is $n$-of-$n$ threshold: no single party — the aggregator included — can
recover any intermediate quantity unilaterally.

Two steps in this vignette would not exist in a real deployment.
The unencrypted reference pass forms the per-site averages that
the protocol keeps encrypted, and the centralized solve pools all
the data. We can run both only because this vignette holds all
three sites' data in one R session. A deployment also has to choose
$\rho$ without the centralized solve.

A deployment can instead choose $\rho$ on a surrogate cohort built
from public design facts, as `vignette("cvxr-consensus-admm-dp")`
does. When the
protocol also spends a privacy budget, as in
`vignette("cvxr-consensus-admm-dp")`, encrypting the sweep is not
enough, and the constants have to be chosen from data that does not
count against the budget.

## Discussion

1. **CVXR problems work unchanged.** Each site solves its local
   Cox-lasso problem in the clear. Only the consensus update is
   encrypted. The [CVXR][] `Problem(Minimize(...))` is the same as
   in the clear, and `run_admm` is called with
   `encrypted_consensus` instead of `plain_consensus`.
2. **Encryption does not change the fit.** The encrypted fit
   matches the unencrypted ADMM to
   `r sci(max(abs(z_enc - z_ref)))` and has the same active
   set. The remaining gap to the centralized solve
   (`r sci(max(abs(z_ref - agg_beta)))`) comes from stopping
   ADMM at a finite tolerance, not from the encryption.
3. **No single party holds the secret key.** Each site generated
   its own share during key generation and kept it. The aggregator
   holds only the joint public key. No single party can decrypt an
   intermediate value, and each round's result is decrypted only
   when every site contributes a partial decryption.
4. **DPP keeps the inner loop fast.** Each site's [CVXR][] problem is
   built once at setup; ADMM iterations only update the `Parameter`
   values.

## Limitations

- **Honest-but-curious trust.** A site that misreports its local
  $(x_k+u_k)$ can corrupt the consensus; detecting this needs
  commitments / zero-knowledge proofs not implemented here.
- **The aggregator sees $\bar w$ at every iteration.** It is
  decrypted before soft-thresholding, so it includes the coordinates
  that $z^t$ sets to zero.
- **No output privacy.** The released $\hat\beta = z^\star$ is the same
  coefficient vector as the centralized fit; output-level protection is
  out of scope here and is demonstrated in the companion DP vignette.

[CVXR]: https://cvxr.rbind.io
