---
title: "Federated Cox-Lasso via Consensus ADMM on DLBCL (Threshold CKKS)"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Federated Cox-Lasso via Consensus ADMM on DLBCL (Threshold CKKS)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r knitr-opts, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error   = FALSE,
  tidy    = FALSE,
  cache   = FALSE
)
```

```{r setup, purl=FALSE, include=FALSE}
## This vignette is the single source of truth for the encrypted
## Cox-lasso consensus-ADMM example. The code chunks below contain the
## real pipeline, gated `eval = RECOMPUTE`: RECOMPUTE defaults to FALSE
## here, so the chunks are shown but NOT run when the vignette builds
## (the ~150-iteration encrypted ADMM is far too slow for that). The
## displayed numbers are read from data(cvxr_consensus), which was
## produced by knitr::purl()-ing exactly these chunks and running them
## (see data-raw/cvxr_consensus.R and the verification note below). The
## openfhe-jss manuscript reads the same purled chunks, so vignette,
## script, manuscript, and data object cannot drift.
if (!exists("RECOMPUTE")) RECOMPUTE <- FALSE

suppressPackageStartupMessages(library(homomorpheR))
data(cvxr_consensus, package = "homomorpheR")
data(DLBCL,          package = "homomorpheR")
data(DLBCL_gex,      package = "homomorpheR")

K          <- cvxr_consensus$params$K
LAMBDA     <- cvxr_consensus$params$LAMBDA
RHO        <- cvxr_consensus$params$RHO
MAX_ITER   <- cvxr_consensus$params$MAX_ITER
TOL        <- cvxr_consensus$params$TOL
P_raw      <- ncol(DLBCL_gex)
N_total    <- nrow(DLBCL)
top_idx    <- cvxr_consensus$top_idx
sigma_K    <- cvxr_consensus$sigma_K
agg_beta   <- cvxr_consensus$agg_beta
z_ref      <- cvxr_consensus$z_ref
z_enc      <- cvxr_consensus$z_enc
trajectory <- cvxr_consensus$trajectory
n_iter_ref <- cvxr_consensus$n_iter_ref
n_iter_enc <- cvxr_consensus$n_iter_enc
```

## Introduction

The companion vignette
[Federated Consensus ADMM](cvxr-consensus-admm.html) develops the
threshold-FHE consensus-ADMM protocol on a small simulated
L2-regularized logistic problem ($p = 4$, three sites). This vignette
takes the same protocol to a real, high-dimensional survival problem:
a stratified **Cox lasso** on the diffuse large-B-cell lymphoma (DLBCL)
gene-expression cohort of @Rosenwald2002, the dataset @BayleFanLou2025
use to motivate distributed Cox estimation.

The lesson is the same as in the `cox` and `cox-threshold` vignettes —
*function-call* optimizers such as `mle()` or `coxph()`'s
Newton-Raphson compose with an encrypted master/worker round without
the optimizer noticing — but it does not extend to arbitrary convex
programs. [CVXR][] operates on a *symbolic* problem: it canonicalizes
once and ships the canonical form to a solver, so we cannot hand it a
callback that secretly performs FHE. The remedy is **federated
consensus ADMM** (Boyd, Parikh, Chu, Peleato and Eckstein, 2011): split
the global problem into per-site subproblems each solved by [CVXR][] in
the clear, and use the encrypted channel only for the cross-site
**consensus average**. Disciplined parametric programming (DPP) keeps
the per-site solves fast because the `Parameter` values change every
iteration but the symbolic structure does not.

We develop the fit in two passes. First we run the whole thing **in the
clear** to fix the target: standardize, screen, and solve the consensus
ADMM with an ordinary plaintext average, checking against the
centralized [CVXR][] solve. Then we put it **under threshold FHE**,
replacing each cross-site sum — the standardization moments, the
screening statistics, and the per-iteration consensus average — with
one round of the encrypted-summation primitive, and confirm the
encrypted fit reproduces the plaintext reference.

> **Reproducibility and verification.** The code chunks in this
> vignette *are* the pipeline; they are gated `eval = RECOMPUTE` and do
> not run when the vignette builds (the encrypted ADMM takes ~150
> iterations of per-site [CVXR][] solves). The displayed numbers come
> from `data(cvxr_consensus)`, which was produced by extracting these
> very chunks with `knitr::purl()` and running them. To verify the
> results, do exactly that yourself:
>
> ```r
> vig <- system.file("doc", "cvxr-cox-lasso-dlbcl.Rmd",
>                    package = "homomorpheR")
> RECOMPUTE <- TRUE                                  # un-gate the chunks
> src <- knitr::purl(vig, output = tempfile(fileext = ".R"), quiet = TRUE)
> source(src)                                        # runs the pipeline (minutes)
> fresh <- cvxr_consensus                            # just recomputed
> data(cvxr_consensus, package = "homomorpheR")      # the shipped copy
> max(abs(fresh$z_enc - cvxr_consensus$z_enc))       # ~1e-7
> ```
>
> Threshold key generation is randomized, so a re-run reproduces the
> shipped coefficients up to CKKS approximation noise (~1e-7), not
> bit-for-bit. This `purl()`-then-`source()` is exactly what
> `data-raw/cvxr_consensus.R` does to build the shipped object.

## The global problem and its consensus split

The three molecular subgroups (GCB, ABC, Type III) are our sites: they
arise from different cells of origin, have systematically different
prognosis, and are diagnosed at different referral centers, so the
subgroup is a biologically and operationally realistic site boundary.
With $N$ sites, per-site standardized design matrices
$X_k \in \mathbb{R}^{n_k \times p}$, event times $t_k$, status
$\delta_k$, a shared coefficient $\beta \in \mathbb{R}^p$, and the
per-stratum Cox partial log-likelihood $\ell_k$ in Breslow form, the
global problem is

$$
\min_{\beta \in \mathbb{R}^p}\;
-\sum_{k=1}^{N}\ell_k(\beta)\;+\;\lambda\lVert\beta\rVert_1 .
$$

Because the partial likelihood factorizes additively across strata, the
consensus split introduces per-site copies $x_k$ and a global consensus
$z$:

$$
\min_{\{x_k\},\,z}\;
\sum_k\bigl(-\ell_k(x_k)\bigr) + \lambda\lVert z\rVert_1
\quad\text{s.t.}\quad x_k = z,\ \forall k,
$$

with augmented-Lagrangian iteration

$$
\begin{aligned}
x_k^{t+1} &= \arg\min_x -\ell_k(x) + \tfrac{\rho}{2}\lVert x - z^t + u_k^t\rVert_2^2,\\
z^{t+1}   &= S_{\lambda/(N\rho)}\!\Bigl(\tfrac{1}{N}\sum_k (x_k^{t+1}+u_k^t)\Bigr),\\
u_k^{t+1} &= u_k^t + (x_k^{t+1}-z^{t+1}),
\end{aligned}
$$

where $S_\tau(v)=\operatorname{sign}(v)\max(|v|-\tau,0)$ is
soft-thresholding (the proximal map of $\tau\lVert\cdot\rVert_1$). The
$x$-update is the per-site [CVXR][] solve; the $z$-update needs the
cross-site average $\bar w=\tfrac1N\sum_k(x_k+u_k)$ — that average is
the only quantity that traverses the encrypted channel; the
soft-threshold is closed-form at the aggregator and adds no
cryptographic depth.

The pipeline needs `survival` and [CVXR][] alongside the encryption
packages.

```{r cvxr-libs, eval=RECOMPUTE}
library(survival)
library(CVXR)
library(openfhe)
library(homomorpheR)
```

## The federated fit in the clear

The three subgroups are our sites. We load the cohort, split it by
subgroup, and keep each site's raw expression matrix, event times, and
status.

```{r cvxr-data, eval=RECOMPUTE}
data(DLBCL,     package = "homomorpheR")  # 235 patients: survival + signatures
data(DLBCL_gex, package = "homomorpheR")  # 235 patients x 6416 Lymphochip probes
dlbcl <- DLBCL
dlbcl$Subgroup <- factor(dlbcl$Subgroup, levels = c("GCB","ABC","Type III"))
stopifnot(identical(as.character(dlbcl$ID), rownames(DLBCL_gex)))  # rows aligned

sites_raw <- lapply(levels(dlbcl$Subgroup), function(s) {
    idx <- which(dlbcl$Subgroup == s)
    list(name = s, X = DLBCL_gex[idx, ], time = dlbcl$time[idx],
         status = dlbcl$status[idx])
})
names(sites_raw) <- levels(dlbcl$Subgroup)
N_sites <- length(sites_raw)
N_total <- sum(vapply(sites_raw, function(s) nrow(s$X), 1L))
P_raw   <- ncol(DLBCL_gex)
```

We standardize the features so the L1 penalty applies uniformly across
predictors on different scales. The pooled mean and variance are
additive over sites: site $k$ contributes $S_k=\sum_{i\in k}x_i$ and
$Q_k=\sum_{i\in k}x_i^2$, from which
$\mu=\tfrac1{N_{\mathrm{tot}}}\sum_k S_k$ and
$\sigma^2=\tfrac1{N_{\mathrm{tot}}}\sum_k Q_k-\mu^2$. In the clear this
is a pair of `colSums`.

```{r cvxr-standardize-plain, eval=RECOMPUTE}
pool_plain <- function(sites, n_total) {
    s <- Reduce(`+`, lapply(sites, function(s) colSums(s$X)))
    q <- Reduce(`+`, lapply(sites, function(s) colSums(s$X^2)))
    mu     <- s / n_total
    sigma2 <- pmax(q / n_total - mu^2, .Machine$double.eps)
    list(mu = mu, sigma = sqrt(sigma2))
}
pool      <- pool_plain(sites_raw, N_total)
sites_std <- lapply(sites_raw, function(s) list(
    name = s$name,
    X    = sweep(sweep(s$X, 2, pool$mu, "-"), 2, pool$sigma, "/"),
    time = s$time, status = s$status))
```

Solving the full Cox-lasso at $p = `r P_raw`$ exhausts memory during
[CVXR][] canonicalization, so we pre-screen to the $K = `r K`$ probes
with the strongest univariate association with survival — the
screen-then-fit pattern @BayleFanLou2025 use on this cohort. For probe
$g$ on stratum $k$ in event-time order with risk set $R_i^{(k)}$ at
event $i$,
$U_g^{(k)}=\sum_{i\in k,\delta_i=1}(X_{i,g}-\overline X_{R_i^{(k)},g})$
is the score and
$I_g^{(k)}=\sum_{i\in k,\delta_i=1}\widehat{\operatorname{Var}}_{R_i^{(k)}}(X_{:,g})$
the information; both sum across sites, and the screen ranks probes by
$|Z_g|=|U_g|/\sqrt{I_g}$.

```{r cvxr-screen-plain, eval=RECOMPUTE}
K <- 100L

score_info_at_zero <- function(X, time, status) {
    p <- ncol(X)
    ord <- order(time, -status)
    X_o <- X[ord, , drop = FALSE]; stat_o <- status[ord]
    n <- nrow(X_o); U <- numeric(p); I <- numeric(p)
    for (i in seq_len(n)) {
        if (stat_o[i] == 1L) {
            risk <- X_o[i:n, , drop = FALSE]
            mu_R <- colMeans(risk)
            U <- U + (X_o[i, ] - mu_R)
            I <- I + colSums(sweep(risk, 2, mu_R, "-")^2) / nrow(risk)
        }
    }
    list(U = U, I = I)
}

screen_plain <- function(sites, K) {
    UI <- lapply(sites, function(s) score_info_at_zero(s$X, s$time, s$status))
    U  <- Reduce(`+`, lapply(UI, `[[`, "U"))
    I  <- Reduce(`+`, lapply(UI, `[[`, "I"))
    Z  <- U / sqrt(pmax(I, .Machine$double.eps))
    order(abs(Z), decreasing = TRUE)[seq_len(K)]
}
top_idx  <- screen_plain(sites_std, K)
sites_KS <- lapply(sites_std, function(s) list(
    name = s$name, X = s$X[, top_idx],
    time = s$time, status = s$status))
sigma_K  <- pool$sigma[top_idx]
```

`sites_KS` now holds each stratum on the $K = `r K`$ screened probes,
and `sigma_K` keeps their pooled SDs for the back-transform to the
original gene-expression scale.

The centralized stratified Cox-lasso fit — our ground truth — is a
single [CVXR][] solve summing the per-stratum Breslow partial
likelihoods plus the L1 penalty.

```{r cvxr-centralized, eval=RECOMPUTE}
LAMBDA <- 5
build_cox_breslow_nll <- function(beta_var, X_s, time_s, status_s) {
    ord <- order(time_s, -status_s)
    X_o <- X_s[ord, , drop = FALSE]; stat_o <- status_s[ord]
    n <- nrow(X_o); eta_o <- X_o %*% beta_var
    terms <- list()
    for (i in seq_len(n)) {
        if (stat_o[i] == 1L) {
            terms[[length(terms) + 1L]] <-
                log_sum_exp(eta_o[i:n, 1]) - eta_o[i, 1]
        }
    }
    Reduce(`+`, terms)
}

beta_var <- Variable(K, name = "beta")
nll_per  <- lapply(sites_KS, function(s)
    build_cox_breslow_nll(beta_var, s$X, s$time, s$status))
agg_prob <- Problem(Minimize(Reduce(`+`, nll_per) +
                                 LAMBDA * p_norm(beta_var, 1)))
suppressMessages(suppressWarnings(
    psolve(agg_prob, solver = "CLARABEL", verbose = FALSE)))
agg_beta <- as.numeric(value(beta_var))
```

The distributed fit splits this objective into per-site subproblems
tied by a consensus variable. Each local problem is built so DPP
applies: $z$ and $u$ enter as `Parameter`s whose values change every
ADMM iteration while the symbolic structure does not; $X_k$, time,
status, and $\rho$ are constants.

```{r cvxr-local, eval=RECOMPUTE}
RHO <- 50

build_local <- function(X_k, time_k, status_k, rho) {
    p <- ncol(X_k)
    x  <- Variable(p); zp <- Parameter(p); up <- Parameter(p)
    nll <- build_cox_breslow_nll(x, X_k, time_k, status_k)
    aug <- (rho / 2) * sum_squares(x - zp + up)
    list(prob = Problem(Minimize(nll + aug)), x = x, zp = zp, up = up)
}
sites_problem <- lapply(sites_KS, function(s)
    build_local(s$X, s$time, s$status, RHO))
```

The ADMM driver below is the whole federated algorithm, and it is
deliberately agnostic to *how* the cross-site average is formed: the
`consensus` argument is a function of the per-site $(x_k,u_k)$ vectors,
and everything else — the local [CVXR][] solves, the soft-threshold
$z$-update, the dual update, the stopping rule — is ordinary R. We call
it once now with a plaintext average and, unchanged, once more under
encryption.

A note on the constants. We fix $\rho = `r RHO`$ and a cap of
`r MAX_ITER` iterations. The dual residual $\rho\lVert z^{t+1}-z^t\rVert$
is the binding term here and decays slowly; with $\rho = `r RHO`$ the
absolute stopping rule `primal < TOL && dual < TOL` (with
`TOL = `r TOL``) trips at iteration `r n_iter_ref`. Smaller $\rho$
reaches the tolerance in fewer iterations but at a looser fit, so we
keep $\rho = `r RHO`$ for the tightest agreement with the centralized
solve.

```{r cvxr-admm, eval=RECOMPUTE}
MAX_ITER <- 200L; TOL <- 5e-3
soft_threshold <- function(v, tau) sign(v) * pmax(abs(v) - tau, 0)

run_admm <- function(sites_problem, consensus) {
    site_x <- replicate(N_sites, rep(0, K), simplify = FALSE)
    site_u <- replicate(N_sites, rep(0, K), simplify = FALSE)
    z_curr <- rep(0, K); trajectory <- list()
    for (iter in seq_len(MAX_ITER)) {
        for (i in seq_len(N_sites)) {
            value(sites_problem[[i]]$zp) <- z_curr
            value(sites_problem[[i]]$up) <- site_u[[i]]
            suppressMessages(suppressWarnings(
                psolve(sites_problem[[i]]$prob, solver = "CLARABEL",
                       verbose = FALSE)))
            site_x[[i]] <- as.numeric(value(sites_problem[[i]]$x))
        }
        w_avg  <- consensus(site_x, site_u)
        z_new  <- soft_threshold(w_avg, LAMBDA / (N_sites * RHO))
        site_u <- Map(function(u, x) u + (x - z_new), site_u, site_x)
        primal <- sqrt(mean(vapply(site_x, function(x) sum((x - z_new)^2), 0)))
        dual   <- RHO * sqrt(sum((z_new - z_curr)^2))
        z_curr <- z_new; trajectory[[iter]] <- z_new
        if (primal < TOL && dual < TOL) break
    }
    list(z = z_curr, trajectory = trajectory)
}

plain_consensus <- function(site_x, site_u)
    Reduce(`+`, Map(`+`, site_x, site_u)) / length(site_x)

ref   <- run_admm(sites_problem, plain_consensus)
z_ref <- ref$z
```

In the clear the consensus is a single line — the average of the
$(x_k+u_k)$ vectors. The plaintext ADMM converges in `r n_iter_ref`
iterations and matches the centralized [CVXR][] fit to
`r signif(max(abs(z_ref - agg_beta)), 2)` in maximum absolute
coefficient difference, so `agg_beta` — equivalently `z_ref` — is the
target the encrypted protocol must reproduce.

## The same fit under threshold FHE

Only three quantities ever cross a site boundary: the standardization
moments $(S_k,Q_k)$, the screening statistics $(U^{(k)},I^{(k)})$, and,
at each ADMM iteration, the consensus sum $\sum_k(x_k+u_k)$. Each is a
sum over sites, so each becomes one round of the same threshold-FHE
summation primitive the master/worker fits use: every site encrypts its
contribution under the joint public key, the aggregator adds the
ciphertexts, and the total is recovered by $n$-of-$n$ partial
decryption. The local [CVXR][] work and `run_admm` are untouched.

We use the same threshold infrastructure as `cox-threshold.Rmd`: a CKKS
context with `Feature$MULTIPARTY` and `make_threshold_master()`, which
runs the chained `multiparty_key_gen()` ceremony internally so no single
party ever holds the secret key.

```{r cvxr-context, eval=RECOMPUTE}
cc <- fhe_context("CKKS",
                  multiplicative_depth = 1L,
                  scaling_mod_size     = 59L,
                  first_mod_size       = 60L,
                  batch_size           = 8192L,
                  features             = c(Feature$MULTIPARTY))
master <- make_threshold_master("Aggregator",
                                crypto_context = cc, n_sites = N_sites)
```

The standardization round is `pool_plain` with the two `colSums`
encrypted: each site encrypts $S_k$ and $Q_k$, the aggregator sums under
encryption and threshold-decrypts the pooled moments. (For brevity we
treat $N_{\mathrm{tot}}$ as known to the aggregator; hiding the per-site
head-counts is one more sum of the same kind.)

```{r cvxr-pool-encrypt, eval=RECOMPUTE}
encrypt_pool <- function(master, sites, n_total, p_raw) {
    s_ct <- lapply(sites, function(s) master_encrypt(master, colSums(s$X)))
    q_ct <- lapply(sites, function(s) master_encrypt(master, colSums(s$X^2)))
    pooled_sum   <- master_decrypt(master, Reduce(`+`, s_ct), len = p_raw)
    pooled_sumsq <- master_decrypt(master, Reduce(`+`, q_ct), len = p_raw)
    mu     <- pooled_sum / n_total
    sigma2 <- pmax(pooled_sumsq / n_total - mu^2, .Machine$double.eps)
    list(mu = mu, sigma = sqrt(sigma2))
}
fhe_pool <- encrypt_pool(master, sites_raw, N_total, P_raw)
```

The encrypted moments agree with the plaintext `pool` to
`r signif(cvxr_consensus$pool_agree$mu, 2)` (mean) and
`r signif(cvxr_consensus$pool_agree$sigma, 2)` (SD) — essentially
machine precision, since these are exact sums under CKKS. The screening
round is structurally identical, the same `score_info_at_zero` summands
$(U^{(k)},I^{(k)})$ encrypted and summed the same way, so we show it
compactly and confirm it selects the same probes.

```{r cvxr-screen-encrypt, eval=RECOMPUTE}
encrypt_screen <- function(master, sites, p_raw, K) {
    enc <- function(v) master_encrypt(master, v)
    UI  <- lapply(sites, function(s) score_info_at_zero(s$X, s$time, s$status))
    U   <- master_decrypt(master, Reduce(`+`, lapply(UI, function(z) enc(z$U))),
                          len = p_raw)
    I   <- master_decrypt(master, Reduce(`+`, lapply(UI, function(z) enc(z$I))),
                          len = p_raw)
    Z   <- U / sqrt(pmax(I, .Machine$double.eps))
    order(abs(Z), decreasing = TRUE)[seq_len(K)]
}
fhe_top <- encrypt_screen(master, sites_std, P_raw, K)
stopifnot(setequal(fhe_top, top_idx))   # same probes as the plaintext screen
```

With standardization and screening recovering the same design, the
consensus round is the only piece left to encrypt. It mirrors
`plain_consensus`: encrypt each $(x_k+u_k)$, sum the ciphertexts, scale
by $1/N$ under encryption (one ciphertext-plaintext multiply, depth 1),
and threshold-decrypt the length-$K$ average. Soft-thresholding stays in
the clear at the aggregator.

```{r cvxr-consensus, eval=RECOMPUTE}
encrypted_consensus <- function(site_x, site_u) {
    cts    <- lapply(seq_along(site_x), function(i)
                  master_encrypt(master, site_x[[i]] + site_u[[i]]))
    ct_avg <- Reduce(`+`, cts) * (1 / length(site_x))
    master_decrypt(master, ct_avg, len = K)
}
fhe        <- run_admm(sites_problem, encrypted_consensus)
z_curr     <- fhe$z
trajectory <- fhe$trajectory
```

```{r cvxr-assemble, eval=RECOMPUTE, echo=FALSE}
## (not shown in the rendered vignette; purled into the regeneration
## script) collect the primitives the vignette, manuscript, and data
## object share. Light derived quantities are recomputed downstream.
cvxr_consensus <- list(
    params     = list(K = K, LAMBDA = LAMBDA, RHO = RHO,
                      MAX_ITER = MAX_ITER, TOL = TOL),
    top_idx    = top_idx,
    sigma_K    = sigma_K,
    agg_beta   = agg_beta,
    z_ref      = z_ref,
    z_enc      = z_curr,
    trajectory = trajectory,
    n_iter_ref = length(ref$trajectory),
    n_iter_enc = length(trajectory),
    pool_agree = list(mu    = max(abs(fhe_pool$mu    - pool$mu)),
                      sigma = max(abs(fhe_pool$sigma - pool$sigma))),
    screen_match = setequal(fhe_top, top_idx))
```

Passing `encrypted_consensus` in place of `plain_consensus` is the
*entire* change. The encrypted ADMM ran for `r n_iter_enc` iterations
and lands on the same coefficients as the plaintext run, differing by
only `r signif(max(abs(z_enc - z_ref)), 2)` — the CKKS approximation
noise.

## Comparison with the centralized fit

We compare the threshold-FHE consensus to the centralized [CVXR][] fit
on both the standardized scale (where ADMM lives) and the
back-transformed original gene-expression scale.

```{r cox-compare, purl=FALSE}
beta_orig_agg <- agg_beta / sigma_K
beta_orig_enc <- z_enc    / sigma_K
cmp <- data.frame(
    check.names    = FALSE,
    Scale          = c("standardized", "original"),
    `Max abs diff` = c(max(abs(z_enc - agg_beta)),
                       max(abs(beta_orig_enc - beta_orig_agg))),
    `L1 diff`      = c(sum(abs(z_enc - agg_beta)),
                       sum(abs(beta_orig_enc - beta_orig_agg))))
knitr::kable(cmp, digits = 4,
             caption = "Threshold-FHE consensus ADMM vs. centralized Cox-lasso")

n_agg   <- sum(abs(agg_beta) > 1e-7)
n_enc   <- sum(abs(z_enc)    > 1e-7)
n_inter <- sum((abs(agg_beta) > 1e-7) & (abs(z_enc) > 1e-7))
```

The active set has `r n_agg` nonzero coefficients in the centralized fit
and `r n_enc` in the encrypted ADMM fit; the intersection is
`r n_inter` — every probe selected by the centralized fit is recovered
by the encrypted distributed protocol.

The figure below shows the consensus trajectory $z^t$ for the eight
probes with largest $|z|$ at convergence, with the centralized fit drawn
as a horizontal reference. The encrypted iterates converge along the
path the centralized solver would take.

```{r cox-fig, purl=FALSE, fig.width=8, fig.height=4.5, fig.cap="Consensus trajectories for the eight largest-magnitude coefficients. Solid lines are the encrypted ADMM iterates; dashed lines are the centralized CVXR fit. Standardized scale."}
top8 <- order(-abs(z_enc))[1:8]
labs <- paste("probe", colnames(DLBCL_gex)[top_idx][top8])
op <- par(mfrow = c(2, 4), mar = c(4, 4, 2, 1), cex = 0.8)
for (j in seq_along(top8)) {
    vals <- vapply(trajectory, function(z) z[top8[j]], numeric(1))
    plot(seq_along(vals), vals, type = "l", lwd = 1.6, col = "steelblue4",
         xlab = "ADMM iteration", ylab = expression(z^t), main = labs[j])
    abline(h = agg_beta[top8[j]], lty = 2)
}
par(op)
```

## What the protocol hides and reveals

The aggregator learns the pooled per-probe mean and SD (round 1), a
per-probe stratified univariate Cox $|Z|$ statistic and the indices of
the top-`r K` screened probes (round 2), and the consensus $z^t$ at
every ADMM iteration. Each is an aggregated population-level statistic
over `r N_total` patients, not patient-level data. None of the per-site
design matrices $X_k$, the per-site $(\mu,\sigma^2)$ contributions, the
per-site $(U,I)$ contributions, or the per-site $(x_k+u_k)$ vectors ever
appear in cleartext anywhere in the protocol. Decryption at every step
is $n$-of-$n$ threshold: no single party — the aggregator included — can
recover any intermediate quantity unilaterally.

## Discussion

1. **CVXR symbolic problems compose with threshold FHE.** The local
   Cox-lasso solve runs in the clear at each site; only the cross-site
   consensus update goes through the encrypted channel. The reader does
   not rewrite their [CVXR][] model for the encrypted setting — the same
   `Problem(Minimize(...))` is used verbatim, and `run_admm` is called
   with `encrypted_consensus` instead of `plain_consensus`.
2. **The encrypted layer is lossless to working precision.** The
   encrypted fit reproduces the plaintext ADMM to
   `r signif(max(abs(z_enc - z_ref)), 2)` and recovers the identical
   active set; the residual gap to the one-shot centralized solve
   (`r signif(max(abs(z_ref - agg_beta)), 2)`) is the ADMM iteration
   budget, not the cryptography.
3. **No single party holds the secret key.** `make_threshold_master()`
   distributes the secret across all sites; encrypted intermediates are
   undecryptable by any one party, and each round's result appears only
   after $n$-of-$n$ partial-decryption fusion.
4. **DPP keeps the inner loop fast.** Each site's [CVXR][] problem is
   built once at setup; ADMM iterations only update the `Parameter`
   values.

## Limitations

- **Honest-but-curious trust.** A site that misreports its local
  $(x_k+u_k)$ can corrupt the consensus; detecting this needs
  commitments / zero-knowledge proofs not implemented here.
- **The aggregator sees the trajectory $\{z^t\}$.** Per-iteration
  consensus values are revealed in plaintext (after fusion) so the loop
  can decide convergence.
- **No output privacy.** The released $\hat\beta = z^\star$ is the same
  coefficient vector as the centralized fit; output-level protection is
  out of scope here and is demonstrated in the companion DP vignette.

[CVXR]: https://cvxr.rbind.io
