---
title: "Consensus ADMM + Output Differential Privacy (Demonstration)"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Consensus ADMM + Output DP (Demonstration)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## Why this vignette exists

This vignette is a **demonstration**, not a recommendation.
`vignette("cvxr-consensus-admm")` shows the lossless consensus-ADMM
protocol over the threshold-FHE channel: bit-identical to the
centralised CVXR fit, no single decrypter, residuals drive
convergence. The aggregator sees the full trajectory $\{z^k\}$ as
plaintext after each iteration's threshold fusion.

A natural question is whether one can compose that lossless
channel with **output differential privacy** to bound the
trajectory leakage formally. The answer in this simulation is: yes,
the cryptographic protocol composes cleanly and the implementation
is one extra `rnorm()` call per site per iteration, but the
accuracy/privacy trade-off does not produce a usable point at
small $\varepsilon$ given the optimiser structure here. This
vignette runs that trade-off end-to-end.

`feedback_he_not_dp` in the package's notes records the framing
explicitly: lossless threshold-FHE is the package's primary path;
DP demonstrations show *what happens if* a user composes output DP
on top, not the recommended deployment for precision-critical
workloads.

## A brief output-DP primer

Given a query $f: \mathcal{D} \to \mathbb{R}^p$, the **Gaussian
mechanism** releases $\tilde f(D) = f(D) + \mathcal{N}(0, \sigma^2 I)$.
Sensitivity-driven $\sigma$ gives single-query
$(\varepsilon, \delta)$ DP. Multi-query composition uses **zCDP**
[Bun & Steinke 2016]: each release is $(\Delta/\sigma)^2/2$-zCDP;
$T$ releases compose linearly to $T \cdot \rho$; convert back to
$(\varepsilon, \delta)$ via $\varepsilon = \rho + 2\sqrt{\rho \log(1/\delta)}$.

## The protocol modification

The setup is identical to `vignette("cvxr-consensus-admm")`. The
only change is **noise injection at each site**: at every ADMM
iteration site $i$ adds an independent draw
$\eta_i \sim \mathcal{N}(0, \sigma^2 N \cdot I)$ to $x_i + u_i$
*before* encrypting. The encrypted noise terms sum under the joint
key, the $1/N$ scaling contracts the variance back to $\sigma^2$
per coordinate, and the recovered $z$ has noise
$\mathcal{N}(0, \sigma^2 I)$.

Distributing the noise across sites (rather than centralising it
at the aggregator) means:

1. No single point ever holds the noiseless value. Even if the
   aggregator is compromised it sees only ciphertexts of *noised*
   contributions until the final fusion.
2. The trust model matches `cox-threshold-dp`: each site is its
   own randomness boundary.

## The stopping rule changes

Residual-based convergence checks are not meaningful under noise:
the residuals never shrink below the per-iteration noise floor.
We replace residual-based stopping with **fixed $T$** equal to
the cleartext convergence iteration count for the chosen $\rho$.
The $\sigma = 0$ row of the table below verifies that the protocol
with the noise mechanism turned off lands on the lossless ADMM
fit at the same fixed $T$.

## Setup

```{r}
suppressPackageStartupMessages({
    library(homomorpheR)
    library(CVXR)
    library(S7)
})

N   <- 3L
p   <- 4L
lam <- 1
```

```{r}
build_local_problem <- function(X_i, y_i, rho_val) {
    x  <- Variable(p)
    zp <- Parameter(p)
    up <- Parameter(p)
    y_signs <- 2 * y_i - 1
    margins <- -y_signs * (X_i %*% x)
    local_loss <- sum(logistic(margins)) +
                  (lam / (2 * N)) * sum_squares(x)
    augmented  <- (rho_val / 2) * sum_squares(x - zp + up)
    prob <- Problem(Minimize(local_loss + augmented))
    value(zp) <- rep(0, p); value(up) <- rep(0, p)
    list(prob = prob, x = x, zp = zp, up = up)
}

ConsensusSite <- new_class("ConsensusSite",
    properties = list(
        name  = class_character,
        n     = class_integer,
        state = class_any))

make_consensus_site <- function(name, X_i, y_i, rho_val) {
    st        <- new.env(parent = emptyenv())
    st$X      <- X_i
    st$y      <- y_i
    built     <- build_local_problem(X_i, y_i, rho_val)
    st$prob   <- built$prob
    st$x_var  <- built$x
    st$zp     <- built$zp
    st$up     <- built$up
    st$x_curr <- rep(0, ncol(X_i))
    st$u_curr <- rep(0, ncol(X_i))
    ConsensusSite(name = name, n = nrow(X_i), state = st)
}

local_update <- function(site, z_curr) {
    st <- site@state
    value(st$zp) <- z_curr
    value(st$up) <- st$u_curr
    suppressMessages(suppressWarnings(psolve(st$prob, solver = "CLARABEL")))
    if (!status(st$prob) %in% c("optimal", "optimal_inaccurate"))
        stop("Local CVXR solve at ", site@name, " did not reach optimal status.")
    st$x_curr <- as.numeric(value(st$x_var))
    invisible(st$x_curr)
}
```

## Simulated cohort

```{r}
set.seed(20260412)
n_per_site <- c(500L, 1000L, 1500L)
beta_true  <- c(intercept = -0.5, age = 0.4, bmi = -0.3, sex = 0.6)

make_site_data <- function(n) {
    X  <- cbind(1, rnorm(n), rnorm(n), rbinom(n, 1, 0.5))
    pr <- plogis(as.numeric(X %*% beta_true))
    y  <- as.integer(runif(n) < pr)
    list(X = X, y = y)
}
site_data <- lapply(n_per_site, make_site_data)
```

## Cleartext $\rho$ sweep

Pick $\rho$ programmatically — same idiom as the lossless ADMM
vignette.

```{r}
tol      <- 1e-3
max_iter <- 60L

sweep_one_rho <- function(rho_val) {
    built <- lapply(site_data,
                    function(s) build_local_problem(s$X, s$y, rho_val))
    x_curr <- u_curr <- replicate(N, rep(0, p), simplify = FALSE)
    z      <- rep(0, p)
    k_conv <- NA_integer_
    for (k in seq_len(max_iter)) {
        for (i in seq_len(N)) {
            value(built[[i]]$zp) <- z
            value(built[[i]]$up) <- u_curr[[i]]
            suppressMessages(suppressWarnings(
                psolve(built[[i]]$prob, solver = "CLARABEL")))
            x_curr[[i]] <- as.numeric(value(built[[i]]$x))
        }
        z_prev <- z
        z <- Reduce(`+`, Map(`+`, x_curr, u_curr)) / N
        for (i in seq_len(N)) u_curr[[i]] <- u_curr[[i]] + x_curr[[i]] - z
        pri <- sqrt(sum(vapply(seq_len(N),
            function(i) sum((x_curr[[i]] - z)^2), 0)) / N)
        dua <- rho_val * sqrt(sum((z - z_prev)^2))
        if (pri < tol && dua < tol) { k_conv <- k; break }
    }
    data.frame(rho = rho_val,
               iters = if (is.na(k_conv)) max_iter else k_conv,
               converged = !is.na(k_conv))
}

rho_grid  <- c(10, 20, 50, 100, 500)
rho_sweep <- do.call(rbind, lapply(rho_grid, sweep_one_rho))
knitr::kable(rho_sweep, caption = "Cleartext consensus-ADMM convergence")

converged_rows <- rho_sweep[rho_sweep$converged, ]
if (nrow(converged_rows) == 0L)
    stop("No rho in the grid converged within max_iter.")

rho_chosen <- converged_rows$rho[which.min(converged_rows$iters)]
T_fixed    <- converged_rows$iters[converged_rows$rho == rho_chosen]
cat(sprintf("Chosen rho = %g (T = %d iterations cleartext).\n",
            rho_chosen, T_fixed))
```

The DP-ADMM loop below runs for exactly $T = `r T_fixed`$
iterations regardless of residuals.

## Threshold-FHE setup

```{r}
cc <- openfhe::fhe_context("CKKS",
                           multiplicative_depth = 1L,
                           scaling_mod_size     = 59L,
                           first_mod_size       = 60L,
                           batch_size           = 8L,
                           features             = c(openfhe::Feature$MULTIPARTY))
```

The DP version of the consensus step. The only change from the
lossless ADMM vignette's `encrypted_consensus()` is the
`rnorm(p, ..., sd = sigma * sqrt(Nv))` term inside the per-site
loop:

```{r}
encrypted_consensus_dp <- function(threshold_master, sites, sigma) {
    Nv  <- length(sites)
    cts <- vector("list", Nv)
    for (i in seq_along(sites)) {
        st  <- sites[[i]]@state
        val <- st$x_curr + st$u_curr +
               rnorm(p, mean = 0, sd = sigma * sqrt(Nv))
        cts[[i]] <- master_encrypt(threshold_master, val)
    }
    ct_sum <- Reduce(`+`, cts)
    ct_avg <- ct_sum * (1 / Nv)
    master_decrypt(threshold_master, ct_avg, len = p)
}
```

## The DP-ADMM loop

```{r}
run_dp_admm <- function(sigma, T_iter = T_fixed, seed = NULL) {
    if (!is.null(seed)) set.seed(seed)
    master <- make_threshold_master("Aggregator",
                                    crypto_context = cc,
                                    n_sites        = N)
    sites <- list(
        make_consensus_site("Site 1", site_data[[1]]$X, site_data[[1]]$y, rho_chosen),
        make_consensus_site("Site 2", site_data[[2]]$X, site_data[[2]]$y, rho_chosen),
        make_consensus_site("Site 3", site_data[[3]]$X, site_data[[3]]$y, rho_chosen))

    z_curr <- rep(0, p)
    z_hist <- matrix(NA_real_, nrow = T_iter, ncol = p,
                     dimnames = list(NULL, names(beta_true)))
    for (k in seq_len(T_iter)) {
        for (s in sites) local_update(s, z_curr)
        z_curr <- encrypted_consensus_dp(master, sites, sigma)
        for (s in sites) {
            s@state$u_curr <- s@state$u_curr + (s@state$x_curr - z_curr)
        }
        z_hist[k, ] <- z_curr
    }
    list(z = z_curr, z_hist = z_hist)
}
```

## Centralised CVXR oracle

```{r}
X_pooled  <- do.call(rbind, lapply(site_data, `[[`, "X"))
y_pooled  <- unlist(lapply(site_data, `[[`, "y"))
beta_var  <- Variable(p)
y_signs_p <- 2 * y_pooled - 1
margins_p <- -y_signs_p * (X_pooled %*% beta_var)
suppressMessages(suppressWarnings(
    psolve(Problem(Minimize(sum(logistic(margins_p)) +
                            (lam / 2) * sum_squares(beta_var))),
           solver = "CLARABEL")))
beta_oracle <- as.numeric(value(beta_var))
names(beta_oracle) <- names(beta_true)
```

## The $\sigma$ sweep

Six $\sigma$ values from zero to one. The $\sigma = 0$ row is the
sanity check that the DP mechanism is a no-op when off.

```{r}
sigma_grid    <- c(0, 1e-4, 1e-3, 1e-2, 1e-1, 1)
sweep_results <- vector("list", length(sigma_grid))
for (j in seq_along(sigma_grid)) {
    sweep_results[[j]] <- run_dp_admm(sigma = sigma_grid[j], seed = 100L + j)
}
names(sweep_results) <- sprintf("sigma=%.0e", sigma_grid)
```

## $\sigma = 0$ sanity check

```{r}
clean_dev <- max(abs(sweep_results[[1]]$z - beta_oracle))
agree_tol <- 10 * tol
if (clean_dev > agree_tol)
    stop("DP-ADMM at sigma = 0 disagrees with the centralised oracle.")
cat(sprintf("Sigma = 0 max coefficient deviation: %.2e (tol %.0e)\n",
            clean_dev, agree_tol))
```

## Summary table

```{r}
summary_df <- do.call(rbind, lapply(seq_along(sigma_grid), function(j) {
    z <- sweep_results[[j]]$z
    data.frame(sigma     = sigma_grid[j],
               intercept = z[1],
               age       = z[2],
               bmi       = z[3],
               sex       = z[4],
               max_dev   = max(abs(z - beta_oracle)))
}))
oracle_row <- data.frame(sigma = NA, intercept = beta_oracle[1],
                         age = beta_oracle[2], bmi = beta_oracle[3],
                         sex = beta_oracle[4], max_dev = 0)
summary_table <- rbind(summary_df, oracle_row)
rownames(summary_table) <- c(sprintf("sigma=%g", sigma_grid), "oracle")
knitr::kable(summary_table, digits = 6,
             caption = "DP-ADMM coefficients vs centralised CVXR")
```

The deviation grows roughly linearly with $\sigma$ in the
small-$\sigma$ regime and then leaves the linear band — at
$\sigma = 1$ the recovered coefficients drift far from the
centralised oracle. The protocol is faithfully running the same
algorithm at every $\sigma$; the deterioration is what the noise
mechanism does to the optimisation, not what the cryptographic
channel does.

## Privacy budget

Per-iteration zCDP: $\rho_{\text{iter}} = (\Delta/\sigma)^2/2$ per
*site* per coordinate. Across $T = `r T_fixed`$ iterations and
$N = `r N`$ sites, total $\rho = T \cdot N \cdot (\Delta/\sigma)^2/2$.
Convert to $(\varepsilon, \delta)$ via the standard formula.

```{r}
zcdp_to_eps <- function(rho, delta = 1e-5) rho + 2 * sqrt(rho * log(1 / delta))

budget <- data.frame(sigma = sigma_grid[sigma_grid > 0])
budget$rho_total                  <- T_fixed * N * (1 / budget$sigma)^2 / 2
budget$epsilon_at_delta_1e_minus_5 <- zcdp_to_eps(budget$rho_total)
knitr::kable(budget, digits = 4,
             caption = "zCDP composition; sensitivity Delta = 1, delta = 1e-5")
```

Read `epsilon_at_delta_1e_minus_5`. The smallest $\varepsilon$
attained — at $\sigma = 1$ where the fit is already broken — is
still in the dozens. Whether large $\varepsilon$ is acceptable is a
use-case decision; we report the numbers and stop.

## What this demonstrates

1. **The cryptographic protocol composes cleanly with output
   DP.** At $\sigma = 0$ the DP-ADMM protocol reproduces the
   lossless ADMM fit. The DP layer is one `rnorm()` call per site
   per iteration and zero new homomorpheR machinery.
2. **The accuracy/privacy trade-off is reported end-to-end.** Six
   $\sigma$ values, one fixed-$T$ DP-ADMM run each, summary table
   showing the recovered coefficients and the deviation from the
   centralised oracle.
3. **At noise scales the optimiser tolerates, $\varepsilon$ is
   large.** zCDP composition over $T \cdot N$ Gaussian releases
   inflates the budget; tightening it requires either fewer
   iterations, tighter sensitivity, or subsampling amplification —
   all out of scope for this demonstration.

## What this does not address

- **Tight sensitivity bounds** for the consensus update. $\Delta = 1$
  is a placeholder.
- **Adaptive noise** schedules across iterations.
- **Subsampling amplification** of the DP accountant.
- **Malicious-site protection.** A hostile site can corrupt either
  its $(x_i + u_i)$ or its noise draw; out of scope.

Readers who need the lossless fit should use
`vignette("cvxr-consensus-admm")` and accept the trajectory
exposure that comes with it.

## References

- Bun & Steinke (2016). *Concentrated Differential Privacy.* TCC.
- Cyffers, Bellet & Upadhyay (2023). *Muffliato:
  Peer-to-Peer Privacy Amplification for Decentralised
  Optimisation and Averaging.* The DP-ADMM analytic companion to
  this empirical demonstration.
