---
title: "Encrypted Logistic Regression Prediction"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Encrypted Logistic Regression Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## Overview

The `secure-inference` vignette showed a *linear* model evaluated
on encrypted patient data. This vignette goes one step further:
the prediction needs the logistic sigmoid

$$\sigma(\eta) = \frac{1}{1 + e^{-\eta}},$$

a non-polynomial function that on its face cannot be evaluated
under encryption — both BFV/BGV and CKKS only support polynomial
arithmetic. Using a Chebyshev polynomial approximation of the
sigmoid, however, we can evaluate the entire logistic
prediction (linear predictor *and* sigmoid) under encryption with
only addition and multiplication.

## The setting

A hospital holds patient data. A researcher holds a trained
logistic-regression model. The researcher wants to score patients
without seeing their data; the hospital wants predictions without
seeing the model coefficients. The same two-party shape as
`secure-inference`, with a non-linear scorer.

## Step 1: train in the clear

```{r}
set.seed(123)
n <- 500
age       <- rnorm(n, 55, 10)
biomarker <- rnorm(n, 0, 1)
prob      <- plogis(-2 + 0.03 * age + 0.8 * biomarker)
outcome   <- rbinom(n, 1, prob)

model <- glm(outcome ~ age + biomarker, family = binomial)
beta  <- coef(model)
cat("Coefficients (intercept, age, biomarker):", round(beta, 4), "\n")
```

## Step 2: encrypt patient data

CKKS needs enough multiplicative depth for the Chebyshev
polynomial. We use depth 8 and enable `Feature$ADVANCEDSHE` for
the polynomial-evaluation primitives.

```{r}
library(openfhe.R)

cc <- fhe_context("CKKS",
                  multiplicative_depth = 8L,
                  scaling_mod_size     = 50L,
                  batch_size           = 16L,
                  features             = c(Feature$ADVANCEDSHE))
keys <- key_gen(cc, eval_mult = TRUE)

new_age <- c(45, 52, 60, 38, 70, 55, 48, 63,
             41, 57, 66, 44, 72, 50, 59, 35)
new_bm  <- c(-0.5,  0.3, 1.2, -1.0, 0.8, 0.1, -0.3, 1.5,
             -0.8,  0.6, 0.9, -0.4, 1.1, 0.0,  0.7, -1.2)

ct_age <- encrypt(keys@public, make_ckks_packed_plaintext(cc, new_age), cc = cc)
ct_bm  <- encrypt(keys@public, make_ckks_packed_plaintext(cc, new_bm),  cc = cc)
```

## Step 3: evaluate the linear predictor (encrypted)

$\eta = \beta_0 + \beta_1\, \text{age} + \beta_2\, \text{biomarker}$,
all arithmetic on encrypted data:

```{r}
ct_eta <- ct_age * beta[2]
ct_eta <- ct_eta + ct_bm * beta[3]
ct_eta <- ct_eta + beta[1]
```

## Step 4: apply the sigmoid (encrypted)

`openfhe.R` exposes Chebyshev approximations of common transcendental
functions, including the logistic sigmoid. The interval $[a, b]$
must cover the realisable range of $\eta$; degree 16 gives good
precision with depth modest enough to fit the budget we declared
above.

```{r}
ct_prob <- eval_logistic(ct_eta, a = -4, b = 4, degree = 16)
```

## Step 5: decrypt predictions

```{r}
result <- decrypt(ct_prob, keys@secret, cc = cc)
set_length(result, 16L)
encrypted_probs <- get_real_packed_value(result)[1:16]

cleartext_probs <- plogis(beta[1] + beta[2] * new_age + beta[3] * new_bm)
max_err <- max(abs(encrypted_probs - cleartext_probs))
cat(sprintf("Max absolute error vs cleartext sigmoid: %.2e\n", max_err))
```

## What this demonstrates

1. **Model coefficients applied to encrypted data.** The
   hospital never sees the model; the researcher never sees the
   patient data. Same transport guarantee as `secure-inference`.
2. **The sigmoid evaluated entirely under encryption.** No
   intermediate decryption, no protocol back-and-forth — the
   whole logistic prediction completes inside the ciphertext
   domain.
3. **CKKS precision is more than enough for medical
   prediction.** The maximum absolute error against the
   cleartext sigmoid is `r signif(max_err, 2)`, well below any
   threshold that would matter for a clinical decision.

## Connection to homomorpheR's Paillier vignettes

The Paillier vignettes (`vignette("homomorphing")`,
`vignette("DHCox")`, `vignette("DHCoxNCP")`) demonstrate
distributed Cox regression and MLE under additive Paillier
encryption. That approach requires:

- sending encrypted *sufficient statistics* (sums) between sites
  rather than evaluating the full likelihood under encryption,
- iterative protocols where the master site decrypts intermediate
  aggregates,
- careful real-number encoding via integer / fractional split.

With CKKS via `openfhe.R`:

- real numbers are encrypted directly — no manual scaling,
- multiplication works — enabling polynomial and sigmoid
  evaluation directly under encryption,
- the full computation can happen in the encrypted domain.

## Limitations

- **Multiplicative depth**: each multiplication consumes one
  level of the ciphertext. Deeper computations require larger
  parameters and more memory.
- **Approximation error**: CKKS is approximate. For deployment,
  validate that the precision is sufficient for your use case.
- **Performance**: encrypted computation is orders of magnitude
  slower than cleartext.
- **Threat model**: this transport-layer demonstration shares the
  model-extraction caveat from `vignette("secure-inference")`.
  The hospital holding the secret key can in principle recover
  the coefficients by issuing carefully chosen queries. The
  Chebyshev sigmoid makes the recovery somewhat less trivial
  than the linear case, but does not eliminate the issue. Real
  deployments still need defences (output DP, query budgets,
  threshold FHE) on top of the FHE transport.

```{r echo=FALSE}
## Detach openfhe.R so its `encrypt`/`decrypt` generics do not bleed
## into other vignettes built later in the same R session.
detach("package:openfhe.R", unload = FALSE, force = TRUE)
```
