---
title: "Introduction to homomorpheR"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to homomorpheR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## The problem this package addresses

A recurring situation in biomedical research: several sites hold
patient records that cannot leave the institutions that collected
them, and you would like to fit a single model to the pooled data.

For a large class of models this is less hopeless than it sounds.
The log-likelihood, the score, and the information matrix are all
*sums over observations*, so they are also sums over sites. An
optimizer fitting such a model never needs an individual record — it
needs the total of the per-site contributions at whatever parameter
value it is currently considering. If the sites can compute that
total without revealing their individual contributions, the fit
proceeds exactly as it would on pooled data.

Homomorphic encryption is what makes the total computable. It is a
form of encryption under which certain arithmetic still works: you
can add two encrypted numbers and get an encryption of their sum,
without decrypting either one. The party doing the adding learns
nothing.

`homomorpheR` supplies the coordination layer — the sites, the
aggregator, and the protocol connecting them — on top of the
encryption provided by the `openfhe.R` package.

The point worth keeping in view is that **the statistical machinery
does not change**. The vignettes here fit Cox models with
`survival`, maximum likelihood with `stats4::mle()`, and penalized
regression with `CVXR`, all unmodified. Only the function that
returns the summary at each iteration is different.

## Terms you will meet

<!-- lint-vocabulary: off -->
<!-- Naming each cryptographic term once is the point of this
     section; the rest of the package avoids them. -->


Encryption brings its own vocabulary, and a few of those words
appear in the function names and in the vignettes that follow. None
of them require a cryptography background. Where a word has a
standard cryptographic name that you will meet in the `openfhe.R`
documentation or in the literature, it is given in parentheses on
first mention and then set aside in favor of the plainer term.

- **Encrypted value** (*ciphertext*). The result of encrypting a
  number or a vector of numbers. Encrypted values can be added to
  each other, and multiplied, without being decrypted first. Every
  encrypted quantity in these vignettes is one of these.

- **Cleartext value** (*plaintext*). The ordinary, unencrypted
  number — what you started with, and what you get back after
  decryption. Note that in the encryption API this word also names
  an intermediate *encoded* form, produced just before encryption;
  that is why function names such as
  `make_ckks_packed_plaintext()` contain it.

- **Slot.** An encrypted value is not a single number but a vector
  with a fixed number of positions, typically thousands. Each
  position is a slot. One arithmetic operation acts on every slot
  at once, which is what makes encrypted vector arithmetic
  affordable. You generally use the first few slots and ignore the
  rest.

- **Site.** A party holding data that must not leave its
  institution. In this package a site is an object carrying its
  local data and a function that returns the site's summary at a
  given parameter value. Sites never share their data, and never
  see each other's contributions.

- **Aggregator** (*computing party*, *evaluator*). The party that
  collects the encrypted per-site summaries, adds them together
  while they are still encrypted, and obtains the total. It sees
  only encrypted quantities in transit.

- **Public and secret key.** The standard public-key pair.
  Encryption uses the public key, which everyone may hold;
  decryption uses the secret key. Who holds the secret key is the
  central design question in these protocols, and the answer
  distinguishes the two kinds of aggregator below.

- **Evaluation keys.** Additional keys that *authorize* particular
  operations on encrypted values — multiplying two of them,
  summing across slots, rotating a vector. They are generated
  alongside the key pair and travel with the public key. They
  permit computation; they do not permit decryption.

- **Threshold keys.** A key arrangement in which no single party
  holds the secret key. Each party holds only a share, the public
  key is built jointly from all of them, and decryption requires
  every party to contribute. No proper subset can decrypt
  anything. This removes the residual trust that an ordinary key
  pair leaves with whoever holds the secret key.

- **Precision budget** (*multiplicative depth*, *levels*). Real-valued
  encrypted arithmetic is approximate, and each multiplication
  consumes part of a finite budget of precision fixed when the
  encryption parameters are chosen. Additions are essentially
  free; multiplications are not. When the budget is exhausted,
  further multiplications return values that are simply wrong. The
  aggregation patterns in these vignettes are deliberately shallow
  — mostly sums — so a small budget suffices.

- **Scheme.** The particular encryption construction in use. Two
  appear here. **CKKS** handles real numbers approximately and is
  what nearly every vignette uses, since statistical quantities are
  real-valued. **BFV** handles integers exactly, and is used where
  the answer is a count and no approximation is acceptable. The
  choice is an implementation detail of each vignette, stated in
  its opening paragraph.

That is the whole vocabulary. The vignettes that follow describe
the protocols using only these terms.

<!-- lint-vocabulary: on -->

## How the pieces fit together

Three objects carry the protocol.

A **site** is built with `make_worker()`. It holds the site's data
and a function `contribution_fn(data, theta)` returning that site's
summary at the parameter value `theta`. If a parameter value breaks
the computation — an extreme value that the site's solver cannot
handle — the function returns `NA`, and that signal propagates back
to the optimizer rather than corrupting the fit.

`make_worker()` builds a `LocalSite` — a site whose records are in
this [R]{.proglang} session. The master reaches every site through
one generic, `contribute(site, theta)`, and what comes back is
**already encrypted**: the site encrypts under the public key the
master broadcast, so no individual site's cleartext contribution ever
reaches the aggregator. That is the property the whole protocol rests
on, and it is why `contribution_fn` returns a plain number while
`contribute()` — not you — does the encrypting.

When the records are not in this session, subclass `RemoteSite` and
give it a `contribute()` method. The package deliberately ships no
implementation: transports differ too much, and a cryptography
package has no business carrying an HTTP client. The extension is
small:

```{r}
#| eval: false
HttpSite <- S7::new_class("HttpSite", parent = RemoteSite,
                          properties = list(url = S7::class_character))

S7::method(contribute, HttpSite) <- function(site, theta) {
    ## The far side holds the public key and encrypts before replying,
    ## so the wire carries encrypted values -- never a bare number.
    fetch_encrypted_contribution(site@url, theta)
}
```

Read `?RemoteSite` before writing one. The contract has a few sharp
edges, the sharpest being that **`NA` and "unreachable" are different
events**. `NA` means *this* `theta` broke the site's solver, and the
optimizer responds sensibly by trying a different parameter. A
network or timeout failure is not that, and backing off to another
`theta` does nothing about it — signal `site_unavailable()` instead,
which aborts the round. Returning `NA` for an offline host, or
quietly dropping the site, changes the set of sites being summed
over between iterations and the fit converges to something that is
not the estimand, with no error raised anywhere.

An **aggregator** is built with either `make_ckks_master()` or
`make_threshold_master()`. The two differ in exactly one respect,
and it is the respect that matters:

- `make_ckks_master()` creates an aggregator holding an ordinary
  key pair. It is appropriate when one party is permitted to hold
  the secret key.
- `make_threshold_master()` takes the sites and runs threshold key
  generation through them: each site generates its own share and
  keeps it, so no single party — including the aggregator itself —
  can decrypt alone. The aggregator it returns holds only public
  material, and is already wired to those sites.

Because both are the same kind of object underneath, the protocol
body is identical for either. Choosing a trust model means choosing
a constructor, not rewriting the analysis. The one setup difference
follows from the cryptography rather than the API: a joint public
key is built *from* the sites, so under threshold keys the sites are
created first and handed to the constructor, instead of being wired
to an aggregator afterwards with `set_workers()`.

**`master_aggregate(master, theta)`** runs one round. The
aggregator sends `theta` to every site; each site computes its
local summary and encrypts it; the aggregator adds the encrypted
summaries together and decrypts only the total. An optimizer calls
this once per iteration, and the fit proceeds.

The topology is a flat fan-out and fan-in, which is how federated
analysis frameworks such as `distcomp` and DataSHIELD are actually
deployed.

### Sites that are not in this R session

Every vignette here runs all parties inside one R session, so that
each is reproducible by running it. That is a simulation of the
deployment, not the deployment itself, and it is worth being
explicit about what changes when the sites are genuinely remote.

The structure does not change. You subclass `RemoteSite` and give
it whatever properties your transport needs — an endpoint, a
credential, a connection object. The object you hold in the
aggregator's session is a *local handle* for a remote party, not
the party itself. Nothing in `master_aggregate()` needs to know the
difference.

Two methods carry the deployment, and they correspond to the only
two moments anything passes between the parties:

- `set_public_params(site, params)` — setup. Send the public
  context and key to the endpoint and have it keep them. Only
  public material travels.
- `contribute(site, theta)` — one round. Send `theta`, and get back
  a value the far end has **already encrypted**.

The ordering in the second one is the whole point: the summary is
encrypted before it leaves the site, so no individual contribution
ever exists in the clear outside the site that produced it.

The package refuses to fake either step for you. The base
`RemoteSite` has no `set_public_params()` method, so wiring a
subclass that has not implemented one fails immediately rather than
leaving a proxy that looks configured while the endpoint was never
told anything. `keygen_round()` and `partial_decrypt()` refuse in
the same way, since performing a remote party's secret-key
operation in this process is precisely what threshold keys exist to
prevent. And `master_aggregate()` checks that what came back is an
encrypted value under this protocol's key before adding it to a
total — a site that replied with a plain number would otherwise
have its answer summed in silently, and the round would report the
correct result having been handed the one quantity the protocol
hides.

What the package does not supply is everything else a deployment
needs: transport, authentication, key storage at the endpoint, retry
and timeout policy, and any defense against a party that actively
deviates rather than merely observes. `?RemoteSite` states the full
contract.

## What "the right answer" means

*Precision* is the companion to this page. It sets out what an
encrypted result being correct actually means — exact for integer
counting, approximate within a measurable bound for real-valued
arithmetic, and merely statistical when an optimizer is involved — and
which comparisons are not meaningful at all. Worth reading before
interpreting any number in the vignettes below.

## Which vignette to read

**Queries and aggregation.** Counting across sites without
revealing who contributed what. These are the simplest complete
protocols in the package and the best place to begin: the
statistical content is a sum, so nothing distracts from the
mechanics.

- *Privacy-Preserving Count Aggregation* — a single encrypted
  total, with one party holding the secret key.
- *Distributed Query Count with Threshold Keys* — the same count
  with no single party able to decrypt.

**Fitting models across sites.** Fitting a model to data you
cannot pool. Each of these wraps an ordinary R fitting routine
that is used unmodified.

- *Distributed Maximum Likelihood Estimation* — the smallest
  complete model fit, and the place to start in this group.
- *Distributed Stratified Cox Regression* — survival analysis
  across sites, using `survival` unchanged.
- *Distributed Cox Regression with Threshold Key Generation* — the
  same fit with no single party able to decrypt.
- *Federated Consensus ADMM with CVXR* — convex optimization
  across sites, using `CVXR` unchanged.
- *Federated Cox-Lasso via Consensus ADMM on DLBCL* — the above at
  realistic scale on gene expression data.

**Prediction and retrieval.** Two parties, one holding a model and
one holding data, neither willing to reveal theirs.

- *Secure Model Inference on Encrypted Data*
- *Encrypted Logistic Regression Prediction*
- *Federated Cosine Similarity with Site-Private Fine-Tuned Models*

**Adding differential privacy.** What changes if you also want to
bound what the sequence of released totals reveals. These are
demonstrations of the composition mechanics, not a recommendation.

- *Threshold Cox + Output Differential Privacy (Demonstration)*
- *Consensus ADMM + Output Differential Privacy (Demonstration)*

For the encryption layer itself — contexts, key generation,
encrypted arithmetic, serialization — see the `openfhe.R` package
and its own vignettes.

## A note on Paillier

Earlier versions of this package implemented the Paillier
cryptosystem natively in R. That code is frozen: it remains
exported for the packages that depend on it, but it is not extended,
and the protocols documented here use the schemes provided through
`openfhe.R` instead. The Paillier vignettes are retained separately
for reference.
