---
title: "Precision"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Precision}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

"The encrypted computation gives the right answer" means different
things in different vignettes, and being clear about which is the
difference between a meaningful check and a misleading one.

```{r}
library(openfhe.R)
set_num_threads(2L)
```

## Exact, for integer schemes

BFV does exact integer arithmetic. A count computed under encryption
is *the same integer* as the count computed in the clear — not a
value near it.

```{r}
cc  <- fhe_context("BFV", plaintext_modulus = 65537L,
                   multiplicative_depth = 1L, batch_size = 8L)
key <- key_gen(cc)

counts <- c(46L, 15L, 52L)
cts    <- lapply(counts, function(n)
  encrypt(key@public, make_packed_plaintext(cc, n), cc = cc))

total <- decrypt(Reduce(`+`, cts), key@secret, cc = cc)
set_length(total, 1L)
get_packed_value(total)[1] == sum(counts)
```

`vignette("privacy-preserving-aggregation")` and
`vignette("query-count-threshold")` assert equality exactly like
this. A tolerance there would be hiding a bug rather than
accommodating one.

## Approximate, for CKKS

CKKS encrypts real numbers and trades exactness for that ability.
Results carry an approximation error that grows with the depth of the
computation and shrinks as the scaling factor widens.

```{r}
ckks_error <- function(depth, scaling_mod_size, values, weights = NULL) {
  cc  <- fhe_context("CKKS", multiplicative_depth = depth,
                     scaling_mod_size = scaling_mod_size,
                     first_mod_size = 60L, batch_size = 8L)
  key <- key_gen(cc, eval_mult = TRUE)

  cts <- lapply(values, function(v)
    encrypt(key@public, make_ckks_packed_plaintext(cc, v), cc = cc))

  if (is.null(weights)) {
    ct       <- Reduce(`+`, cts)
    expected <- Reduce(`+`, values)
  } else {
    ct       <- Reduce(`+`, Map(function(c, w) c * w, cts, weights))
    expected <- Reduce(`+`, Map(function(v, w) v * w, values, weights))
  }

  res <- decrypt(ct, key@secret, cc = cc)
  set_length(res, length(expected))
  got <- get_real_packed_value(res)[seq_along(expected)]

  absolute <- max(abs(got - expected))
  c(absolute = absolute, relative = absolute / max(abs(expected)))
}

## The three settings at a given magnitude are run on the SAME draw, so
## the comparison between them is controlled: only the parameter under
## study changes. Redrawing per row would confound the setting with the
## sample.
settings <- list(
  list(name = "sum",                         depth = 1L, sms = 50L, w = NULL),
  list(name = "sum, wider scale",            depth = 1L, sms = 59L, w = NULL),
  list(name = "weighted sum (one multiply)", depth = 2L, sms = 50L,
       w = list(0.35, -0.20, 0.50)))

set.seed(1)
rows <- do.call(rbind, lapply(c(1, 1e2, 1e4), function(m) {
  values <- replicate(3, rnorm(8, m, m / 10), simplify = FALSE)
  do.call(rbind, lapply(settings, function(s) {
    e <- ckks_error(s$depth, s$sms, values, s$w)
    data.frame(computation = s$name, magnitude = m,
               depth = s$depth, scaling_mod_size = s$sms,
               absolute = e[["absolute"]], relative = e[["relative"]])
  }))
}))
```

```{r echo=FALSE}
## Scientific notation explicitly: these span ten orders of magnitude,
## and any fixed number of decimal places renders most of them as zero.
shown <- transform(rows,
                   absolute = sprintf("%.2e", absolute),
                   relative = sprintf("%.2e", relative))
knitr::kable(shown,
             caption = "CKKS error against the same computation in the clear.")
```

```{r echo=FALSE}
## Every ratio quoted below is computed from `rows`, never typed. A
## hand-written "about two orders" beside a table that says otherwise
## is the drift this package works to avoid.
pick  <- function(comp, m, col = "absolute")
  rows[[col]][rows$computation == comp & rows$magnitude == m]
gain  <- function(m) pick("sum", m) / pick("sum, wider scale", m)
mult  <- function(m, col) pick("weighted sum (one multiply)", m, col) /
                          pick("sum", m, col)
fmt   <- function(x) signif(x, 2)
```

Three things to read off that table.

**State tolerances relatively, not absolutely.** For the plain sum the
absolute error rises from `r sprintf("%.1e", pick("sum", 1))` at
magnitude 1 to `r sprintf("%.1e", pick("sum", 1e4))` at magnitude
$10^4$, a factor of about `r fmt(pick("sum", 1e4) / pick("sum", 1))`,
while the relative error *falls* — from
`r sprintf("%.1e", pick("sum", 1, "relative"))` to
`r sprintf("%.1e", pick("sum", 1e4, "relative"))`. At magnitude 1 a
fixed noise floor is large compared to the answer; by magnitude $10^4$
it is negligible against it. A tolerance calibrated on standardized
covariates is therefore far too tight for a log-likelihood in the
hundreds, which is why the Cox vignettes raise `scaling_mod_size`
above the default rather than loosening a comparison.

**Widening the scaling factor helps only where that floor binds.** At
magnitude 1 it improves the absolute error by a factor of
`r fmt(gain(1))`. At magnitude $10^4$ the same change gives a factor of
`r fmt(gain(1e4))` — that is, nothing, and the two settings land within
a small multiple of each other in either direction. Once the floor is
no longer what limits the answer, a wider scale stops buying accuracy
while still consuming modulus budget. It is a targeted fix for
small-magnitude work, not a general accuracy dial.

**Each multiplication costs precision as well as budget.** The depth-2
row is worse than the depth-1 sum at every magnitude: by a factor of
`r fmt(mult(1, "absolute"))` in absolute terms at magnitude 1, and
`r fmt(mult(1e4, "relative"))` in relative terms at magnitude $10^4$.
The budget is consumed whether or not the extra precision is missed.

These are comparisons of the encrypted result against *the same
computation performed in the clear on the same data*. That is the
actual cryptographic claim, and it is the tightest bound any vignette
here can make.

## Statistical, for fitted estimates

When an optimizer runs through the encrypted channel, what comes back
is compared against a centralized fit. The relevant tolerance is then
not CKKS precision but a *statistical* one: two optimizers with
different stopping rules do not halt at the same point even on
identical data, and a difference between them is meaningless if it is
small relative to the standard error.

`vignette("cox")` states its agreement in standard-error units for
exactly that reason. A difference of a thousandth of one standard
error is invisible in any reported result, and quoting it as a bare
number would invite the reader to compare it against a CKKS error
bound — a category mistake.

## What cannot be compared

**Differentially private runs.** `vignette("cox-threshold-dp")` and
`vignette("cvxr-consensus-admm-dp")` draw fresh noise on every query,
so two runs of the same code on the same data disagree by
construction. Only distributional behavior is comparable: how the
error grows with $\sigma$, and which optimizers degrade first.

**Iteration counts.** A loop that stops when a residual first falls
below a threshold is reporting where a continuous quantity crossed a
line. CKKS noise alone can move it by one, and redrawing the data
moves it further. It is a property of the sample, not of the method.

## Where the data comes from

**Measured data ships with the package.** The DLBCL cohort is real,
so every vignette that uses it reads the same values each time and
its numbers are reproducible in the strict sense.

**Simulated data is drawn fresh.** The other vignettes generate their
cohorts on each run. The claim they make is that the encrypted
protocol reproduces the cleartext answer *on whatever data it was
given*, which is a stronger claim than reproducing one fixed dataset.

**Expensive fits are precomputed.** `data(cvxr_consensus)` carries the
~150-iteration ADMM results that `vignette("cvxr-cox-lasso-dlbcl")`
would otherwise recompute on every build. The script that produced it
ships with the package, so the artifact is regenerable rather than
merely asserted.

## Categorical order is protocol semantics

Sites are visited in a fixed order, and the first is the lead
decryptor in the threshold ceremony. In the Cox vignettes the sites
come from splitting the cohort on a factor, so that factor's *level*
order determines which site leads:

```{r}
data(DLBCL, package = "homomorpheR")
levels(DLBCL$Subgroup)
```

That is not alphabetical:

```{r}
sort(unique(as.character(DLBCL$Subgroup)))
```

`split()` orders its output by factor levels, so the protocol visits
GCB first. Had `Subgroup` been stored as a character vector, `split()`
would have sorted it and put ABC first instead — silently swapping
which site holds the lead decryption role. The column is a factor with
a deliberate level order for that reason, and it is worth knowing
before anyone "tidies" it.

```{r echo=FALSE}
## Detach openfhe.R so its encrypt/decrypt generics do not bleed into
## other vignettes built later in the same R session.
detach("package:openfhe.R", unload = FALSE, force = TRUE)
```
