---
title: "Precision"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Precision}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```
## Introduction

Some computations on encrypted data are exact while others are
approximate. We describe the differences below.

```{r}
library(openfhe.R)
```

## Exact, for integer schemes

BFV does exact integer arithmetic. A count computed under encryption
is *the same integer* as the count computed in the clear — not a
value near it.

```{r}
cc  <- fhe_context("BFV", plaintext_modulus = 65537L,
                   multiplicative_depth = 1L, batch_size = 8L)
key <- key_gen(cc)

counts <- c(46L, 15L, 52L)
cts    <- lapply(counts, function(n)
  encrypt(key@public, make_packed_plaintext(cc, n), cc = cc))

total <- decrypt(Reduce(`+`, cts), key@secret, cc = cc)
set_length(total, 1L)
get_packed_value(total)[1] == sum(counts)
```

`vignette("privacy-preserving-aggregation")` and
`vignette("query-count-threshold")` assert equality exactly like
this. A tolerance there would be hiding a bug rather than
accommodating one.

## Approximate, for CKKS

CKKS encrypts real numbers and trades exactness for that ability.
Results carry an approximation error that grows with the depth of the
computation and shrinks as the scaling factor widens.

```{r}
ckks_error <- function(depth, scaling_mod_size, values, weights = NULL) {
  cc  <- fhe_context("CKKS", multiplicative_depth = depth,
                     scaling_mod_size = scaling_mod_size,
                     first_mod_size = 60L, batch_size = 8L)
  key <- key_gen(cc, eval_mult = TRUE)

  cts <- lapply(values, function(v)
    encrypt(key@public, make_ckks_packed_plaintext(cc, v), cc = cc))

  if (is.null(weights)) {
    ct       <- Reduce(`+`, cts)
    expected <- Reduce(`+`, values)
  } else {
    ct       <- Reduce(`+`, Map(function(c, w) c * w, cts, weights))
    expected <- Reduce(`+`, Map(function(v, w) v * w, values, weights))
  }

  res <- decrypt(ct, key@secret, cc = cc)
  set_length(res, length(expected))
  got <- get_real_packed_value(res)[seq_along(expected)]

  absolute <- max(abs(got - expected))
  c(absolute = absolute, relative = absolute / max(abs(expected)))
}

## The three settings at a given magnitude are run on the SAME draw, so
## the comparison between them is controlled: only the parameter under
## study changes. Redrawing per row would confound the setting with the
## sample.
settings <- list(
  list(name = "sum",                         depth = 1L, sms = 50L, w = NULL),
  list(name = "sum, wider scale",            depth = 1L, sms = 59L, w = NULL),
  list(name = "weighted sum (one multiply)", depth = 2L, sms = 50L,
       w = list(0.35, -0.20, 0.50)))

set.seed(1)
rows <- do.call(rbind, lapply(c(1, 1e2, 1e4), function(m) {
  values <- replicate(3, rnorm(8, m, m / 10), simplify = FALSE)
  do.call(rbind, lapply(settings, function(s) {
    e <- ckks_error(s$depth, s$sms, values, s$w)
    data.frame(computation = s$name, magnitude = m,
               depth = s$depth, scaling_mod_size = s$sms,
               absolute = e[["absolute"]], relative = e[["relative"]])
  }))
}))
```

```{r echo=FALSE}
## Scientific notation explicitly: these span ten orders of magnitude,
## and any fixed number of decimal places renders most of them as zero.
shown <- transform(rows,
                   absolute = sprintf("%.2e", absolute),
                   relative = sprintf("%.2e", relative))
knitr::kable(shown,
             caption = "CKKS error against the same computation in the clear.")
```

```{r echo=FALSE}
## Every ratio quoted below is computed from `rows`, never typed. A
## hand-written "about two orders" beside a table that says otherwise
## is the drift this package works to avoid.
pick  <- function(comp, m, col = "absolute")
  rows[[col]][rows$computation == comp & rows$magnitude == m]
gain  <- function(m) pick("sum", m) / pick("sum, wider scale", m)
mult  <- function(m, col) pick("weighted sum (one multiply)", m, col) /
                          pick("sum", m, col)
fmt   <- function(x) signif(x, 2)
```

Three things to read off that table.

**State tolerances relatively, not absolutely.** For the plain sum the
absolute error rises from `r sprintf("%.1e", pick("sum", 1))` at
magnitude 1 to `r sprintf("%.1e", pick("sum", 1e4))` at magnitude
$10^4$, a factor of about `r fmt(pick("sum", 1e4) / pick("sum", 1))`,
while the relative error *falls* — from
`r sprintf("%.1e", pick("sum", 1, "relative"))` to
`r sprintf("%.1e", pick("sum", 1e4, "relative"))`. At magnitude 1 a
fixed noise floor is large compared to the answer; by magnitude $10^4$
it is negligible against it. A tolerance calibrated on standardized
covariates is therefore far too tight for a log-likelihood in the
hundreds, which is why the Cox vignettes raise `scaling_mod_size`
above the default rather than loosening a comparison.

**Widening the scaling factor helps only where that floor binds.** At
magnitude 1 it improves the absolute error by a factor of
`r fmt(gain(1))`. At magnitude $10^4$ the same change gives a factor of
`r fmt(gain(1e4))` — that is, nothing, and the two settings land within
a small multiple of each other in either direction. Once the floor is
no longer what limits the answer, a wider scale stops buying accuracy
while still consuming modulus budget. It is a targeted fix for
small-magnitude work, not a general accuracy dial.

**Each multiplication costs precision as well as budget.** The depth-2
row is worse than the depth-1 sum at every magnitude: by a factor of
`r fmt(mult(1, "absolute"))` in absolute terms at magnitude 1, and
`r fmt(mult(1e4, "relative"))` in relative terms at magnitude $10^4$.
The budget is consumed whether or not the extra precision is missed.

These are comparisons of the encrypted result against *the same
computation performed in the clear on the same data*. 

