---
title: "Distributed Query Count under Threshold BFV"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Distributed Query Count under Threshold BFV}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## The problem

Several sites each hold a private table of patient records. A
coordinator wants the answer to a single aggregate query — *how many
patients across all sites satisfy some condition?* — without any site
revealing its own records, and without the coordinator learning any
individual site's count. Only the grand total should ever become
visible, and it should be **exact**: a count is an integer, and an
answer of "about 4" is not a count.

We simulate three sites, each with `sex`, `age`, and a biomarker `bm`.

```{r}
set.seed(130)
sample_size <- c(60, 15, 25)
query_data <- local({
    tmp   <- c(0, cumsum(sample_size))
    start <- tmp[1:3] + 1
    end   <- tmp[-1]
    id_list <- Map(seq, from = start, to = end)
    lapply(seq_along(sample_size), function(i) {
        data.frame(
            id  = sprintf("P%4d", id_list[[i]]),
            sex = sample(c("F", "M"), sample_size[i], replace = TRUE),
            age = sample(40:70,      sample_size[i], replace = TRUE),
            bm  = rnorm(sample_size[i]),
            stringsAsFactors = FALSE)
    })
})
```

The query we will run is `age < 50 & sex == "F" & bm < 0.2`.

## The target answer

If the data could be pooled in one place, the query is a single line
of R. We compute it here only to have a reference to check the
distributed protocol against.

```{r}
query <- quote(age < 50 & sex == "F" & bm < 0.2)

pooled <- do.call(rbind, query_data)
cleartext_count <- sum(eval(query, pooled))
cleartext_count
```

The pooled answer is `r cleartext_count`. No site will actually pool
its data; this number is the ground truth the encrypted protocol must
reproduce.

## Why the classic solution needed two non-cooperating parties

The additive structure of the problem — the grand total is the sum of
the per-site counts — is exactly what an *additively homomorphic*
encryption scheme computes. Let $c_i$ be site $i$'s local count and
$E(\cdot)$ the encryption function. Given $E(c_1), E(c_2), E(c_3)$, the
scheme lets anyone compute $E(c_1) + E(c_2) + E(c_3) = E(c_1 + c_2 +
c_3)$ without decrypting the parts.

The historical difficulty was not the arithmetic but the *trust
model*. The Paillier cryptosystem [@Paillier99], which `homomorpheR`
originally used for exactly this kind of aggregation, has a **single
private key**. Whoever holds it can decrypt anything — including a lone
$E(c_i)$. So if the coordinator holds the key, it can read each site's
count individually, defeating the purpose.

The workaround was two **non-cooperating parties**, NCP1 and NCP2, who
do not collude. Each site draws a fresh random $r_i$ and sends the two
additive shares

$$
E(c_i - r_i) \;\longrightarrow\; \text{NCP1},
\qquad
E(c_i + r_i) \;\longrightarrow\; \text{NCP2}.
$$

NCP1 sums its shares to get $E\!\left(\sum_i (c_i - r_i)\right)$ and
NCP2 gets $E\!\left(\sum_i (c_i + r_i)\right)$; neither total means
anything on its own because of the random masks. The coordinator adds
the two encrypted totals, the masks cancel, and it decrypts

$$
\textstyle\sum_i (c_i - r_i) + \sum_i (c_i + r_i)
  \;=\; \sum_i 2c_i \;=\; 2c,
$$

recovering the grand total $c$ after halving. No party ever sees an
individual $c_i$. This works, but it is machinery erected to
compensate for a single-key scheme: two extra parties, a
non-collusion assumption between them, and a masking dance. The full
Paillier implementation is preserved in the package's archived
`QueryNCP` vignette.

## Threshold encryption removes the need for masking

`openfhe.R` supports **threshold** (multiparty) encryption, and it
changes the trust model at the root. There is no single private key.
Key generation is chained across the sites: each site holds only a
secret *share* $sk_i$, and the joint public key $pk_{1..n}$ is built by
passing the running public key from one site to the next. Everything is
encrypted under $pk_{1..n}$, but **decryption requires every site to
contribute a partial decryption** — no one, coordinator included, can
decrypt anything alone.

Under this model a lone $E(c_i)$ is simply not decryptable by any
single party, so the random masks and the two non-cooperating parties
are no longer needed. The coordinator sums the encrypted per-site
counts and asks the sites to jointly decrypt only the total. Individual
counts are protected not because they are hidden behind random noise,
but because the ability to decrypt them does not exist in any one
place.

We use the **BFV** scheme rather than CKKS here. BFV is exact integer
arithmetic: the decrypted total is the integer sum, bit for bit, with
no approximation. CKKS, used elsewhere in this package for real-valued
likelihoods, would return the count as a floating-point value close to
the integer and require rounding. For a count, exactness is the whole
point, so BFV is the right instrument.

## The threshold-BFV implementation

The crypto context is BFV with the `MULTIPARTY` feature enabled. The
plaintext modulus only has to exceed the largest total the query could
return, so the default is comfortable for counts.

```{r}
library(homomorpheR)

cc <- openfhe.R::fhe_context("BFV",
                             plaintext_modulus    = 65537L,
                             multiplicative_depth = 1L,
                             features             = c(openfhe.R::Feature$MULTIPARTY))
```

`make_threshold_master()` runs the chained key generation across the
three sites and returns a master holding the joint public key and the
per-site secret shares. The same master class drives CKKS or BFV — it
reads the scheme back from the context — so the only thing that changed
from the real-valued vignettes is the context above.

```{r}
master <- make_threshold_master("Coordinator",
                                crypto_context = cc,
                                n_sites        = 3)
```

Each site is a worker whose local function evaluates the query against
its own private data and returns a count. The master broadcasts the
query; no site sees another site's data, and the counts it returns are
encrypted before they leave.

```{r}
local_count <- function(data, query) sum(eval(query, data))

workers <- Map(
    function(nm, d) make_worker(nm, data = d, local_fn = local_count),
    c("Site 1", "Site 2", "Site 3"),
    query_data)

set_workers(master, workers)
```

One call runs the protocol: each worker's encrypted count is summed
homomorphically, and the total is jointly decrypted by the three
secret-share holders.

```{r}
encrypted_count <- master_aggregate(master, theta = query)
encrypted_count
```

## The encrypted answer is exact

```{r}
comparison <- data.frame(
    method = c("pooled cleartext", "threshold-BFV distributed"),
    count  = c(cleartext_count, as.integer(encrypted_count)))
knitr::kable(comparison,
             caption = "Distributed encrypted query count vs. the pooled answer")
```

The threshold-BFV protocol returns `r as.integer(encrypted_count)`,
identical to the pooled answer of `r cleartext_count`. Because BFV is
exact integer arithmetic, the two agree with no tolerance to argue
about — the equality is bit-for-bit, not approximate.

```{r}
stopifnot(identical(as.integer(encrypted_count), as.integer(cleartext_count)))
```

## Discussion

- **The trust improvement is structural.** No single party holds a
  decryption key, so no single party — coordinator or site — can
  decrypt an individual site's count. This is a stronger guarantee
  than the Paillier scheme's, and it is achieved without the two
  non-cooperating parties, the non-collusion assumption, or the random
  masking the single-key scheme required.

- **Only the aggregate is ever revealed.** The sites jointly decrypt
  the total and nothing else. Individual counts are never decrypted by
  anyone, because doing so would require all the secret shares to be
  brought together for that specific encrypted value — which the
  protocol never does.

- **BFV gives exact counts.** For integer-valued aggregates —
  counts, sums, contingency-table cells — BFV decrypts to the exact
  integer. The exactness that the original Paillier demonstration
  offered is retained; the trust model is what improved.

- **Honest-but-curious parties.** As with the other distributed
  protocols in this package, sites are assumed to follow the protocol.
  A malicious site could submit a wrong count or a corrupted partial
  decryption; detecting that requires verifiable decryption, which is
  beyond this demonstration.

## References
