---
title: "Secure Model Inference on Encrypted Data"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Secure Model Inference on Encrypted Data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## The use case

A diagnostic lab develops a proprietary scoring model for disease
risk based on a panel of biomarkers. A hospital wants to score its
patients but cannot send raw patient data to the lab in cleartext.
The lab does not want to send its model coefficients to the
hospital in cleartext either.

With FHE the hospital encrypts patient biomarkers, sends the
encrypted values to the lab, the lab applies its weighted-sum
scoring to them while they stay encrypted, and the hospital
decrypts the scores.
The biomarker values never appear in cleartext on the lab's
machine, and the lab's coefficients are never sent to the
hospital. These are the *transport* guarantees and they are what
the pipeline below illustrates. They are not the whole story of
deploying a model-as-a-service — see *Threat model: model
extraction from the hospital side* below.

This vignette is the two-party companion to the multi-site
master/worker pattern (`vignette("mle")`, `vignette("cox")`).
Unlike those, there is no aggregator and no consensus: one
client (hospital) holds the secret key and queries one server
(lab) which holds the model coefficients in cleartext. The
homomorphic primitives all come from the `openfhe.R` package.

## Scenario: weighted biomarker score

The lab's model computes a weighted score:

$$\text{score} = w_1 x_1 + w_2 x_2 + w_3 x_3 + w_4 x_4 + b$$

where $x_i$ are biomarker values and $w_i, b$ are proprietary
coefficients.

## Setup: hospital's context and patient data

```{r}
library(openfhe.R)

cc <- fhe_context("CKKS",
                  multiplicative_depth = 2L,
                  scaling_mod_size     = 50L,
                  batch_size           = 8L)
keys <- key_gen(cc, eval_mult = TRUE)

## 8 patients, each with 4 biomarker values. We pack each biomarker
## across patients (SIMD layout): one ciphertext per biomarker, with
## patient values in the slots.
biomarker1 <- c(1.2, 0.8, 1.5, 0.3, 2.1, 0.9, 1.1, 1.8)
biomarker2 <- c(0.5, 1.1, 0.3, 0.8, 0.2, 1.4, 0.7, 0.6)
biomarker3 <- c(2.0, 1.5, 2.3, 1.0, 1.8, 2.1, 1.6, 2.5)
biomarker4 <- c(0.1, 0.4, 0.2, 0.6, 0.3, 0.1, 0.5, 0.2)

ct1 <- encrypt(keys@public, make_ckks_packed_plaintext(cc, biomarker1), cc = cc)
ct2 <- encrypt(keys@public, make_ckks_packed_plaintext(cc, biomarker2), cc = cc)
ct3 <- encrypt(keys@public, make_ckks_packed_plaintext(cc, biomarker3), cc = cc)
ct4 <- encrypt(keys@public, make_ckks_packed_plaintext(cc, biomarker4), cc = cc)
```

## Lab side: apply the model to encrypted data

The lab receives the encrypted biomarkers and applies its
proprietary model — without ever seeing patient values.

```{r}
## Lab's proprietary model weights (never shared with the hospital)
w <- c(0.35, -0.20, 0.50, 0.15)
b <- 1.2

## Encrypted score = w1*x1 + w2*x2 + w3*x3 + w4*x4 + b
ct_score <- ct1 * w[1] + ct2 * w[2] + ct3 * w[3] + ct4 * w[4] + b
```

## Hospital side: decrypt the results

```{r}
result <- decrypt(ct_score, keys@secret, cc = cc)
set_length(result, 8L)
scores <- get_real_packed_value(result)[1:8]

for (i in seq_len(8)) {
    risk <- if (scores[i] > 2.0) "HIGH"
            else if (scores[i] > 1.5) "MODERATE"
            else "LOW"
    cat(sprintf("  Patient %d: %.3f (%s)\n", i, scores[i], risk))
}
```

## Verification

```{r}
cleartext_scores <- w[1] * biomarker1 + w[2] * biomarker2 +
                    w[3] * biomarker3 + w[4] * biomarker4 + b
max_error <- max(abs(scores - cleartext_scores))
sprintf("Maximum error vs cleartext: %.2e", max_error)
```

CKKS gives essentially the same answer as cleartext, within
floating-point precision.

## Transport guarantees

The pipeline above delivers two concrete protections:

1. **Biomarker values never appear in cleartext outside the
   hospital.** The lab's view of the protocol consists of the
   encrypted values it received and the encrypted score it
   returned.
2. **The lab's coefficients $w$ and $b$ are never sent to the
   hospital in cleartext.** They are used only to construct the
   returned encrypted score inside the lab's R session.

| Party | Cleartext view |
|-------|----------------|
| Hospital | Patient biomarkers (local), decrypted scores |
| Lab | Encrypted values only — no cleartext biomarker values, no cleartext scores |

These are necessary conditions for any model-as-a-service
deployment that does not trust the lab with cleartext patient
data. They are not sufficient conditions, as the next section
shows.

## Threat model: model extraction from the hospital side

The hospital holds the secret key and decides what goes into the
encrypted queries. Nothing in the FHE pipeline restricts the
biomarker values the hospital encrypts. For a linear model with
four biomarkers and a bias, the hospital can recover every
coefficient with five queries by submitting the standard basis:

- $\mathbf{e}_0 = (0,0,0,0)\ \Rightarrow\ \text{score} = b$
- $\mathbf{e}_1 = (1,0,0,0)\ \Rightarrow\ \text{score} = w_1 + b$
- $\mathbf{e}_2 = (0,1,0,0)\ \Rightarrow\ \text{score} = w_2 + b$
- $\mathbf{e}_3 = (0,0,1,0)\ \Rightarrow\ \text{score} = w_3 + b$
- $\mathbf{e}_4 = (0,0,0,1)\ \Rightarrow\ \text{score} = w_4 + b$

Subtracting the first score from each of the others recovers the
four weights exactly. We can run this attack in the same R
session: wrap the lab's scoring pipeline as a function that
closes over $w$ and $b$ without revealing them, then pack the
five probes across slots 1–5 of the four encrypted biomarker
vectors.

```{r}
## Lab pipeline wrapped as a function. Closes over `w` and `b`;
## the caller (hospital) never reads either.
lab_score <- function(ct_bio) {
    ct_bio[[1]] * w[1] + ct_bio[[2]] * w[2] +
    ct_bio[[3]] * w[3] + ct_bio[[4]] * w[4] + b
}

## Hospital crafts five probes packed across slots 1..5.
## Slot 1 is e_0 (all zeros, probes b). Slot j+1 is e_j (a one in
## position j, probes w_j + b).
probe_bio1 <- c(0, 1, 0, 0, 0, 0, 0, 0)
probe_bio2 <- c(0, 0, 1, 0, 0, 0, 0, 0)
probe_bio3 <- c(0, 0, 0, 1, 0, 0, 0, 0)
probe_bio4 <- c(0, 0, 0, 0, 1, 0, 0, 0)

ct_probe <- list(
    encrypt(keys@public, make_ckks_packed_plaintext(cc, probe_bio1), cc = cc),
    encrypt(keys@public, make_ckks_packed_plaintext(cc, probe_bio2), cc = cc),
    encrypt(keys@public, make_ckks_packed_plaintext(cc, probe_bio3), cc = cc),
    encrypt(keys@public, make_ckks_packed_plaintext(cc, probe_bio4), cc = cc)
)

ct_probe_score <- lab_score(ct_probe)
probe_result   <- decrypt(ct_probe_score, keys@secret, cc = cc)
set_length(probe_result, 5L)
probe_scores <- get_real_packed_value(probe_result)[1:5]

b_hat <- probe_scores[1]
w_hat <- probe_scores[2:5] - b_hat

recovered <- rbind(
    true      = c(b, w),
    recovered = c(b_hat, w_hat)
)
colnames(recovered) <- c("b", "w1", "w2", "w3", "w4")
round(recovered, 6)
```

The recovered coefficients match the lab's true coefficients to
CKKS precision. Five queries — one bias probe plus one per
biomarker — are enough because the scoring function is linear in
the biomarkers; a linear function of $k$ inputs is fully specified
by any $k + 1$ affinely independent point evaluations. FHE does
not impede this extraction: the hospital is the party that
decrypts, the scoring function is the thing being released, and
the standard basis is a legal query vector.

## What this means for deployment

A deployment that needs the scoring function to stay proprietary
has to layer defences *on top of* the FHE transport:

- **Output perturbation.** Adding calibrated noise to the
  decrypted score — a differential-privacy mechanism — turns
  each query into a noisy observation rather than an exact
  readout. Extraction accuracy degrades with the noise variance
  and a query budget can be accounted for explicitly.
- **Query-budget accounting.** Capping the number of queries a
  client can issue in a session bounds how much of the model can
  leak before the budget is exhausted. This composes with output
  perturbation rather than replacing it.
- **Threshold FHE.** Splitting the secret key across several
  parties (see `vignette("cox-threshold")` for the master/worker
  case) changes who can decrypt and under what audit trail. The
  hospital alone can no longer mount the adaptive-query attack;
  decryption requires cooperation from the other key-holders.
- **Non-linear scoring structure.** A linear model is the
  trivial extraction case. Non-linear scorers
  (`openfhe.R::eval_logistic`, polynomial compositions) resist
  closed-form basis-vector attacks, though approximation attacks
  from the model-extraction literature still apply.

The vignette above shows the transport mechanics end-to-end. It
is a *toy illustration* in that the linear scorer plus
hospital-held secret key is exactly the configuration where the
five-query attack works; a production model-as-a-service
deployment is the transport layer plus at least one of the
defences above.

## Network protocol: serialization

In practice the hospital and lab are on different machines. All
objects serialize for network transport:

```{r}
tdir <- tempdir()
fhe_serialize(cc, file.path(tdir, "context.bin"))
fhe_serialize(keys@public, file.path(tdir, "pubkey.bin"))
fhe_serialize(ct1, file.path(tdir, "patient_bm1.bin"))

## Lab receives the serialized files
cc_lab <- fhe_deserialize(file.path(tdir, "context.bin"), "CryptoContext")
ct_lab <- fhe_deserialize(file.path(tdir, "patient_bm1.bin"), "Ciphertext")

## Lab applies its weights to the deserialized ciphertext
ct_weighted <- ct_lab * 0.35

## Lab returns the result
fhe_serialize(ct_weighted, file.path(tdir, "weighted.bin"))

## Hospital receives, deserializes, decrypts
ct_recv <- fhe_deserialize(file.path(tdir, "weighted.bin"), "Ciphertext")
result  <- decrypt(ct_recv, keys@secret, cc = cc)
set_length(result, 8L)
get_real_packed_value(result)[1:8]
```

```{r echo=FALSE}
## Detach openfhe.R so its `encrypt`/`decrypt` generics do not bleed
## into other vignettes built later in the same R session.
detach("package:openfhe.R", unload = FALSE, force = TRUE)
```
