---
title: "Federated Cosine Similarity with Site-Private Fine-Tuned Models"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Federated Cosine Similarity with Site-Private Fine-Tuned Models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## Introduction
A recurring problem in federated medical analytics is similarity
search across silos. A clinician at one hospital sees an unusual
case and wants to ask "do other hospitals in our network have
similar patients?" — without sending the patient out, without
the other hospitals revealing which of their patients were
checked, and without any single party (including a coordinating
master) seeing similarity scores in the clear.

The natural primitive is cosine similarity between embeddings.
Modern foundation models map patient data (radiology reports,
chest X-rays, pathology slides) into a fixed-dimensional vector
space where geometric proximity tracks clinical similarity.
Cosine similarity reduces to an inner product on unit-norm
vectors, which is exactly what threshold CKKS computes
efficiently.

The wrinkle is that hospitals do not all use the *same* model.
Each hospital starts from a public foundation model and
fine-tunes it on its own data for its own purposes — it does not
constrain the result to be an isometry of the public model. The
fine-tuned embeddings therefore differ from the public baseline
by a generically *non-isometric* transformation of the geometry,
and a query in the public-model space is no longer directly
comparable to a database vector in a hospital's private space.

This vignette closes the gap with a per-site *compatibility
adapter* $A_k$, fit post hoc on a public anchor cohort. The
adapter lives on a one-parameter family indexed by a
near-isometry penalty $\mu$: orthogonal Procrustes at one end
($\mu\to\infty$), unconstrained least squares at the other
($\mu = 0$), near-orthogonal maps in between. It is deployed in
either of two ways — applied to the encrypted query (the
homomorphic matrix–vector multiply, valid as a cosine near the
orthogonal end), or folded into the database offline so the
encrypted query reduces to a plain inner product. Threshold
decryption returns the top-`k` matches, with no party able to
decrypt any intermediate value unilaterally.

The accompanying `similarity-sideexp.md` document describes a
deferred experiment on a real fine-tuned model. The vignette
uses synthetic data so it renders quickly during package build.

## The setup

Three sites — labeled $S_1, S_2, S_3$ — each hold a cohort of
patients. Each cohort is embedded with the site's privately
fine-tuned model into $\mathbf{R}^p$ for some fixed dimension
$p$ (the architecture is shared across sites; only the weights
differ).

The query enters the system from a clinician who does not have
access to any site's private model. They embed their candidate
patient with the *public* foundation model, producing $q \in
\mathbf{R}^p$. The query is encrypted under a joint threshold
key whose secret-key material is shared $n$-of-$n$ across the
three sites.

The retrieval target is the top-$k$ patients across all sites
whose private-model embeddings are most cosine-similar to the
query, returned as a list of $(\text{site\_id},
\text{local\_patient\_id}, \text{score})$ tuples.

The friction is that $q$ lives in the public-model geometry
while each site's database lives in the site's private-model
geometry. We close the gap with a per-site *compatibility
adapter* $A_k$ fit on a public anchor cohort. $A_k$ is held
privately by site $k$ and never leaves; it is applied to
encrypted queries to bring them into agreement with site $k$'s
geometry before the inner-product step (or, equivalently, folded
into the site's database offline).

## Threat model

Three sites and one untrusted aggregator (the master):

- **Sites $S_1, S_2, S_3$** each hold private patient
  embeddings, a private compatibility adapter $A_k$, and a secret-key
  share $\mathit{sk}_k$. They are honest-but-curious among
  themselves and toward the master. Each site sees the master's
  encrypted query but cannot decrypt it; each site sees its own
  database in the clear (it is its own data); each site never
  sees other sites' databases or scores.
- **Master** holds no secret-key material. It receives the
  encrypted query from the querier, broadcasts it to the sites,
  collects their encrypted scores, orchestrates the n-of-n
  partial-decryption ceremony for the top-$k$ results, and
  returns them to the querier. A curious or compromised master
  cannot decrypt anything by itself.
- **Querier** sees only the top-$k$ result tuples. Repeated
  adaptive queries leak structural information about the
  database in the same Hyrum-style sense as any retrieval
  system; this leakage is acknowledged but not eliminated.

What the master sees, by stage:

1. The encrypted query $\mathit{ct}_q$. Not decryptable alone.
2. Per-site encrypted scores. Not decryptable alone.
3. Partial decryptions of the top-$k$ scores from each site.
   Not decryptable individually.
4. After fusion, the top-$k$ scores in the clear. Released to
   the querier as the protocol output.

Step 4 is the only cleartext the master ever holds, and only
for the top-$k$ scores released to the querier — not for any
intermediate quantity.

The site-private compatibility adapters $\{A_k\}$ never leave
their sites. Each is applied to an encrypted vector while it
stays encrypted — an unencrypted matrix times an encrypted
vector — or else folded into the site's database offline, so the
adapter never appears in the clear at any party other than its
owner.

## The protocol

The protocol has three phases. The setup phase runs once,
offline; the query phase and result phase run for each query.

**Setup phase.**

1. The three sites jointly generate a CKKS threshold key pair
   ($n$-of-$n$). Joint rotation keys for slot rotations
   $1, 2, \ldots, p-1$ are also generated by an $n$-of-$n$
   ceremony.
2. A public anchor cohort of patients with public-model
   embeddings is published.
3. Each site embeds the anchor cohort with its private
   fine-tuned model and fits its $p \times p$ compatibility
   adapter $A_k$ along the $\mu$ family (orthogonal Procrustes
   at $\mu\to\infty$, least squares at $\mu = 0$).
4. Each site stores its local cohort, embedded with the
   private model and normalized to unit length — and, for the
   fold (Design 1) deployment, the public-compatible images of
   those vectors.

**Query phase.** The querier embeds their candidate patient
with the public model, normalizes to unit length, encrypts
into a single slot-packed encrypted vector under the joint
public key, and broadcasts that to the master.

For each site $k$:

5. The site receives the encrypted query $\mathit{ct}_q$.
6. The site applies its private $A_k$ to the encrypted query via
   diagonal-encoded matrix–vector multiplication, producing
   $\mathit{ct}_{A_k q}$ (Design 2). Near the orthogonal end of
   the family the result is unit-norm, so the score is a genuine
   cosine; alternatively the site skips this step and scores the
   query directly against its folded public-compatible database
   (Design 1).
7. The site computes inner products against each of its local
   embeddings $v_{k,i}$ via slot-wise multiplication followed
   by log-$p$ rotation-and-add slot summation, producing one
   encrypted scalar per local patient.
8. The site returns its encrypted scores to the master,
   alongside its local patient indices in the clear.

**Result phase.**

9. The master concatenates encrypted scores across all sites.
   In v1, local indices are kept in the clear alongside the
   encrypted scores; encrypted top-$k$ selection in CKKS is
   feasible but adds depth and complexity orthogonal to this
   vignette's pedagogical aim.
10. The master collects partial decryption shares from all
    three sites for the encrypted scores. Fusion yields the
    scores in the clear; the master sorts them there and
    returns the top-$k$ to the querier.

## Synthetic data

The vignette uses fully synthetic data so it renders quickly
during package build. The phenotype-mixture model below is the
simplest setup that gives a meaningful retrieval ground truth.

```{r}
suppressPackageStartupMessages({
  library(homomorpheR)
  library(openfhe.R)
})
set.seed(20260428)

p             <- 32L
n_sites       <- 3L
cohort_sizes  <- c(80L, 60L, 100L)
n_anchor      <- 100L
n_phenotypes  <- 5L
top_k         <- 5L
```

A patient is one of `n_phenotypes` clinical phenotypes. Each
phenotype is a Gaussian cluster in $\mathbf{R}^p$. Phenotype
labels are the retrieval ground truth: a query of phenotype
$j$ should be matched to database patients of phenotype $j$.

```{r}
phenotype_centers <- matrix(rnorm(n_phenotypes * p, sd = 1),
                            n_phenotypes, p)

embed_public <- function(n) {
  labels  <- sample.int(n_phenotypes, n, replace = TRUE)
  centers <- phenotype_centers[labels, , drop = FALSE]
  noise   <- matrix(rnorm(n * p, sd = 0.4), n, p)
  z <- centers + noise
  z <- z / sqrt(rowSums(z^2))            # unit norm
  list(z = z, label = labels)
}
```

A site fine-tunes for its own purposes; it does not constrain
the result to be an isometry of the public model. We therefore
model the per-site drift as a *non-isometric* linear map
$B_k = Q_k D_k$ — a random rotation $Q_k$ composed with a
diagonal stretch $D_k = \mathrm{diag}(e^{\beta g})$, $g$
standard normal. The parameter $\beta$ is the non-isometry
magnitude: $\beta = 0$ recovers an exactly orthogonal drift
(the special case where the geometry is merely rotated), and
$\beta > 0$ stretches the embedding directions anisotropically,
as a freely fine-tuned model generically would. The site's
embeddings are unit-normalized after the map, so the drift acts
on the sphere.

```{r}
random_drift <- function(p, beta) {
  ## Free fine-tuning, simulated as B = Q D: a random rotation Q
  ## composed with an anisotropic stretch D = diag(exp(beta g)).
  ## beta = 0 gives an exactly orthogonal (isometric) drift;
  ## beta > 0 is non-isometric, the generic fine-tuned case.
  Q <- qr.Q(qr(matrix(rnorm(p * p), p, p)))
  if (beta == 0) return(Q)
  Q %*% diag(exp(beta * rnorm(p)))
}

embed_private <- function(z_public, B_k) {
  ## Site k's fine-tuned model in column-vector convention:
  ## f_k(x) = B_k · f(x). For a matrix z_public of n row-stacked
  ## vectors, the private embeddings are z_public %*% t(B_k),
  ## then unit-normalized (the drift acts on the sphere; under a
  ## non-isometric B the normalization is a genuine nonlinearity).
  v <- z_public %*% t(B_k)
  v / sqrt(rowSums(v^2))
}
```

The public anchor cohort is a small, publicly-available set
of patient examples. Each site embeds the anchors twice — once
with the public model (giving $Z_{\text{pub}}$, identical at
every site by construction) and once with its own private
fine-tuned model (giving $Z_{\text{priv},k}$). The pair drives a
post-hoc *compatibility adapter* $A_k$ that maps the site's
private geometry back to the public protocol, fit by

$$
\min_{A}\ \lVert Z_{\text{priv},k}\,A - Z_{\text{pub}}\rVert_F^2
        \;+\; \mu\,\lVert A^\top A - I\rVert_F^2 .
$$

The penalty parameter $\mu$ traces a one-parameter family. At
$\mu \to \infty$ the adapter is forced orthogonal and the fit is
the classical orthogonal Procrustes problem, solved in closed
form by the singular value decomposition — this is the rigid
endpoint, exact only when the drift is itself isometric. At
$\mu = 0$ the adapter is an unconstrained least-squares map, the
best public-space fit but generically non-isometric. Intermediate
$\mu$ interpolates: a *near-orthogonal* map. We deploy $A_k$ by
applying it to the encrypted query, $\langle A_k q,\, v\rangle$,
which recovers the public-space cosine $\langle q, u\rangle$ when
the adapter inverts the drift (exactly, in the orthogonal limit).

```{r}
fit_adapter <- function(Z_priv, Z_pub, mu) {
  ## Compatibility adapter A (private -> public): Z_priv A ~ Z_pub,
  ## with a near-isometry penalty mu * ||A^T A - I||^2.
  p <- ncol(Z_priv)
  if (is.infinite(mu)) {                       # orthogonal Procrustes
    sv <- svd(crossprod(Z_priv, Z_pub))
    return(sv$u %*% t(sv$v))
  }
  A_ls <- solve(crossprod(Z_priv) + 1e-6 * diag(p), crossprod(Z_priv, Z_pub))
  if (mu == 0) return(A_ls)                     # least squares
  fn <- function(par) { A <- matrix(par, p, p)  # near-orthogonal
    sum((Z_priv %*% A - Z_pub)^2) + mu * sum((crossprod(A) - diag(p))^2) }
  gr <- function(par) { A <- matrix(par, p, p)
    as.vector(2 * crossprod(Z_priv, Z_priv %*% A - Z_pub) +
              4 * mu * (A %*% (crossprod(A) - diag(p)))) }
  matrix(optim(as.vector(A_ls), fn, gr, method = "L-BFGS-B",
               control = list(maxit = 400))$par, p, p)
}
```

A convex alternative one might reach for is *Gram / metric
learning* (§10.4 of the companion exploration): fit a
positive-semidefinite $M$ matching the public Gram matrix,
$\min_{M\succeq0}\lVert Z_{\text{priv}}MZ_{\text{priv}}^\top -
Z_{\text{pub}}Z_{\text{pub}}^\top\rVert_F^2$. Its minimizer is
analytic — $M = A_{\text{LS}}A_{\text{LS}}^\top$ — and we deploy
its symmetric square root. We include it only to show, below,
that it is the *wrong* tool here.

```{r}
fit_gram <- function(Z_priv, Z_pub) {
  M <- tcrossprod(solve(crossprod(Z_priv) + 1e-6 * diag(ncol(Z_priv)),
                        crossprod(Z_priv, Z_pub)))
  e <- eigen(M, symmetric = TRUE)
  e$vectors %*% (sqrt(pmax(e$values, 0)) * t(e$vectors))   # symmetric M^{1/2}
}

unit_rows <- function(Z) Z / sqrt(rowSums(Z^2))            # for Design-1 folding
```

A moderate non-isometry magnitude drives the protocol
walk-through; the sweeps later vary it:

```{r}
beta_demo <- 0.6
mu_demo   <- Inf      # orthogonal endpoint for the encrypted walk-through
```

The site cohorts and anchor cohort are sampled from the same
phenotype-mixture model (in real deployments the anchor cohort
is publicly distributed and its phenotype distribution is set
once; we approximate by independent sampling).

```{r}
public_anchor <- embed_public(n_anchor)
public_query  <- embed_public(1L)        # one query for the protocol walk-through

site_cohorts <- lapply(cohort_sizes, embed_public)
```

We later fit one adapter $A_k$ per site, build both the
raw-private and folded public-compatible databases, and report
retrieval recall as the non-isometry $\beta$ and the penalty
$\mu$ vary.

## Threshold key generation

The three sites jointly construct a CKKS keypair so that the
secret key is split $n$-of-$n$ across them. The familiar
`make_threshold_master()` from `homomorpheR` wires the joint
public key and the per-site secret-key shares for us; the
master is given the joint public key but never holds usable
secret material.

```{r}
cc <- fhe_context(
  scheme               = "CKKS",
  multiplicative_depth = 3L,
  scaling_mod_size     = 45L,
  batch_size           = p,
  features             = c(Feature$MULTIPARTY, Feature$KEYSWITCH))

master <- make_threshold_master(name = "master",
                                crypto_context = cc,
                                n_sites = n_sites)
```

The joint public key is `master@joint_pubkey`; the secret-key
shares are `master@secret_keys[[k]]` for $k = 1, \ldots, n$.
The `master_encrypt(master, x)` and `master_decrypt(master,
ct, len)` generics dispatch on `ThresholdMaster` and run the
$n$-of-$n$ partial-decryption ceremony internally — no party
ever holds a usable secret unilaterally.

## Joint rotation keys

The matrix–vector multiply step in the protocol is implemented
as a *diagonal-encoded matvec*. For a $p \times p$ matrix $M$
and an encrypted vector $q \in \mathbf{R}^p$ held one component
per slot:

$$
M \cdot q \;=\; \sum_{i=0}^{p-1} d_i \odot \mathrm{rot}(q, i),
$$

where $d_i$ is the $i$-th diagonal of $M$ (a $p$-vector, and
never encrypted), $\odot$ is slot-wise multiplication, and
$\mathrm{rot}(q, i)$ cyclically rotates the slots of $q$ by
$i$. Each rotation requires a precomputed *rotation key*. In
single-key CKKS these are generated from the secret key; in
threshold CKKS they are generated by an $n$-of-$n$ ceremony
that mirrors the encryption-key ceremony.

The inner-product step adds a second use of rotation keys: the
log-$p$ rotation-and-add reduction that sums the slots of an
encoded $p$-vector into slot 0. The same set of rotation
indices serves both purposes; we generate keys for indices
$1, \ldots, p-1$ and rely on the subset that each step needs.

```{r}
rotation_indices <- seq_len(p - 1L)
sks  <- master@secret_keys
pks  <- list(master@joint_pubkey)        # daisy-chain pubkey at step 1
joint_pk_tag <- get_key_tag(master@joint_pubkey)

## Lead-party rotation-key generation. Populates the crypto
## context's automorphism-key registry under the lead party's
## secret-key tag.
eval_rotate_key_gen(cc, sks[[1]], rotation_indices)

lead_tag    <- get_key_tag(sks[[1]])
rot_running <- get_eval_automorphism_key_map(lead_tag)

## Sites 2..n contribute their rotation-key shares in turn.
## At each step, the running joint share is registered under
## the cumulative-pubkey tag — which for our `make_threshold_master`
## is the same `joint_pk_tag` at every step (the master daisy-chains
## pubkeys forward but only retains the final joint pubkey).
for (k in 2:n_sites) {
  share_k <- multi_eval_at_index_key_gen(
    cc, sks[[k]], rot_running,
    index_list = rotation_indices,
    key_tag = joint_pk_tag)
  rot_running <- multi_add_eval_automorphism_keys(
    cc, rot_running, share_k, key_tag = joint_pk_tag)
}

insert_eval_automorphism_key(rot_running, key_tag = joint_pk_tag)
```

After insertion, any value encrypted under the joint
public key (i.e., under `master@joint_pubkey`) can be rotated
by any index in `rotation_indices` via `eval_rotate(ct, idx)`.

A round-trip smoke test confirms the joint rotation key works.
We encrypt a known vector, rotate by 3 slots, decrypt via the
$n$-of-$n$ ceremony, and check that slot $i$ now holds
$x_{(i+3) \bmod p}$:

```{r}
x <- 1:p
ct_x      <- master_encrypt(master, x)
ct_rot    <- eval_rotate(ct_x, 3L)
recovered <- master_decrypt(master, ct_rot, len = p)

expected <- x[((seq_len(p) - 1L + 3L) %% p) + 1L]
stopifnot(max(abs(recovered - expected)) < 1e-6)
cat("rotation round-trip max error:",
    sprintf("%.2e\n", max(abs(recovered - expected))))
```

## Per-site adapter fit and database setup

For the encrypted walk-through we fit the adapter at the
orthogonal endpoint ($\mu = \infty$), so the map applied to the
query is exactly norm-preserving and the matrix–vector multiply
below produces a unit-norm result — the regime in which the
raw-private-database matvec (Design 2) is a valid cosine. The
recall sweeps later vary $\mu$ and the drift. Each site's
fine-tuned model is the non-isometric $B_k$; the adapter $A_k$
is fit post hoc on the public anchor cohort.

```{r}
B <- lapply(seq_len(n_sites), function(k) random_drift(p, beta_demo))

## Each site's private database, embedded under that site's
## fine-tuned model and unit-normalized on the sphere.
db <- lapply(seq_len(n_sites), function(k) {
  cohort <- site_cohorts[[k]]
  list(z = embed_private(cohort$z, B[[k]]),
       label = cohort$label)
})

## Each site fits its compatibility adapter on the anchor cohort.
A_hat <- lapply(seq_len(n_sites), function(k) {
  Z_pub  <- public_anchor$z
  Z_priv <- embed_private(Z_pub, B[[k]])
  fit_adapter(Z_priv, Z_pub, mu_demo)
})

## Design-1 deployment: the adapter folded into the database
## offline, giving unit-norm public-compatible vectors that an
## encrypted public-model query scores directly. (At mu = Inf the
## adapter is orthogonal, so folding is norm-preserving and
## Design 1 and Design 2 coincide; they part company at finite mu.)
db_fold <- lapply(seq_len(n_sites), function(k)
  list(z = unit_rows(db[[k]]$z %*% A_hat[[k]]), label = db[[k]]$label))

## Setup diagnostics the master would receive: anchor-reconstruction
## error and the adapter's departure from isometry.
cat("Per-site adapter diagnostics (beta =", beta_demo, ", mu = Inf):\n")
for (k in seq_len(n_sites)) {
  Zr <- embed_private(public_anchor$z, B[[k]])
  recon <- norm(Zr %*% A_hat[[k]] - public_anchor$z, "F")
  aniso <- norm(crossprod(A_hat[[k]]) - diag(p), "F")
  cat(sprintf("  site %d: anchor recon %.2e,  ||A^T A - I||_F %.2e\n",
              k, recon, aniso))
}
```

## Diagonal-encoded matrix-vector multiply

The CKKS-friendly way to multiply an unencrypted $p \times p$
matrix into an encrypted vector is the *diagonal encoding*. The
matrix $M$ is decomposed into its $p$ generalized diagonals,
each an unencrypted length-$p$ vector:

$$
d_i[j] = M[\,j,\, (j + i) \bmod p\,], \qquad i = 0, 1, \ldots, p-1.
$$

The matrix–vector product becomes

$$
M \cdot q \;=\; \sum_{i=0}^{p-1} d_i \,\odot\, \mathrm{rot}(q, i),
$$

where $\odot$ is slot-wise multiplication and $\mathrm{rot}(q, i)$
is the $i$-step cyclic slot rotation. Each term multiplies an
encrypted vector by an unencrypted one and adds the result to a
running encrypted total. The cost is $p$ rotations and $p$ such
multiplications per matrix-vector product, and it consumes a
single level of the precision budget.

Multiplying by an unencrypted vector rather than an encrypted
one is what keeps that cost down: the adapter $A_k$ is the
site's own, so it never needs encrypting, and the operation is
correspondingly cheaper than multiplying two encrypted
quantities together.

```{r}
build_diagonals <- function(M, p) {
  ## d_i[j] = M[j, ((j-1 + i) %% p) + 1]   (1-based R indexing)
  lapply(0:(p - 1L), function(i) {
    vapply(seq_len(p),
           function(j) M[j, ((j - 1L + i) %% p) + 1L],
           numeric(1L))
  })
}

encrypted_matvec <- function(ct_q, M, cc, p) {
  diags <- build_diagonals(M, p)
  ct_acc <- NULL
  for (i in 0:(p - 1L)) {
    d_pt <- make_ckks_packed_plaintext(cc, diags[[i + 1L]])
    ct_term <- if (i == 0L) {
      eval_mult(ct_q, d_pt)
    } else {
      eval_mult(eval_rotate(ct_q, i), d_pt)
    }
    ct_acc <- if (is.null(ct_acc)) ct_term else eval_add(ct_acc, ct_term)
  }
  ct_acc
}
```

A round-trip smoke test on a known query vector confirms the
matvec recovers $A_1 \cdot q$ to floating-point precision:

```{r}
q_demo <- public_query$z[1, ]
ct_q  <- master_encrypt(master, q_demo)
ct_Aq <- encrypted_matvec(ct_q, A_hat[[1]], cc, p)
Aq_recovered <- master_decrypt(master, ct_Aq, len = p)
Aq_expected  <- as.numeric(A_hat[[1]] %*% q_demo)
cat(sprintf("matvec max error (site 1): %.2e\n",
            max(abs(Aq_recovered - Aq_expected))))
```

## Inner product against the local database

After the matvec, the encrypted query has been mapped by the
adapter into site $k$'s private geometry (unit-norm at the
orthogonal endpoint used here). The cosine similarity against a
private-database vector $v$ (also unit-norm, and never
encrypted since it never leaves the site) is just

$$
\cos(A_k q,\, v) \;=\; \langle A_k q,\, v \rangle
\;=\; \sum_{j=1}^{p} (A_k q)_j \cdot v_j.
$$

In CKKS this is a slot-wise multiply of the matvec output by the
unencrypted $v$, followed by a log-$p$ rotate-and-add
*slot-summation reduction* that places the summed inner product
in slot 0 (and uninteresting partial sums in the other slots).

```{r}
slot_sum_reduction <- function(ct, p) {
  ## Standard CKKS log-p reduction. After the loop, slot 0 of
  ## the result holds sum_{j=1}^{p} ct[j]; other slots hold
  ## partial sums and are not used.
  step <- p %/% 2L
  while (step >= 1L) {
    ct <- eval_add(ct, eval_rotate(ct, step))
    step <- step %/% 2L
  }
  ct
}

encrypted_inner_product <- function(ct_x, v_plain, cc, p) {
  pt_v <- make_ckks_packed_plaintext(cc, v_plain)
  ct_prod <- eval_mult(ct_x, pt_v)
  slot_sum_reduction(ct_prod, p)
}
```

A smoke test against a single database vector confirms the
inner product matches the unencrypted computation at slot 0:

```{r}
v_test <- db[[1]]$z[1, ]
ct_score <- encrypted_inner_product(ct_Aq, v_test, cc, p)
score_recovered <- master_decrypt(master, ct_score, len = 1L)
score_expected  <- sum(Aq_expected * v_test)
cat(sprintf("inner-product error (site 1, patient 1): %.2e\n",
            abs(score_recovered - score_expected)))
```

The same `encrypted_inner_product` serves the **Design 1**
deployment without any matvec: the encrypted *public-model*
query is scored directly against the folded, public-compatible
database vectors. At the orthogonal endpoint this matches the
matvec route exactly; for a non-isometric adapter it is the
route that stays a valid cosine.

```{r}
ct_score_fold <- encrypted_inner_product(ct_q, db_fold[[1]]$z[1, ], cc, p)
fold_recovered <- master_decrypt(master, ct_score_fold, len = 1L)
fold_expected  <- sum(q_demo * db_fold[[1]]$z[1, ])
cat(sprintf("Design-1 inner-product error (site 1, patient 1): %.2e\n",
            abs(fold_recovered - fold_expected)))
```

## Site `local_fn`: full per-site protocol step

The site-side computation closes over the site's adapter
$A_k$ and database $D_k$, takes the encrypted query as input,
and returns a list of encrypted scores plus the corresponding
local indices in the clear. This is the **Design 2** branch:
the adapter is applied to the encrypted query (the matvec), and
scoring runs against the site's raw private database.

```{r}
make_similarity_site_fn <- function(A_k, db_k, cc, p) {
  function(ct_q) {
    ct_Aq <- encrypted_matvec(ct_q, A_k, cc, p)
    n_local <- nrow(db_k$z)
    ct_scores <- lapply(seq_len(n_local), function(i) {
      encrypted_inner_product(ct_Aq, db_k$z[i, ], cc, p)
    })
    list(scores = ct_scores,
         local_index = seq_len(n_local),
         label = db_k$label)
  }
}

site_fns <- lapply(seq_len(n_sites), function(k) {
  make_similarity_site_fn(A_hat[[k]], db[[k]], cc, p)
})
```

A timed end-to-end run on the demo query through site 1
returns one encrypted score per local patient:

```{r}
t0 <- proc.time()
site1_out <- site_fns[[1]](ct_q)
elapsed <- (proc.time() - t0)[["elapsed"]]
cat(sprintf("site 1 produced %d encrypted scores in %.2f s\n",
            length(site1_out$scores), elapsed))
```

## Master orchestration and threshold-decrypted top-`k`

The master broadcasts the encrypted query to every site,
collects per-site encrypted scores plus the local patient
indices in the clear, and runs the $n$-of-$n$ threshold
decryption ceremony for each score so it can sort them once
they are in the clear and return the top-`k`.

The decryption pattern is per-patient: each encrypted score
goes through one threshold-decrypt round
(`master_decrypt()` calls `multiparty_decrypt_lead` on site 1,
`multiparty_decrypt_main` on each remaining site, and
`multiparty_decrypt_fusion` to combine). For our `r sum(cohort_sizes)`
total patients this runs in a few seconds; production
deployments would pack many patients into the slots of a
single encrypted value and amortize the ceremony.

```{r}
run_similarity_query <- function(ct_q, site_fns, master, top_k) {
  ## Fan out to every site.
  site_results <- lapply(seq_along(site_fns), function(k) {
    out <- site_fns[[k]](ct_q)
    out$site_id <- k
    out
  })

  ## Threshold-decrypt each per-patient inner-product
  ## ciphertext. v1 runs one ceremony per patient; a packed
  ## variant that fuses multiple inner products into a single
  ## ciphertext (via slot tiling) is a natural extension.
  rows <- list()
  for (s in site_results) {
    for (i in seq_along(s$scores)) {
      score <- master_decrypt(master, s$scores[[i]], len = 1L)
      rows[[length(rows) + 1L]] <- data.frame(
        site_id     = s$site_id,
        local_index = s$local_index[i],
        label       = s$label[i],
        score       = score)
    }
  }
  scored <- do.call(rbind, rows)
  scored <- scored[order(-scored$score), ]
  head(scored, top_k)
}

t0 <- proc.time()
top_result <- run_similarity_query(ct_q, site_fns, master, top_k = top_k)
elapsed <- (proc.time() - t0)[["elapsed"]]
cat(sprintf("Top-%d retrieval over %d sites and %d patients in %.1f s\n",
            top_k, n_sites, sum(cohort_sizes), elapsed))
cat(sprintf("Query phenotype label: %d\n", public_query$label))
print(top_result, row.names = FALSE)
```

## Reference-truth comparison

The protocol output is meaningful only if the inner products
computed under encryption agree with the same quantities computed
in the clear, to within CKKS approximation error. The reference
computation runs the same $A_k \cdot q$ adapter application and
inner-product reduction unencrypted on each site's database:

```{r}
plaintext_top_k <- function(q, site_data, A_list, top_k) {
  rows <- list()
  for (k in seq_along(site_data)) {
    Aq  <- as.numeric(A_list[[k]] %*% q)
    s_k <- site_data[[k]]
    scores <- as.numeric(s_k$z %*% Aq)
    for (i in seq_along(scores)) {
      rows[[length(rows) + 1L]] <- data.frame(
        site_id     = k,
        local_index = i,
        label       = s_k$label[i],
        score       = scores[i])
    }
  }
  scored <- do.call(rbind, rows)
  scored <- scored[order(-scored$score), ]
  head(scored, top_k)
}

plain_top <- plaintext_top_k(q_demo, db, A_hat, top_k)

## Compare encrypted-domain top-k against the plaintext reference
## by joining on (site_id, local_index).
compare <- merge(top_result, plain_top,
                 by = c("site_id", "local_index"),
                 suffixes = c("_enc", "_plain"))
score_err <- max(abs(compare$score_enc - compare$score_plain))
cat(sprintf("Top-%d encrypted vs plaintext score max error: %.2e\n",
            top_k, score_err))

## Whether the encrypted-domain top-k contains the same
## (site_id, local_index) pairs as the plaintext reference.
enc_set   <- paste(top_result$site_id, top_result$local_index, sep = ":")
plain_set <- paste(plain_top$site_id,  plain_top$local_index,  sep = ":")
cat(sprintf("Top-%d set match: %d of %d\n",
            top_k, length(intersect(enc_set, plain_set)), top_k))
```

## What the adapter buys: fidelity, posture, and a sweet spot

The encrypted mechanics work to floating-point precision, as the
smoke tests and reference comparison above confirm (max error
$\sim 10^{-11}$). The substantive questions are statistical, and
because the encrypted and unencrypted scores agree to within
CKKS approximation error we answer them in the clear — fast
enough to average over many queries and trace stable curves.
Three questions:

1. **Fidelity.** When the drift is genuinely non-isometric, does
   a near-orthogonal or least-squares adapter recover retrieval
   that the rigid orthogonal Procrustes endpoint ($\mu\to\infty$)
   cannot?
2. **Posture.** Where do the two deployments agree — Design 1
   (fold the adapter into the database offline) and Design 2
   (apply it to the encrypted query, the matvec) — and where must
   we prefer one? They coincide only when the adapter is
   orthogonal, so the per-vector norms it produces are all one.
3. **Sweet spot.** Does an intermediate $\mu$ ever beat both
   endpoints, and when?

The sweeps use a less-forgiving cluster configuration than the
walk-through (unit-norm phenotype centers, so noise competes with
separation), letting alignment quality rather than trivial
separability drive recall.

```{r}
centers_u <- phenotype_centers / sqrt(rowSums(phenotype_centers^2))
embed_cfg <- function(n, sep = 0.85, sd = 0.40) {
  labels <- sample.int(n_phenotypes, n, replace = TRUE)
  z <- sep * centers_u[labels, , drop = FALSE] +
       matrix(rnorm(n * p, sd = sd), n, p)
  list(z = unit_rows(z), label = labels)
}

recall_at_k <- function(query_label, top_rows, k = top_k)
  sum(top_rows$label == query_label) / k

## Federated recall of a query population under one deployment.
## design 1: fold A into the db (unit) and score <q, z>;
## design 2: apply A to the query and score <Aq, h> against raw db.
fed_recall <- function(q_pop, db_list, A_list, design) {
  mean(vapply(seq_len(nrow(q_pop$z)), function(qi) {
    q <- q_pop$z[qi, ]
    parts <- lapply(seq_along(db_list), function(k) {
      s <- db_list[[k]]
      sc <- if (design == 1L) as.numeric(unit_rows(s$z %*% A_list[[k]]) %*% q)
            else              as.numeric(s$z %*% as.numeric(A_list[[k]] %*% q))
      data.frame(label = s$label, score = sc)
    })
    scored <- do.call(rbind, parts)
    recall_at_k(q_pop$label[qi],
                head(scored[order(-scored$score), ], top_k))
  }, numeric(1)))
}

## A world at non-isometry beta with an anchor cohort of size na.
make_world <- function(beta, na) {
  anchor  <- embed_cfg(na)
  cohorts <- lapply(cohort_sizes, embed_cfg)
  Bs <- lapply(seq_len(n_sites), function(k) random_drift(p, beta))
  list(anchor  = anchor,
       pub_db  = cohorts,                          # public embeddings (ideal ref)
       priv_db = lapply(seq_len(n_sites), function(k)
         list(z = embed_private(cohorts[[k]]$z, Bs[[k]]),
              label = cohorts[[k]]$label)),
       Hanchor = lapply(seq_len(n_sites), function(k)
         embed_private(anchor$z, Bs[[k]])))
}

I_list  <- replicate(n_sites, diag(p), simplify = FALSE)
n_rep   <- 3L
n_q     <- 40L
mu_grid <- c(0, 0.1, 1, 10, Inf)

## --- mu sweep at fixed beta, ample anchor: fidelity + posture + Gram ---
mu_sweep <- function(beta, na) {
  tab <- 0; ideal <- 0; unaligned <- 0; gram <- 0
  for (r in seq_len(n_rep)) {
    w  <- make_world(beta, na)
    qp <- embed_cfg(n_q)
    ideal     <- ideal     + fed_recall(qp, w$pub_db,  I_list, 2L)
    unaligned <- unaligned + fed_recall(qp, w$priv_db, I_list, 2L)
    Ag   <- lapply(seq_len(n_sites), function(k) fit_gram(w$Hanchor[[k]], w$anchor$z))
    gram <- gram + fed_recall(qp, w$priv_db, Ag, 1L)
    rows <- lapply(mu_grid, function(mu) {
      A <- lapply(seq_len(n_sites), function(k)
        fit_adapter(w$Hanchor[[k]], w$anchor$z, mu))
      data.frame(mu = mu,
                 d1 = fed_recall(qp, w$priv_db, A, 1L),
                 d2 = fed_recall(qp, w$priv_db, A, 2L),
                 aniso = mean(vapply(A, function(Ak)
                   norm(crossprod(Ak) - diag(p), "F"), numeric(1))))
    })
    tab <- tab + do.call(rbind, rows)
  }
  tab <- tab / n_rep
  tab$mu <- mu_grid
  list(tab = tab, ideal = ideal / n_rep,
       unaligned = unaligned / n_rep, gram = gram / n_rep)
}

mu_main   <- mu_sweep(beta = 0.6, na = 100L)   # ample calibration
mu_scarce <- mu_sweep(beta = 0.6, na = 24L)    # anchor < p = 32

## --- beta sweep, ample anchor: Procrustes vs near-orthogonal vs LS ---
## Common random numbers: within a replicate the public data and the
## per-site drift directions are fixed and only the stretch magnitude
## beta varies, so the no-drift ideal is a single drift-independent
## reference rather than a wandering curve.
beta_grid <- c(0, 0.3, 0.6, 1.0)
beta_acc <- 0; ideal_b <- 0
for (r in seq_len(n_rep)) {
  anchor  <- embed_cfg(n_anchor)
  cohorts <- lapply(cohort_sizes, embed_cfg)
  qp      <- embed_cfg(n_q)
  Qg <- lapply(seq_len(n_sites), function(k)
    list(Q = qr.Q(qr(matrix(rnorm(p * p), p, p))), g = rnorm(p)))
  ideal_b  <- ideal_b + fed_recall(qp, cohorts, I_list, 2L)
  beta_acc <- beta_acc + do.call(rbind, lapply(beta_grid, function(b) {
    Bs   <- lapply(Qg, function(qg)
      if (b == 0) qg$Q else qg$Q %*% diag(exp(b * qg$g)))
    priv <- lapply(seq_len(n_sites), function(k)
      list(z = embed_private(cohorts[[k]]$z, Bs[[k]]),
           label = cohorts[[k]]$label))
    Hanc <- lapply(seq_len(n_sites), function(k)
      embed_private(anchor$z, Bs[[k]]))
    fitb <- function(mu) lapply(seq_len(n_sites), function(k)
      fit_adapter(Hanc[[k]], anchor$z, mu))
    data.frame(beta = b,
               LS   = fed_recall(qp, priv, fitb(0), 1L),
               near = fed_recall(qp, priv, fitb(1), 1L),
               Proc = fed_recall(qp, priv, fitb(Inf), 2L))
  }))
}
beta_tab <- beta_acc / n_rep; beta_tab$beta <- beta_grid
ideal_b  <- ideal_b / n_rep

cat("mu sweep (beta=0.6, anchor=100):  ideal=",
    sprintf("%.3f", mu_main$ideal), " unaligned=",
    sprintf("%.3f", mu_main$unaligned), " Gram=",
    sprintf("%.3f\n", mu_main$gram), sep = "")
print(round(mu_main$tab, 3), row.names = FALSE)
cat("\nbeta sweep (anchor=100, common random numbers):  ideal=",
    sprintf("%.3f\n", ideal_b), sep = "")
print(round(beta_tab, 3), row.names = FALSE)
cat("\nmu sweep at scarce anchor=24:\n")
print(round(mu_scarce$tab[c("mu", "d1", "d2")], 3), row.names = FALSE)
```

Three readings come out of the sweeps. **Fidelity** (β panel):
at $\beta = 0$ the drift is isometric and Procrustes, the
near-orthogonal map, and least squares all match the no-drift
ideal; as the drift bends, Procrustes falls away while the
relaxed adapters track the ideal — the richer map earns its keep
exactly when fine-tuning is non-isometric. **Posture** (μ panel):
Design 1 and Design 2 coincide at $\mu \to \infty$ (where the
adapter is orthogonal and the per-vector norms are all one) and
separate as $\mu$ shrinks, so the raw-database matvec is a valid
cosine only near the orthogonal endpoint; the departure-from-
isometry panel quantifies the budget. **Sweet spot** (scarce-
anchor panel): when the calibration cohort is smaller than the
embedding dimension, an intermediate $\mu$ regularizes the
adapter and can beat both endpoints — a secondary effect that
appears only under scarce calibration. The Gram-PSD reference
sits near the unaligned floor: its rotation invariance loses the
public-frame orientation, so it is the wrong tool for a
public-query retrieval (see the discussion).

```{r recall-figure, fig.width=9, fig.height=7, fig.align="center"}
op <- par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.4, 1.0))
xi <- seq_along(mu_grid)
mulab <- function(m) ifelse(is.infinite(m), "Inf", formatC(m, format = "g"))

## (1) recall vs mu: Design 1 vs Design 2, with Gram / ideal / floor
plot(xi, mu_main$tab$d1, type = "b", pch = 19, ylim = c(0, 1), xaxt = "n",
     xlab = expression(mu ~ "(0 = LS    ->    Inf = Procrustes)"),
     ylab = sprintf("recall@%d", top_k), main = "Recall vs mu (beta = 0.6)")
axis(1, xi, mulab(mu_grid))
lines(xi, mu_main$tab$d2, type = "b", pch = 1, lty = 2, col = "firebrick")
abline(h = mu_main$ideal, lty = 3, col = "darkgreen")
abline(h = mu_main$unaligned, lty = 3, col = "grey60")
abline(h = mu_main$gram, lty = 4, lwd = 2, col = "orange")
legend("right", bty = "n", cex = 0.75,
       legend = c("Design 1 (fold)", "Design 2 (matvec)", "Gram-PSD",
                  "ideal", "unaligned"),
       col = c("black", "firebrick", "orange", "darkgreen", "grey60"),
       lty = c(1, 2, 4, 3, 3), pch = c(19, 1, NA, NA, NA))

## (2) recall vs beta: Procrustes vs near-orthogonal vs LS
plot(beta_tab$beta, beta_tab$Proc, type = "b", pch = 19, ylim = c(0, 1),
     xlab = expression(beta ~ "(non-isometry)"), ylab = sprintf("recall@%d", top_k),
     main = "Procrustes vs relaxed adapters")
lines(beta_tab$beta, beta_tab$near, type = "b", pch = 1, lty = 2, col = "blue")
lines(beta_tab$beta, beta_tab$LS, type = "b", pch = 2, lty = 3, col = "firebrick")
abline(h = ideal_b, lty = 3, col = "darkgreen")
legend("bottomleft", bty = "n", cex = 0.75,
       legend = c("Procrustes (mu=Inf)", "near-orth (mu=1)", "LS (mu=0)", "ideal"),
       col = c("black", "blue", "firebrick", "darkgreen"),
       lty = c(1, 2, 3, 3), pch = c(19, 1, 2, NA))

## (3) recall vs mu at scarce anchor (< p): the regularization sweet spot
plot(xi, mu_scarce$tab$d1, type = "b", pch = 19, ylim = c(0, 1), xaxt = "n",
     xlab = expression(mu), ylab = sprintf("recall@%d (Design 1)", top_k),
     main = "Scarce anchor (n = 24 < p = 32)")
axis(1, xi, mulab(mu_grid))
lines(xi, mu_scarce$tab$d2, type = "b", pch = 1, lty = 2, col = "firebrick")
legend("bottomleft", bty = "n", cex = 0.75,
       legend = c("Design 1 (fold)", "Design 2 (matvec)"),
       col = c("black", "firebrick"), lty = c(1, 2), pch = c(19, 1))

## (4) departure from isometry vs mu (the Design-2 / matvec budget)
plot(xi, mu_main$tab$aniso + 1e-12, type = "b", pch = 19, log = "y", xaxt = "n",
     xlab = expression(mu), ylab = expression("||" * A^T * A - I * "||"[F]),
     main = "Departure from isometry")
axis(1, xi, mulab(mu_grid))
par(op)
```

## Discussion
- **Federated retrieval across heterogeneous fine-tuned
  models.** Each of $n$ sites holds its own privately
  fine-tuned variant of a public foundation model; the
  protocol returns the top-$k$ patients across all sites
  whose private-model embeddings are most cosine-similar to a
  public-model query.
- **Site-private compatibility adapter on the $\mu$ axis.** Each
  site fits an adapter $A_k$ on a public anchor cohort along the
  family $\min_A \lVert Z_{\text{priv}}A - Z_{\text{pub}}\rVert^2
  + \mu\lVert A^\top A - I\rVert^2$ — orthogonal Procrustes at
  $\mu\to\infty$, least squares at $\mu = 0$. The adapter never
  leaves the site and never appears in the clear at the master.
- **Two deployments, one axis.** The adapter is either applied
  to the encrypted query (Design 2, the matvec) or folded into
  the database offline (Design 1, a plain encrypted inner
  product). The two coincide exactly at the orthogonal endpoint
  (where every per-vector norm is one) and part company as $\mu$
  relaxes; Design 1 then gives correct cosines while Design 2
  does not. The matvec is thus the large-$\mu$ branch.
- **Threshold key generation, $n$-of-$n$.** The CKKS secret
  key is split across all sites. The master holds no usable
  secret material. Any subset short of the full $n$ cannot
  decrypt anything along the way.
- **Diagonal-encoded matvec under threshold CKKS.** A
  $p \times p$ matrix–vector multiply on an encrypted vector,
  implemented as $p$ rotations, $p$ multiplications by
  unencrypted vectors, and $p$ encrypted additions, all within
  one level of the precision budget. The
  joint rotation keys for the cyclic-slot rotations come
  from an $n$-of-$n$ ceremony that mirrors the encryption-key
  ceremony.
- **Inner-product reduction via $\log p$ rotation-and-add.**
  Slot-wise multiply by the site's own unencrypted database
  vector followed by a halve-and-fold reduction places the
  cosine similarity in slot 0.
- **Top-$k$ over cleartext indices, scores threshold-decrypted.**
  Local patient indices stay in the clear at each site; only the
  similarity scores are encrypted, and the threshold ceremony
  reveals values only for the released top-$k$. The encrypted
  protocol's output agrees with the same computation run
  unencrypted to $\sim 10^{-11}$.
- **Fidelity under non-isometric drift.** When fine-tuning is a
  genuine non-isometry, the rigid orthogonal endpoint leaves
  recall on the table while a relaxed (least-squares or
  near-orthogonal) adapter tracks the no-drift ideal. At the
  isometric special case the two coincide — that special case is
  the orthogonal protocol of the earlier design.
- **Why not metric learning?** The convex Gram / Mahalanobis fit
  ($\min_{M\succeq0}\lVert Z_{\text{priv}}MZ_{\text{priv}}^\top -
  Z_{\text{pub}}Z_{\text{pub}}^\top\rVert^2$) is tempting, but its
  minimizer is exactly $A_{\text{LS}}A_{\text{LS}}^\top$ and its
  symmetric-root deployment carries the adapter's orthogonal
  polar factor — an unrecoverable rotation against a public-frame
  query. Its recall collapses to the unaligned floor in the sweep
  above. Gram matching is the right tool for private↔private
  retrieval (where the rotation cancels), not for a public query
  against a private database; pointwise alignment, the $\mu$
  family, is what pins the frame.

## Limitations
- **Encrypted top-$k$ selection.** We sort in the clear
  after threshold decrypting the per-patient scores. Encrypted
  argmax / top-$k$ via polynomial sign approximation is
  feasible in CKKS but adds depth and complexity orthogonal
  to this vignette's pedagogical aim. With encrypted top-$k$,
  the master would never see the full score distribution —
  only the top-$k$ release.
- **Normalization, and what the $\mu$ knob trades.** The
  near-isometry of the adapter controls the per-vector norms
  $\lVert A_k^\top v\rVert$. At the orthogonal end they are all
  one, which buys two things at once: the homomorphic
  inverse-square-root (the depth-dominating Newton iteration of
  Qu & Xu 2023 / Prantl et al. *De Bello Homomorphico* 2023) is
  avoided, *and* every per-site score is a cosine in $[-1,1]$ on
  the same scale across sites, so the raw-database matvec
  (Design 2) is directly comparable. As $\mu$ relaxes for better
  fidelity, those norms spread; the cure is to fold the adapter
  into the database offline and unit-normalize there (Design 1),
  which restores comparable cosines at the cost of materializing
  public-compatible embeddings and re-folding them on a
  public-model upgrade. So $\mu$ trades fidelity against the
  raw-database-matvec posture, with offline folding as the
  release valve — not a single forced choice.
- **Slot-tiling for production scale.** The vignette runs at
  $p = 32$ with one encrypted value per database vector.
  Production at $p = 512$ would pack many database vectors into
  the slots of a single encrypted value, amortizing the matvec
  and inner-product cost across patients within a site.
- **Real-model validation.** The synthetic data above drifts by
  a parameterized non-isometric $B_k = Q_k D_k$, a controllable
  stand-in for free fine-tuning. Real fine-tuned foundation
  models (BiomedCLIP fine-tuned per site, PubMedBERT, etc.) drift
  in ways no parameterized family captures exactly. The deferred
  side experiment in `similarity-sideexp.md` examines the
  alignment question empirically on a real fine-tuned model.
- **Adaptive-query leakage.** Repeated adaptive queries by
  the querier (or a coalition with the querier) leak
  structural information about the cohort — the same
  Hyrum-style observation that applies to any retrieval
  system. We do not eliminate this leakage; we acknowledge
  it as the cost of any released-function output.
- **Malicious-secure threshold protocol.** The trust model
  here is honest-but-curious. A malicious-secure variant
  would require zero-knowledge proofs of correct partial
  decryption and verifiable computation on the encrypted
  scores; that is heavy machinery and out of scope.

## Where this fits

This vignette extends the threshold-FHE family in
`homomorpheR` — `cox-threshold` and `cvxr-consensus-admm` —
with a third protocol shape. Cox-threshold and ADMM use a
master/worker fan-in — each worker's encrypted contribution is
summed by the master; the similarity protocol uses a
broadcast-and-aggregate shape: encrypted query in, per-site
scoring, master-side threshold decryption of the released
top-$k$. The actor surface (`make_threshold_master`,
`make_worker`, `master_encrypt`/`master_decrypt`) carries over
unchanged.

The closest precedents in the literature combine some but
not all of the ingredients used here: FRAG (Lin et al.,
arXiv:2410.13272) federates encrypted retrieval across
distrusting parties but assumes a shared embedding model;
FedE4RAG (arXiv:2504.19101) federates *training* of RAG
retrievers under CKKS-on-gradients, with each client running
its own fine-tuned model, but does the cross-client alignment
through knowledge distillation rather than a fitted
compatibility adapter; the AMPPERE three-party setup for PPER
(arXiv:2405.18430) uses CKKS for entity resolution but
assumes homogeneous models and tokenization-based alignment.
The composition of (a) site-private fine-tuned models, (b)
a private per-site compatibility adapter on the near-isometry
($\mu$) family, fit on a public anchor cohort, (c) threshold
CKKS with $n$-of-$n$ decryption, and (d) two interchangeable
deployments (encrypted-query matvec or offline fold) with the
adapter held privately at each site does not appear to have
been studied as a unified pipeline.

```{r echo = FALSE}
## Detach openfhe.R so its `encrypt`/`decrypt` generics do not
## bleed into other vignettes built later in the same R session.
detach("package:openfhe.R", unload = TRUE, character.only = TRUE)
```
