---
title: "Federated Cosine Similarity with Site-Private Fine-Tuned Models"
author: "Balasubramanian Narasimhan"
date: '`r Sys.Date()`'
bibliography: homomorphing.bib
output:
  html_document:
  fig_caption: yes
  theme: cerulean
  toc: yes
  toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Federated Cosine Similarity with Site-Private Fine-Tuned Models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)
```

## Why this vignette exists

A recurring problem in federated medical analytics is similarity
search across silos. A clinician at one hospital sees an unusual
case and wants to ask "do other hospitals in our network have
similar patients?" — without sending the patient out, without
the other hospitals revealing which of their patients were
checked, and without any single party (including a coordinating
master) seeing similarity scores in the clear.

The natural primitive is cosine similarity between embeddings.
Modern foundation models map patient data (radiology reports,
chest X-rays, pathology slides) into a fixed-dimensional vector
space where geometric proximity tracks clinical similarity.
Cosine similarity reduces to an inner product on unit-norm
vectors, which is exactly what threshold CKKS computes
efficiently.

The wrinkle is that hospitals do not all use the *same* model.
Each hospital starts from a public foundation model and
fine-tunes it on its own data, producing a privately fine-tuned
variant whose embeddings differ from the public baseline by an
unknown rotation of the embedding geometry. A query in the
public-model space is no longer directly comparable to a
database vector in a hospital's private space.

This vignette implements a protocol that handles the gap. Each
site fits a private orthogonal *translation map* on a public
anchor cohort, applies it homomorphically to encrypted queries,
and computes encrypted inner products against its private
database. Threshold decryption returns the top-`k` matches to
the querier, with no party able to decrypt intermediate
ciphertexts unilaterally.

The accompanying `similarity-sideexp.md` document describes a
deferred experiment that validates the orthogonal-alignment
assumption empirically on a real fine-tuned model. The vignette
itself uses synthetic data so that it renders quickly during
package build.

## The setup

Three sites — labeled $S_1, S_2, S_3$ — each hold a cohort of
patients. Each cohort is embedded with the site's privately
fine-tuned model into $\mathbf{R}^p$ for some fixed dimension
$p$ (the architecture is shared across sites; only the weights
differ).

The query enters the system from a clinician who does not have
access to any site's private model. They embed their candidate
patient with the *public* foundation model, producing $q \in
\mathbf{R}^p$. The query is encrypted under a joint threshold
key whose secret-key material is shared $n$-of-$n$ across the
three sites.

The retrieval target is the top-$k$ patients across all sites
whose private-model embeddings are most cosine-similar to the
query, returned as a list of $(\text{site\_id},
\text{local\_patient\_id}, \text{score})$ tuples.

The friction is that $q$ lives in the public-model geometry
while each site's database lives in the site's private-model
geometry. We close the gap with a per-site *orthogonal
translation map* $M_k$ fit on a public anchor cohort. $M_k$ is
held privately by site $k$ and never leaves; it is applied to
encrypted queries to translate them into site $k$'s private
geometry before the inner-product step.

## Threat model

Three sites and one untrusted aggregator (the master):

- **Sites $S_1, S_2, S_3$** each hold private patient
  embeddings, a private translation map $M_k$, and a secret-key
  share $\mathit{sk}_k$. They are honest-but-curious among
  themselves and toward the master. Each site sees the master's
  encrypted query but cannot decrypt it; each site sees its own
  database in the clear (it is its own data); each site never
  sees other sites' databases or scores.
- **Master** holds no secret-key material. It receives the
  encrypted query from the querier, broadcasts it to the sites,
  collects their encrypted scores, orchestrates the n-of-n
  partial-decryption ceremony for the top-$k$ results, and
  returns them to the querier. A curious or compromised master
  cannot decrypt anything by itself.
- **Querier** sees only the top-$k$ result tuples. Repeated
  adaptive queries leak structural information about the
  database in the same Hyrum-style sense as any retrieval
  system; this leakage is acknowledged but not eliminated.

What the master sees, by stage:

1. The encrypted query $\mathit{ct}_q$. Not decryptable alone.
2. Per-site encrypted score ciphertexts. Not decryptable alone.
3. Partial decryptions of the top-$k$ scores from each site.
   Not decryptable individually.
4. After fusion, the plaintext top-$k$ scores. Released to the
   querier as the protocol output.

Step 4 is the only plaintext the master ever holds, and only
for the top-$k$ scores released to the querier — not for any
intermediate quantity.

The site-private translation maps $\{M_k\}$ never leave their
sites. They are applied to ciphertexts as plaintext-matrix ×
ciphertext-vector multiplications inside the homomorphic
evaluation, so the translation never appears in the clear at
any party other than its owner.

## The protocol

The protocol has three phases. The setup phase runs once,
offline; the query phase and result phase run for each query.

**Setup phase.**

1. The three sites jointly generate a CKKS threshold key pair
   ($n$-of-$n$). Joint rotation keys for slot rotations
   $1, 2, \ldots, p-1$ are also generated by an $n$-of-$n$
   ceremony.
2. A public anchor cohort of patients with public-model
   embeddings is published.
3. Each site embeds the anchor cohort with its private
   fine-tuned model and solves orthogonal Procrustes to produce
   a private $p \times p$ orthogonal matrix $M_k$ that maps
   public-model space into the site's private-model space.
4. Each site stores its local cohort, embedded with the
   private model and normalized to unit length.

**Query phase.** The querier embeds their candidate patient
with the public model, normalizes to unit length, encrypts
into a slot-batched ciphertext under the joint public key,
and broadcasts the ciphertext to the master.

For each site $k$:

5. The site receives the encrypted query $\mathit{ct}_q$.
6. The site applies its private $M_k$ to the ciphertext via
   diagonal-encoded matrix–vector multiplication, producing
   $\mathit{ct}_{M_k q}$ — the query in the site's private
   geometry, still encrypted, still unit-norm by orthogonality
   of $M_k$.
7. The site computes inner products against each of its local
   embeddings $v_{k,i}$ via slot-wise multiplication followed
   by log-$p$ rotation-and-add slot summation, producing one
   encrypted scalar per local patient.
8. The site returns its score ciphertexts to the master,
   alongside plaintext local patient indices.

**Result phase.**

9. The master concatenates encrypted scores across all sites.
   In v1, plaintext local indices are kept alongside
   encrypted scores; encrypted top-$k$ selection in CKKS is
   feasible but adds depth and complexity orthogonal to this
   vignette's pedagogical aim.
10. The master collects partial decryption shares from all
    three sites for the encrypted score ciphertexts. Fusion
    yields plaintext scores; the master sorts plaintext-side
    and returns the top-$k$ to the querier.

## Synthetic data

The vignette uses fully synthetic data so it renders quickly
during package build. The phenotype-mixture model below is the
simplest setup that gives a meaningful retrieval ground truth.

```{r}
suppressPackageStartupMessages({
  library(homomorpheR)
  library(openfhe)
})
set.seed(20260428)

p             <- 32L
n_sites       <- 3L
cohort_sizes  <- c(80L, 60L, 100L)
n_anchor      <- 100L
n_phenotypes  <- 5L
top_k         <- 5L
```

A patient is one of `n_phenotypes` clinical phenotypes. Each
phenotype is a Gaussian cluster in $\mathbf{R}^p$. Phenotype
labels are the retrieval ground truth: a query of phenotype
$j$ should be matched to database patients of phenotype $j$.

```{r}
phenotype_centers <- matrix(rnorm(n_phenotypes * p, sd = 1),
                            n_phenotypes, p)

embed_public <- function(n) {
  labels  <- sample.int(n_phenotypes, n, replace = TRUE)
  centers <- phenotype_centers[labels, , drop = FALSE]
  noise   <- matrix(rnorm(n * p, sd = 0.4), n, p)
  z <- centers + noise
  z <- z / sqrt(rowSums(z^2))            # unit norm
  list(z = z, label = labels)
}
```

The site-private model is simulated as a per-site orthogonal
rotation $M_k$ of the public-model geometry. Light, moderate,
and heavy fine-tuning regimes correspond to three drift
magnitudes; the vignette will report retrieval recall as the
drift varies.

```{r}
random_orthogonal_drift <- function(p, alpha) {
  ## Cayley transform of a random skew-symmetric matrix gives
  ## an exactly orthogonal M without needing matrix exponentials.
  Z <- matrix(rnorm(p * p), p, p)
  S <- (Z - t(Z)) / 2                    # skew-symmetric
  half <- (alpha / 2) * S
  ## (I - half)^{-1} (I + half) is orthogonal whenever S is
  ## skew-symmetric. Closed form, base R only.
  Ip <- diag(p)
  solve(Ip - half) %*% (Ip + half)
}

embed_private <- function(z_public, M_k) {
  ## Site k's fine-tuned model in column-vector convention:
  ## f_k(x) = M_k · f(x). For a matrix z_public of n row-stacked
  ## unit vectors, the private embeddings are
  ## (M_k %*% t(z_public))^T == z_public %*% t(M_k).
  v <- z_public %*% t(M_k)
  v / sqrt(rowSums(v^2))                 # unit norm preserved by orthogonality
}
```

The public anchor cohort is a small, publicly-available set
of patient examples. Each site embeds the anchors twice — once
with the public model (giving $Z_{\text{pub}}$, identical at
every site by construction) and once with its own private
fine-tuned model (giving $Z_{\text{priv},k}$). The pair drives
the orthogonal Procrustes fit.

```{r}
fit_procrustes <- function(Z_pub, Z_priv) {
  ## Orthogonal Procrustes in the column-vector convention:
  ## return M such that M · z_pub ≈ z_priv (z's treated as column
  ## vectors). Closed form: M = U V^T where SVD(Z_priv^T Z_pub) = U Σ V^T.
  sv <- svd(crossprod(Z_priv, Z_pub))
  sv$u %*% t(sv$v)
}
```

Three drift regimes for the per-site fine-tuning:

```{r}
drift_alpha <- c(light = 0.05, moderate = 0.20, heavy = 0.50)
```

The site cohorts and anchor cohort are sampled from the same
phenotype-mixture model (in real deployments the anchor cohort
is publicly distributed and its phenotype distribution is set
once; we approximate by independent sampling).

```{r}
public_anchor <- embed_public(n_anchor)
public_query  <- embed_public(1L)        # one query for the protocol walk-through

site_cohorts <- lapply(cohort_sizes, embed_public)
```

We will later fit one $M_k$ per site per drift regime, embed
the cohorts under each regime's private model, and report
retrieval recall as a function of drift.

## Threshold key generation

The three sites jointly construct a CKKS keypair so that the
secret key is split $n$-of-$n$ across them. The familiar
`make_threshold_master()` from `homomorpheR` wires the joint
public key and the per-site secret-key shares for us; the
master is given the joint public key but never holds usable
secret material.

```{r}
cc <- fhe_context(
  scheme               = "CKKS",
  multiplicative_depth = 3L,
  scaling_mod_size     = 45L,
  batch_size           = p,
  features             = c(Feature$MULTIPARTY, Feature$KEYSWITCH))

master <- make_threshold_master(name = "master",
                                crypto_context = cc,
                                n_sites = n_sites)
```

The joint public key is `master@joint_pubkey`; the secret-key
shares are `master@secret_keys[[k]]` for $k = 1, \ldots, n$.
The `master_encrypt(master, x)` and `master_decrypt(master,
ct, len)` generics dispatch on `ThresholdMaster` and run the
$n$-of-$n$ partial-decryption ceremony internally — no party
ever holds a usable secret unilaterally.

## Joint rotation keys

The matrix–vector multiply step in the protocol is implemented
as a *diagonal-encoded matvec*. For a $p \times p$ matrix $M$
and a slot-batched ciphertext encoding a vector
$q \in \mathbf{R}^p$:

$$
M \cdot q \;=\; \sum_{i=0}^{p-1} d_i \odot \mathrm{rot}(q, i),
$$

where $d_i$ is the $i$-th diagonal of $M$ (a $p$-vector of
plaintexts), $\odot$ is slot-wise multiplication, and
$\mathrm{rot}(q, i)$ cyclically rotates the slots of $q$ by
$i$. Each rotation requires a precomputed *rotation key*. In
single-key CKKS these are generated from the secret key; in
threshold CKKS they are generated by an $n$-of-$n$ ceremony
that mirrors the encryption-key ceremony.

The inner-product step adds a second use of rotation keys: the
log-$p$ rotation-and-add reduction that sums the slots of an
encoded $p$-vector into slot 0. The same set of rotation
indices serves both purposes; we generate keys for indices
$1, \ldots, p-1$ and rely on the subset that each step needs.

```{r}
rotation_indices <- seq_len(p - 1L)
sks  <- master@secret_keys
pks  <- list(master@joint_pubkey)        # daisy-chain pubkey at step 1
joint_pk_tag <- get_key_tag(master@joint_pubkey)

## Lead-party rotation-key generation. Populates the crypto
## context's automorphism-key registry under the lead party's
## secret-key tag.
eval_rotate_key_gen(cc, sks[[1]], rotation_indices)

lead_tag    <- get_key_tag(sks[[1]])
rot_running <- get_eval_automorphism_key_map(lead_tag)

## Sites 2..n contribute their rotation-key shares in turn.
## At each step, the running joint share is registered under
## the cumulative-pubkey tag — which for our `make_threshold_master`
## is the same `joint_pk_tag` at every step (the master daisy-chains
## pubkeys forward but only retains the final joint pubkey).
for (k in 2:n_sites) {
  share_k <- multi_eval_at_index_key_gen(
    cc, sks[[k]], rot_running,
    index_list = rotation_indices,
    key_tag = joint_pk_tag)
  rot_running <- multi_add_eval_automorphism_keys(
    cc, rot_running, share_k, key_tag = joint_pk_tag)
}

insert_eval_automorphism_key(rot_running, key_tag = joint_pk_tag)
```

After insertion, any ciphertext encrypted under the joint
public key (i.e., under `master@joint_pubkey`) can be rotated
by any index in `rotation_indices` via `eval_rotate(ct, idx)`.

A round-trip smoke test confirms the joint rotation key works.
We encrypt a known vector, rotate by 3 slots, decrypt via the
$n$-of-$n$ ceremony, and check that slot $i$ now holds
$x_{(i+3) \bmod p}$:

```{r}
x <- 1:p
ct_x      <- master_encrypt(master, x)
ct_rot    <- eval_rotate(ct_x, 3L)
recovered <- master_decrypt(master, ct_rot, len = p)

expected <- x[((seq_len(p) - 1L + 3L) %% p) + 1L]
stopifnot(max(abs(recovered - expected)) < 1e-6)
cat("rotation round-trip max error:",
    sprintf("%.2e\n", max(abs(recovered - expected))))
```

## Per-site Procrustes fit and database setup

For the protocol walk-through we use the **moderate** drift
regime; the recall-vs-drift figure later in the vignette
re-runs everything across all three regimes. Each site's
private fine-tuned model is simulated as a left-multiplication
by a private orthogonal matrix; the same matrix is recovered by
orthogonal Procrustes on the public anchor cohort.

```{r}
alpha_demo <- drift_alpha["moderate"]
M <- lapply(seq_len(n_sites),
            function(k) random_orthogonal_drift(p, alpha_demo))

## Each site's private database, embedded under that site's
## fine-tuned model (already unit-norm because M_k is orthogonal).
db <- lapply(seq_len(n_sites), function(k) {
  cohort <- site_cohorts[[k]]
  list(z = embed_private(cohort$z, M[[k]]),
       label = cohort$label)
})

## Each site fits orthogonal Procrustes on the anchor cohort.
## In our synthetic setup the recovery is exact to machine
## precision because the data-generating model is itself
## orthogonal — the recall-vs-drift figure below isolates the
## protocol-mechanics contribution from this estimation.
M_hat <- lapply(seq_len(n_sites), function(k) {
  Z_pub  <- public_anchor$z
  Z_priv <- embed_private(Z_pub, M[[k]])
  fit_procrustes(Z_pub, Z_priv)
})

## Held-out anchor reconstruction error per site (a diagnostic
## the master would receive at protocol setup time).
cat("Procrustes recovery |M_hat - M|_F per site:\n")
for (k in seq_len(n_sites)) {
  cat(sprintf("  site %d: %.2e\n", k, norm(M_hat[[k]] - M[[k]], "F")))
}
```

## Diagonal-encoded matrix-vector multiply

The CKKS-friendly way to multiply a $p \times p$ plaintext
matrix into a slot-batched encrypted vector is the
*diagonal encoding*. The matrix $M$ is decomposed into its $p$
generalized diagonals, each a length-$p$ plaintext vector:

$$
d_i[j] = M[\,j,\, (j + i) \bmod p\,], \qquad i = 0, 1, \ldots, p-1.
$$

The matrix–vector product becomes

$$
M \cdot q \;=\; \sum_{i=0}^{p-1} d_i \,\odot\, \mathrm{rot}(q, i),
$$

where $\odot$ is slot-wise multiplication and $\mathrm{rot}(q, i)$
is the $i$-step cyclic slot rotation. Each term is a single
ciphertext × plaintext multiply followed by an encrypted-
vector add. The total cost is $p$ rotations and $p$
plaintext-multiplies per matrix-vector product; the
multiplicative depth is one.

```{r}
build_diagonals <- function(M, p) {
  ## d_i[j] = M[j, ((j-1 + i) %% p) + 1]   (1-based R indexing)
  lapply(0:(p - 1L), function(i) {
    vapply(seq_len(p),
           function(j) M[j, ((j - 1L + i) %% p) + 1L],
           numeric(1L))
  })
}

encrypted_matvec <- function(ct_q, M, cc, p) {
  diags <- build_diagonals(M, p)
  ct_acc <- NULL
  for (i in 0:(p - 1L)) {
    d_pt <- make_ckks_packed_plaintext(cc, diags[[i + 1L]])
    ct_term <- if (i == 0L) {
      eval_mult(ct_q, d_pt)
    } else {
      eval_mult(eval_rotate(ct_q, i), d_pt)
    }
    ct_acc <- if (is.null(ct_acc)) ct_term else eval_add(ct_acc, ct_term)
  }
  ct_acc
}
```

A round-trip smoke test on a known query vector confirms the
matvec recovers $M_1 \cdot q$ to floating-point precision:

```{r}
q_demo <- public_query$z[1, ]
ct_q  <- master_encrypt(master, q_demo)
ct_Mq <- encrypted_matvec(ct_q, M_hat[[1]], cc, p)
Mq_recovered <- master_decrypt(master, ct_Mq, len = p)
Mq_expected  <- as.numeric(M_hat[[1]] %*% q_demo)
cat(sprintf("matvec max error (site 1): %.2e\n",
            max(abs(Mq_recovered - Mq_expected))))
```

## Inner product against the local database

After the matvec, the encrypted query lives in site $k$'s
private geometry as a unit-norm vector. The cosine similarity
against a private-database vector $v$ (also unit-norm and held
in plaintext at the site) is just

$$
\cos(M_k q,\, v) \;=\; \langle M_k q,\, v \rangle
\;=\; \sum_{j=1}^{p} (M_k q)_j \cdot v_j.
$$

In CKKS this is a slot-wise multiply of the matvec output by the
plaintext-encoded $v$, followed by a log-$p$ rotate-and-add
*slot-summation reduction* that places the summed inner product
in slot 0 (and uninteresting partial sums in the other slots).

```{r}
slot_sum_reduction <- function(ct, p) {
  ## Standard CKKS log-p reduction. After the loop, slot 0 of
  ## the result holds sum_{j=1}^{p} ct[j]; other slots hold
  ## partial sums and are not used.
  step <- p %/% 2L
  while (step >= 1L) {
    ct <- eval_add(ct, eval_rotate(ct, step))
    step <- step %/% 2L
  }
  ct
}

encrypted_inner_product <- function(ct_x, v_plain, cc, p) {
  pt_v <- make_ckks_packed_plaintext(cc, v_plain)
  ct_prod <- eval_mult(ct_x, pt_v)
  slot_sum_reduction(ct_prod, p)
}
```

A smoke test against a single database vector confirms the
inner product matches its plaintext counterpart at slot 0:

```{r}
v_test <- db[[1]]$z[1, ]
ct_score <- encrypted_inner_product(ct_Mq, v_test, cc, p)
score_recovered <- master_decrypt(master, ct_score, len = 1L)
score_expected  <- sum(Mq_expected * v_test)
cat(sprintf("inner-product error (site 1, patient 1): %.2e\n",
            abs(score_recovered - score_expected)))
```

## Site `local_fn`: full per-site protocol step

The site-side computation closes over the site's private
$M_k$ and database $D_k$, takes the encrypted query
ciphertext as input, and returns a list of encrypted scores
plus plaintext local indices.

```{r}
make_similarity_site_fn <- function(M_k, db_k, cc, p) {
  function(ct_q) {
    ct_Mq <- encrypted_matvec(ct_q, M_k, cc, p)
    n_local <- nrow(db_k$z)
    ct_scores <- lapply(seq_len(n_local), function(i) {
      encrypted_inner_product(ct_Mq, db_k$z[i, ], cc, p)
    })
    list(scores = ct_scores,
         local_index = seq_len(n_local),
         label = db_k$label)
  }
}

site_fns <- lapply(seq_len(n_sites), function(k) {
  make_similarity_site_fn(M_hat[[k]], db[[k]], cc, p)
})
```

A timed end-to-end run on the demo query through site 1
returns one encrypted score per local patient:

```{r}
t0 <- proc.time()
site1_out <- site_fns[[1]](ct_q)
elapsed <- (proc.time() - t0)[["elapsed"]]
cat(sprintf("site 1 produced %d encrypted scores in %.2f s\n",
            length(site1_out$scores), elapsed))
```

## Master orchestration and threshold-decrypted top-`k`

The master broadcasts the encrypted query to every site,
collects per-site encrypted score ciphertexts plus plaintext
local patient indices, and runs the $n$-of-$n$ threshold
decryption ceremony for each score so it can sort them
plaintext-side and return the top-`k`.

The decryption pattern is per-patient: each encrypted score
ciphertext goes through one threshold-decrypt round
(`master_decrypt()` calls `multiparty_decrypt_lead` on site 1,
`multiparty_decrypt_main` on each remaining site, and
`multiparty_decrypt_fusion` to combine). For our `r sum(cohort_sizes)`
total patients this runs in a few seconds; production
deployments would pack many patients per ciphertext via
slot tiling and amortize the ceremony.

```{r}
run_similarity_query <- function(ct_q, site_fns, master, top_k) {
  ## Fan out to every site.
  site_results <- lapply(seq_along(site_fns), function(k) {
    out <- site_fns[[k]](ct_q)
    out$site_id <- k
    out
  })

  ## Threshold-decrypt each per-patient inner-product
  ## ciphertext. v1 runs one ceremony per patient; a packed
  ## variant that fuses multiple inner products into a single
  ## ciphertext (via slot tiling) is a natural extension.
  rows <- list()
  for (s in site_results) {
    for (i in seq_along(s$scores)) {
      score <- master_decrypt(master, s$scores[[i]], len = 1L)
      rows[[length(rows) + 1L]] <- data.frame(
        site_id     = s$site_id,
        local_index = s$local_index[i],
        label       = s$label[i],
        score       = score)
    }
  }
  scored <- do.call(rbind, rows)
  scored <- scored[order(-scored$score), ]
  head(scored, top_k)
}

t0 <- proc.time()
top_result <- run_similarity_query(ct_q, site_fns, master, top_k = top_k)
elapsed <- (proc.time() - t0)[["elapsed"]]
cat(sprintf("Top-%d retrieval over %d sites and %d patients in %.1f s\n",
            top_k, n_sites, sum(cohort_sizes), elapsed))
cat(sprintf("Query phenotype label: %d\n", public_query$label))
print(top_result, row.names = FALSE)
```

## Reference-truth comparison

The protocol output is meaningful only if the encrypted-domain
inner products agree with their plaintext-domain counterparts
to within CKKS noise. The reference computation runs the same
$M_k \cdot q$ translation and inner-product reduction in
plaintext on each site's database:

```{r}
plaintext_top_k <- function(q, site_data, M_list, top_k) {
  rows <- list()
  for (k in seq_along(site_data)) {
    Mq  <- as.numeric(M_list[[k]] %*% q)
    s_k <- site_data[[k]]
    scores <- as.numeric(s_k$z %*% Mq)
    for (i in seq_along(scores)) {
      rows[[length(rows) + 1L]] <- data.frame(
        site_id     = k,
        local_index = i,
        label       = s_k$label[i],
        score       = scores[i])
    }
  }
  scored <- do.call(rbind, rows)
  scored <- scored[order(-scored$score), ]
  head(scored, top_k)
}

plain_top <- plaintext_top_k(q_demo, db, M_hat, top_k)

## Compare encrypted-domain top-k against the plaintext oracle
## by joining on (site_id, local_index).
compare <- merge(top_result, plain_top,
                 by = c("site_id", "local_index"),
                 suffixes = c("_enc", "_plain"))
score_err <- max(abs(compare$score_enc - compare$score_plain))
cat(sprintf("Top-%d encrypted vs plaintext score max error: %.2e\n",
            top_k, score_err))

## Whether the encrypted-domain top-k contains the same
## (site_id, local_index) pairs as the plaintext oracle.
enc_set   <- paste(top_result$site_id, top_result$local_index, sep = ":")
plain_set <- paste(plain_top$site_id,  plain_top$local_index,  sep = ":")
cat(sprintf("Top-%d set match: %d of %d\n",
            top_k, length(intersect(enc_set, plain_set)), top_k))
```

## Why the alignment step matters: recall vs drift

The protocol's mechanics work to floating-point precision under
threshold CKKS, demonstrated above. The remaining question is
*pedagogical*: does the per-site Procrustes alignment $M_k$
actually buy anything in the federated retrieval? Or could a
simpler protocol — broadcast the public-model query, score it
directly against each site's private-model database — work
just as well?

The next sweep answers that empirically. We compare two
retrieval strategies across the three drift regimes (light,
moderate, heavy):

- **With alignment**: the protocol implemented above. Each
  site's private translation $M_k$ is fit on the public anchor
  cohort and applied (homomorphically, in deployment) to the
  encrypted query.
- **Without alignment**: the encrypted query's slots are
  scored directly against each site's private-model database,
  ignoring drift. Equivalent to setting $M_k = I$ at every
  site.

Because step 4 confirmed the encrypted protocol agrees with
its plaintext counterpart to $\sim 10^{-11}$, the recall sweep
runs in plaintext — fast enough to average over many queries
per regime and produce a stable curve. The encrypted protocol
would produce the same recall numbers within CKKS noise.

```{r}
recall_at_k <- function(query_label, top_rows, k = top_k) {
  matched <- sum(top_rows$label == query_label)
  matched / k
}

run_one_query <- function(q_pub, q_label, db_list, M_list,
                          align = TRUE) {
  rows <- list()
  for (k_site in seq_along(db_list)) {
    Mq  <- if (align) as.numeric(M_list[[k_site]] %*% q_pub) else q_pub
    s_k <- db_list[[k_site]]
    scores <- as.numeric(s_k$z %*% Mq)
    for (i in seq_along(scores)) {
      rows[[length(rows) + 1L]] <- data.frame(
        site_id     = k_site,
        local_index = i,
        label       = s_k$label[i],
        score       = scores[i])
    }
  }
  scored <- do.call(rbind, rows)
  scored <- scored[order(-scored$score), ]
  head(scored, top_k)
}

n_queries  <- 50L
sweep_grid <- c(0.00, 0.05, 0.10, 0.20, 0.35, 0.50, 0.75, 1.00)
sweep <- vector("list", length(sweep_grid) * 2L)
idx <- 0L
for (alpha_k in sweep_grid) {
  ## Refit per-site M and rebuild private databases at this drift.
  M_regime <- lapply(seq_len(n_sites),
                     function(k) random_orthogonal_drift(p, alpha_k))
  db_regime <- lapply(seq_len(n_sites), function(k) {
    cohort <- site_cohorts[[k]]
    list(z = embed_private(cohort$z, M_regime[[k]]),
         label = cohort$label)
  })
  M_hat_regime <- lapply(seq_len(n_sites), function(k) {
    Z_pub  <- public_anchor$z
    Z_priv <- embed_private(Z_pub, M_regime[[k]])
    fit_procrustes(Z_pub, Z_priv)
  })
  query_pop <- embed_public(n_queries)
  for (cond in c("with_alignment", "without_alignment")) {
    align <- cond == "with_alignment"
    rec <- numeric(n_queries)
    for (q in seq_len(n_queries)) {
      top_rows <- run_one_query(query_pop$z[q, ], query_pop$label[q],
                                db_regime, M_hat_regime, align = align)
      rec[q] <- recall_at_k(query_pop$label[q], top_rows)
    }
    idx <- idx + 1L
    sweep[[idx]] <- data.frame(
      alpha       = alpha_k,
      condition   = cond,
      mean_recall = mean(rec),
      se_recall   = sd(rec) / sqrt(n_queries))
  }
}
sweep_df <- do.call(rbind, sweep)
print(sweep_df, row.names = FALSE)
```

The contrast between the two conditions is the headline of the
figure. With alignment, recall stays high across all three drift
magnitudes because $M_k$ recovers the same orthogonal map that
generated the data — the inner product reduces to its
public-space form $\langle q, \text{cohort}\rangle$. Without
alignment, recall degrades as drift increases: the inner
product becomes $\langle q, M_k \cdot \text{cohort}\rangle$,
which drifts away from the public-space inner product as $M_k$
moves away from the identity.

```{r recall-figure, fig.width=6, fig.height=4, fig.align="center"}
op <- par(mar = c(4.5, 4.5, 1.5, 1.0))
ymin <- min(sweep_df$mean_recall - sweep_df$se_recall) - 0.05
ymax <- 1.0
plot(NA, NA,
     xlim = c(0, max(sweep_grid) * 1.05),
     ylim = c(ymin, ymax),
     xlab = expression(paste("Drift magnitude  ", alpha)),
     ylab = sprintf("Recall@%d (averaged over %d queries)", top_k, n_queries),
     bty = "L")
abline(h = 1, lty = 3, col = "grey60")
for (cond in c("with_alignment", "without_alignment")) {
  d <- sweep_df[sweep_df$condition == cond, ]
  d <- d[order(d$alpha), ]
  col_cond <- if (cond == "with_alignment") "black" else "firebrick"
  lty_cond <- if (cond == "with_alignment") 1 else 2
  pch_cond <- if (cond == "with_alignment") 19 else 17
  arrows(d$alpha, d$mean_recall - d$se_recall,
         d$alpha, d$mean_recall + d$se_recall,
         angle = 90, code = 3, length = 0.04, col = col_cond)
  lines(d$alpha, d$mean_recall, lty = lty_cond, col = col_cond)
  points(d$alpha, d$mean_recall, pch = pch_cond, col = col_cond)
}
legend("bottomleft",
       legend = c("with Procrustes alignment", "no alignment"),
       col = c("black", "firebrick"),
       lty = c(1, 2), pch = c(19, 17), bty = "n")
par(op)
```

## What this delivers

- **Federated retrieval across heterogeneous fine-tuned
  models.** Each of $n$ sites holds its own privately
  fine-tuned variant of a public foundation model; the
  protocol returns the top-$k$ patients across all sites
  whose private-model embeddings are most cosine-similar to a
  public-model query.
- **Site-private orthogonal alignment.** Each site fits an
  orthogonal Procrustes map $M_k$ on a public anchor cohort
  and applies it homomorphically to the encrypted query so
  that scoring happens in the site's private geometry. The
  matrix $M_k$ never leaves the site and never appears in the
  clear at the master.
- **Threshold key generation, $n$-of-$n$.** The CKKS secret
  key is split across all sites. The master holds no usable
  secret material. Any subset short of the full $n$ cannot
  decrypt anything along the way.
- **Diagonal-encoded matvec under threshold CKKS.** A
  $p \times p$ matrix–vector multiply on a slot-batched
  ciphertext, implemented as $p$ rotations + $p$ plaintext-
  multiplies + $p$ ciphertext-additions at depth one. The
  joint rotation keys for the cyclic-slot rotations come
  from an $n$-of-$n$ ceremony that mirrors the encryption-key
  ceremony.
- **Inner-product reduction via $\log p$ rotation-and-add.**
  Slot-wise multiply by the plaintext database vector
  followed by a halve-and-fold reduction places the cosine
  similarity in slot 0.
- **Top-$k$ over plaintext indices, scores threshold-decrypted.**
  Local patient indices stay plaintext at each site; only the
  similarity scores are encrypted, and the threshold ceremony
  reveals plaintext only for the released top-$k$. The
  encrypted protocol's output agrees with its plaintext oracle
  to $\sim 10^{-11}$.
- **Recall invariant under orthogonal drift.** Across the
  full range of drift magnitudes simulated above, the
  Procrustes-aligned protocol returns the same top-$k$ as
  cosine retrieval in public space. Without alignment, recall
  collapses for moderate-to-heavy drift.

## What this still does not deliver

- **Encrypted top-$k$ selection.** We sort plaintext-side
  after threshold decrypting the per-patient scores. Encrypted
  argmax / top-$k$ via polynomial sign approximation is
  feasible in CKKS but adds depth and complexity orthogonal
  to this vignette's pedagogical aim. With encrypted top-$k$,
  the master would never see the full score distribution —
  only the top-$k$ release.
- **Server-side normalization.** The orthogonality constraint
  on $M_k$ removes the need for a homomorphic inverse-square-
  root: $\|M_k q\| = \|q\| = 1$ by construction. For richer
  fine-tuning regimes where the alignment is not exactly an
  isometry, server-side renormalization via Newton iteration
  would be required and would dominate the depth budget.
- **Slot-tiling for production scale.** The vignette runs at
  $p = 32$ with one ciphertext per database vector. Production
  at $p = 512$ would pack many database vectors per ciphertext
  via slot tiling, amortizing the matvec and inner-product
  cost across patients within a site.
- **Real-model validation.** The synthetic data above is
  generated by an exactly-orthogonal $M_k$, which the
  Procrustes fit recovers to machine precision. Real
  fine-tuned foundation models (BiomedCLIP fine-tuned per
  site, PubMedBERT, etc.) are unlikely to drift along an
  isometry exactly. The deferred side experiment in
  `similarity-sideexp.md` validates the orthogonal-alignment
  assumption empirically on a real fine-tuned model.
- **Adaptive-query leakage.** Repeated adaptive queries by
  the querier (or a coalition with the querier) leak
  structural information about the cohort — the same
  Hyrum-style observation that applies to any retrieval
  system. We do not eliminate this leakage; we acknowledge
  it as the cost of any released-function output.
- **Malicious-secure threshold protocol.** The trust model
  here is honest-but-curious. A malicious-secure variant
  would require zero-knowledge proofs of correct partial
  decryption and verifiable computation on the score
  ciphertexts; that is heavy machinery and out of scope.

## Where this fits

This vignette extends the threshold-FHE family in
`homomorpheR` — `cox-threshold` and `cvxr-consensus-admm` —
with a third protocol shape. Cox-threshold and ADMM use a
round-robin sum over sites; the similarity protocol uses a
broadcast-and-aggregate shape: encrypted query in, per-site
scoring, master-side threshold decryption of the released
top-$k$. The actor surface (`make_threshold_master`,
`make_site`, `master_encrypt`/`master_decrypt`) carries over
unchanged.

The closest precedents in the literature combine some but
not all of the ingredients used here: FRAG (Lin et al.,
arXiv:2410.13272) federates encrypted retrieval across
distrusting parties but assumes a shared embedding model;
FedE4RAG (arXiv:2504.19101) federates *training* of RAG
retrievers under CKKS-on-gradients, with each client running
its own fine-tuned model, but does the cross-client alignment
through knowledge distillation rather than orthogonal
Procrustes; the AMPPERE three-party setup for PPER
(arXiv:2405.18430) uses CKKS for entity resolution but
assumes homogeneous models and tokenization-based alignment.
The composition of (a) site-private fine-tuned models, (b)
private per-site orthogonal alignment fit on a public anchor
cohort, (c) threshold CKKS with $n$-of-$n$ decryption, and
(d) encrypted query application with the alignment matrix
held privately at each site does not appear to have been
studied as a unified pipeline.

```{r echo = FALSE}
## Detach openfhe so its `encrypt`/`decrypt` generics do not
## bleed into other vignettes built later in the same R session.
detach("package:openfhe", unload = TRUE, character.only = TRUE)
```
